### 2016-03-18 exam2.tex
exam

\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
\usepackage{siunitx}
\usepackage{amsmath}
\usepackage{url}
\sisetup{detect-weight=true, detect-family=true}
\newcommand{\micron}[1] {\SI{#1}{\micro\meter}}

%\usepackage{setspace}
%\onehalfspace

\title{}
\author{Machine Learning Influenced of High Dimensional Time Varying Data}
\committee{ Dr. Michael Grossberg(Advisor)}
\submitted{}
\abstract{}
\begin{document}
\makefrontmatter

\section{Introduction}

Large datasets of climate and weather data are being produced from multiple sources. With cheaper and increased computer power, and cheaper disk storage, it is now possible to more easily create long model simulations at higher resolutions. Increasingly  high 
resolution satellite data and reanalysis projects which blend both model prediction and measurements create readily available large uniformly sampled spatio-temporal resolution datasets. Measured data from networks of ground stations and proxy data from trees, 
glaciers, oceans, corals, fossils, and historical records are also becoming more accessible, 
yielding sometimes irregularly sampled data that provides a host of 
information on climate variability. The great size of many of these data sets mean that just providing a download link does not really make the data accessible to the public or even other scientists. It is important for the scientists who produce the data to have a way to make it explorable..

%%some sorta transition from the generic to the specific tools-and shift this to be more tool oriented

Munzner \cite{Munzner14} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day. She contrasts this with datasets that are amassed over time where the length of the record does not intrinsically mean the data is time varying; for example [insert something that isn't horse racing]

Often the observations in these time-varying datasets are also spatially varying, spreading over the earth or the brain or another variable space where all the observations in the high dimensional space are of the same unit. %%note: must get some reference/notation/way to discuss this

This poses a difficult visualization task because the researcher is trying to capture multiple levels of interaction:
\begin{enumerate}
	\item intraobservational: snowfall in New York and snowfall in New Jersey on August 9th 2015)
	\item interobservational: global snowfall on August 9th and August 10th
	\item a mixture thereof: does snowfall in New York on the 9th affect snowfall in New Jersey on the 10th?
\end{enumerate}

\section{Clustering Based Aggregation}
\label{sec:clustering}
Wijk and Selow \cite{WijkSelow99} worked with a special case of spatio-temperal data wherein the spatial domain is also time. They were exploring trends on multiple time scales simultaneously, so the spatial domain was in effect the relationship of hours to neighbouring hours (typically a pixel in a more typical geographic map) and how those are blocked into days (geographical borders) or even months (continents) and then how those higher level groupings relate to each other. 
Wikj and Selow studied the energy consumption of and number of employees in an office building over the course of a year. They first built a 3D visualization %%include graphics
wherein the days and hours are mapped on different axis, colors and height are both used to indicate kilowatt usage. Because the aggregated monthly patterns get lost in the 3D, and would likely be difficult to see in even a 2D heatmap because of the high temperoal resolution, Wikj and Selow turn to machine learning and common idioms to reveal aggregated patterns.

They first turn the data into a vector space wherein each day is an observation consisting of the energy consumption measurements for each hour. They then cluster the days to find which days have similar energy consumption patterns. To then make this data intuitive to the user, they display this information as linked 2d views: one side is a calendar wherein the days are colored based on their cluster membership and the other side of the visualization is a simple line plot of the cluster means, themselves colored in the same scheme as the calendar. %insert screenshot here

\section{Spatio-Temporal Earth Science}
\label{sec:climate}
Climate studies are a common application of spatio-temporal visualization techniques because they exemplify the importance of visualizing the temporal, spatial, and spatio-temporal relationships. 
%%Actually need to pull the vistrails paper, possibly also UV-CDAT paper (modern one)

\subsection{Visualization Toolkits}
Some research institutions provide a tool to explore data they created, such as the \cite{src:esrlpsd}, but since all these tools are site specific, a researcher would either have to use the datasets on the website or submit their datasets for uploading 
(assuming their dataset meets the submission criteria for an aggregation site). A scientist also has the option of using the CDAT suite of libraries, developed by the \cite{WilliamsEtAl13}, GrADS, \cite{src:grads},and Ferret, \cite{src:HankinEtAl96}. The major critique with all of these though is that they mostly render the data as is, with limited support for statistical analysis and the application of typical meterological analysis. Support for even common machine learning techniques often requires extending the library; therefore for the purpose of this discussion it is somewhat useful to treat these tools more as plotting libraries than visualization systems. %%somewhere scope out/differentiate between plotting and viz

%%pull these papers out of SunZhou
\subsection{Temporal Trajectory Analysis}
\subsection{Spatial Pattern Indices}

%%this paper is kinda terrible and I should probably not use it
Many analysis and visualizations of climate data rely on machine learning techniques such as clustering, but they treat each observation as independent or they do not ascertain if a pixel affects its neighbor. Lin, Xie, Song, and Wu explore those relationships through a tight clustering method that is evaluated using a Value-Process measurement  \cite{LinXieSongWu09}.

They first create a matrix of the form time stamp by measurement for each location (latitude, longitude pair) on earth. The temporal resolution is monthly from 1901 to 200 and the measurements are similarity estimation, temperature, precipitation, annual amplitude of temperature, annual amplitude of precipitation. They apply a principal component to each of these matrices to extract the first principal component, which is what they then use to filter our dissimilar points within a radius R of the point. They then refine their filter by computing the distance between the remaining points using the Pearson product moment correlation coefficient. Afterwards they draw a similarity graph amongst the remaining points.

A fairly common problem in visualizing spatio-temporal data is to visualize changes in land ise and cover-whether the land gets drier or more covered with vegetation-over time. Since there are many techniques that will show the temporal dynamics over time, Sun and Zhou \cite{SunZhou16} developed a method to quantify and visualize how both the time and neighboring pixels effected changes in the land cover in the Yuli County in the Xinjiang Uygur
Autonomous Region of western China.

Sun and Zhou acquired data from four satellites over a period of time where vegetation would strongly stand out from other land types; they then classified the land types in the satellite images into one of 6 classes using a Maximum Likelihood Classifier and finally merged the the six classes into either farmland or other. They coded farmland as a binary 1 and others as 0 to create a binary image for each timestamp and then computed the trajectory of each pixel by examining the change in class of each pixel over time. Because the set of all possible trajectories was too large to feasibly look at, they generalized the trajectories to four categories based on whether it had changed from one state to another and how long it had stayed in that state. They then visualized these trajectories by assigning a color to each of the most common ones and placing it on a heat map, essentially encoding distinct temporal patterns using colors.

\section{Ensemble Visualizations}
Ensemble models are a special case of spatio-temporal visualizations; they are created by running the same model (often a time dependent one) using different initial conditions. Each of the resulting set of modeled data is called a member of the ensemble. Ensembles have the same spatio-temperoal data visualization difficulties as any other spatio-temporally varying dataset and the added complication of incorporating the information gained from using an ensemble of models rather than a singular model.

%%Nir paper? (My work)

\section{Conclusion}
\label{sec:conclusion}

\pagebreak
\bibliographystyle{abbrv}
\bibliography{exam2}

\end{document}

### 2016-06-02 exam2.tex
started sketching scope and rearranging

@@
 
-Large datasets of climate and weather data are being produced from multiple sources.
+%Large datasets of climate and weather data are being produced from multiple sources.
 With cheaper and increased computer power, and cheaper disk storage, it is now possible to more easily create long model simulations at higher resolutions.
 Increasingly  high
-resolution satellite data and reanalysis projects which blend both model prediction and measurements create readily available large uniformly sampled spatio-temporal resolution datasets.
+%resolution satellite data and reanalysis projects which blend both model prediction and measurements create readily available large uniformly sampled spatio-temporal resolution datasets.
 Measured data from networks of ground stations and proxy data from trees,
-glaciers, oceans, corals, fossils, and historical records are also becoming more accessible, 
-yielding sometimes irregularly sampled data that provides a host of 
-information on climate variability.
+%glaciers, oceans, corals, fossils, and historical records are also becoming more accessible, 
+%yielding sometimes irregularly sampled data that provides a host of 
+%information on climate variability.
 The great size of many of these data sets mean that just providing a download link does not really make the data accessible to the public or even other scientists.
 It is important for the scientists who produce the data to have a way to make it explorable..
+
+\subsection{Scope}
+
+This paper focuses on the visualization of:
+\begin{itemize}
+\item climate and weather data - observational and model
+\item has both a spaitial and temporal component
+\item requires aggregation(ML or stats) before can be visualized with simple viz (line, bar, pie, scatter, heatmap)
+\end{itemize}
 

### 2016-06-16 exam2.tex
updated working title
more fleshed out outline + more intro folded into latex

@@
 
-\title{}
-\author{Machine Learning Influenced of High Dimensional Time Varying Data}
-\committee{ Dr. Michael Grossberg(Advisor)}
+\title{Visualizing Relationships Amongst and With Climate Ensembles}
+\author{Hannah Aizenman}
+\committee{ Dr. Michael Grossberg(Advisor), Robert Haralick, Huy Vo}
 \submitted{}
-\abstract{}
+\abstract{Climate datasets often inherently try to encapsulate complex systems.
+The variables are often high dimensional and often have long ranging interactions both temporally and spatially.
+Ensemble climate models often add another layer to the complexity because they add uncertainty to the already complex interactions.
+Scientists often try to visualize these interactions either by attempting to encode many if them in a visualization, using a machine learning algorithm to model the interaction and traditional approaches to visualize them, or building out a complex toolkit that relies on users to compare graphs of each of the sub interactions to build out the relationships.
+Tamara'Munzner's domain-task-idiom framework can be an effective way of grouping the visualization techniques that have become dominant in the ensemble evaluation community and understanding what it is they're trying to accomplish.}
 \begin{document}
@@
 It is important for the scientists who produce the data to have a way to make it explorable..
+
+And that those explorations are understandable...while many papers(\cite{Munzner} \cite{Acquired Codes} \cite{...}discuss how more complex tasks inherently encourage more complex visualizations, they also warn that more complex visual idioms are often more difficult to understand and so they may not be beneficial.
+Some authors though discuss how highly specialized dashboards may be more productive once specialists are trained on the software.
+
+Organization:
+\begin{table}
+data & task & idiom  & AUDIENCE\\
+
+\end{table}
 
@@
 
-\section{Clustering Based Aggregation}
-\label{sec:clustering}
-Wijk and Selow \cite{WijkSelow99} worked with a special case of spatio-temperal data wherein the spatial domain is also time.
-They were exploring trends on multiple time scales simultaneously, so the spatial domain was in effect the relationship of hours to neighbouring hours (typically a pixel in a more typical geographic map) and how those are blocked into days (geographical borders) or even months (continents) and then how those higher level groupings relate to each other.
-Wikj and Selow studied the energy consumption of and number of employees in an office building over the course of a year. They first built a 3D visualization %%include graphics
-wherein the days and hours are mapped on different axis, colors and height are both used to indicate kilowatt usage.
-Because the aggregated monthly patterns get lost in the 3D, and would likely be difficult to see in even a 2D heatmap because of the high temperoal resolution, Wikj and Selow turn to machine learning and common idioms to reveal aggregated patterns.
+\section{Simple idioms, hard math}
 
-They first turn the data into a vector space wherein each day is an observation consisting of the energy consumption measurements for each hour.
-They then cluster the days to find which days have similar energy consumption patterns.
-To then make this data intuitive to the user, they display this information as linked 2d views: one side is a calendar wherein the days are colored based on their cluster membership and the other side of the visualization is a simple line plot of the cluster means, themselves colored in the same scheme as the calendar.
-%insert screenshot here
+\subsection{Probabilistic aggregations/information gain metrics}
+\label{sec:probabilistic}
+%%Dump Nir into here-it kinda ties into the band depth stuff that'll show up later
+\subsection{markov chain analysis}
+\subsection{clustering}
 
-\section{Spatio-Temporal Earth Science}
-\label{sec:climate}
-Climate studies are a common application of spatio-temporal visualization techniques because they exemplify the importance of visualizing the temporal, spatial, and spatio-temporal relationships. 
-%%Actually need to pull the vistrails paper, possibly also UV-CDAT paper (modern one)
+\section{Let's throw Everything in one picture}                                                        
+%% Evolution of Single Task/Single Idiom to Higher Order 
+\subsection{Scatter, matrix of scatter, scagnostics}                                        
+\subsection{Quiver, Flow, Spaghetti, Lasagna}
+\subsection{BoxPlots, Functional Box Plots, Countour Box Plots}
 
-\subsection{Visualization Toolkits}
+Countour Boxplots \cite are a cross between boxplots and the  
+\subsection{Maps & Heatmaps -> Glyphs}
+
+\subsection{Let's Jam all the simple together as Toolkits}
 Some research institutions provide a tool to explore data they created, such as the \cite{src:esrlpsd}, but since all these tools are site specific, a researcher would either have to use the datasets on the website or submit their datasets for uploading
@@
 %%somewhere scope out/differentiate between plotting and viz
-
-%%pull these papers out of SunZhou
-\subsection{Temporal Trajectory Analysis}
-\subsection{Spatial Pattern Indices}
-
-%%this paper is kinda terrible and I should probably not use it
-Many analysis and visualizations of climate data rely on machine learning techniques such as clustering, but they treat each observation as independent or they do not ascertain if a pixel affects its neighbor.
-Lin, Xie, Song, and Wu explore those relationships through a tight clustering method that is evaluated using a Value-Process measurement  \cite{LinXieSongWu09}.
-
-They first create a matrix of the form time stamp by measurement for each location (latitude, longitude pair) on earth.
-The temporal resolution is monthly from 1901 to 200 and the measurements are similarity estimation, temperature, precipitation, annual amplitude of temperature, annual amplitude of precipitation.
-They apply a principal component to each of these matrices to extract the first principal component, which is what they then use to filter our dissimilar points within a radius R of the point.
-They then refine their filter by computing the distance between the remaining points using the Pearson product moment correlation coefficient.
-Afterwards they draw a similarity graph amongst the remaining points.
-
-A fairly common problem in visualizing spatio-temporal data is to visualize changes in land ise and cover-whether the land gets drier or more covered with vegetation-over time.
-Since there are many techniques that will show the temporal dynamics over time, Sun and Zhou \cite{SunZhou16} developed a method to quantify and visualize how both the time and neighboring pixels effected changes in the land cover in the Yuli County in the Xinjiang Uygur
-Autonomous Region of western China.
-
-Sun and Zhou acquired data from four satellites over a period of time where vegetation would strongly stand out from other land types; they then classified the land types in the satellite images into one of 6 classes using a Maximum Likelihood Classifier and finally merged the the six classes into either farmland or other.
-They coded farmland as a binary 1 and others as 0 to create a binary image for each timestamp and then computed the trajectory of each pixel by examining the change in class of each pixel over time.
-Because the set of all possible trajectories was too large to feasibly look at, they generalized the trajectories to four categories based on whether it had changed from one state to another and how long it had stayed in that state.
-They then visualized these trajectories by assigning a color to each of the most common ones and placing it on a heat map, essentially encoding distinct temporal patterns using colors.
-
-\section{Ensemble Visualizations}
-Ensemble models are a special case of spatio-temporal visualizations; they are created by running the same model (often a time dependent one) using different initial conditions.
-Each of the resulting set of modeled data is called a member of the ensemble.
-Ensembles have the same spatio-temperoal data visualization difficulties as any other spatio-temporally varying dataset and the added complication of incorporating the information gained from using an ensemble of models rather than a singular model.
-
-%%Nir paper? (My work)
 

### 2016-06-21 exam2.tex
countor box first draft and found a host of related papers-and I have to do spagetthit and lasgana if I'm doing cb...

@@
 \subsection{BoxPlots, Functional Box Plots, Countour Box Plots}                                      
+A box plot is a really simple measure of uncertainity: there's a line and then two bars represent the uncertainity of that point. %insert image
+While this is fairly readable for small datasets, %%\cite{} argues that the upper limit is x/y/z.
 
-Countour Boxplots \cite are a cross between boxplots and the  
+Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
+creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
+statistics for a measure called band depth.
+Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.
+The bands are then sorted by band depth such that the median is the ensemble with about 50\% of its members withen all envelopes formed by other bands (so most centered).
+Outer bands are chosen according to the task at hand.
+In \cite{Whitaker2013}, they apply the countour boxplots to temperature visualizations.
+%%insert figures here.
+The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in (fig) because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
+Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray).
+It retains much of the information of the spagethhit, but removes the visual noise of the lines.
 \subsection{Maps & Heatmaps -> Glyphs}

### 2016-06-24 exam2.tex
mostly just reference revision
updated abstract & stuff based on Tufte's talk...

@@
 
-\title{Visualizing Relationships Amongst and With Climate Ensembles}
+\title{Contextualizing Climate Ensembles}
 \author{Hannah Aizenman}
@@
 \submitted{}
-\abstract{Climate datasets often inherently try to encapsulate complex systems.
-The variables are often high dimensional and often have long ranging interactions both temporally and spatially.
-Ensemble climate models often add another layer to the complexity because they add uncertainty to the already complex interactions.
-Scientists often try to visualize these interactions either by attempting to encode many if them in a visualization, using a machine learning algorithm to model the interaction and traditional approaches to visualize them, or building out a complex toolkit that relies on users to compare graphs of each of the sub interactions to build out the relationships.
-Tamara'Munzner's domain-task-idiom framework can be an effective way of grouping the visualization techniques that have become dominant in the ensemble evaluation community and understanding what it is they're trying to accomplish.}
+\abstract{
+  Problem: Ensembles inherently encode uncertainity since they are different
+  variations on the same model. While there are many ways to visualize
+  ensembles, and the uncertianity in the ensembles, these visualizations are
+  often at least a step or two removed from what is happening in the real
+  world. 
+  Importance: Climate & weather ensembles are widley used by scientists and
+  policy makers to evaluate risk; this is readily seen in Hurrican season when
+  governor's have to decide on whether to move people based on how sure they
+  are that the model's predictions are really accurate. Decontextualized
+  visualizations that are more about the method than the data can occlude this
+  decision process.
+
+  Solution: THere are a kitchen sinks worth of methods and toolkits for
+  visualizing ensembles, but the best ones clearly articulate the
+  spatio-temporal relationships in the ensembles and between the variables such
+  that the user can definitively parse out what a line or dot is saying about
+  the temperature in New Jersey.  
+}
+
 \begin{document}
@@
 \section{Introduction}
+
+``Contextualize all the things'' is important. Tufte stresses it because he's
+confirmatory, but Munzner's exploratray task-data-idiom approach is also
+grounded in it. This is because while very pretty visualizations look well
+pretty, at the end of the day they're semi-useless if there are too many steps
+removed between the visualization and the underlying data.
 

### 2016-07-20 exam2.tex
found Wickham Survey on boxplots...fills out most of that section

@@
 
-  Solution: THere are a kitchen sinks worth of methods and toolkits for
-  visualizing ensembles, but the best ones clearly articulate the
+  Solution: While there are many worth of methods and toolkits for
+  visualizing ensembles, the best ones clearly articulate the
   spatio-temporal relationships in the ensembles and between the variables such
@@
 
-Organization:
-\begin{table}
-data & task & idiom  & AUDIENCE\\
-
-\end{table}
-
-\subsection{Scope}
-
-This paper focuses on the visualization of:
-\begin{itemize}
-\item climate and weather data - observational and model
-\item has both a spaitial and temporal component
-\item requires aggregation(ML or stats) before can be visualized with simple viz (line, bar, pie, scatter, heatmap)
-\end{itemize}
-
-%%some sorta transition from the generic to the specific tools-and shift this to be more tool oriented
-
 Munzner \cite{Munzner14} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day.
@@
 
-\section{Simple idioms, hard math}
+\section{Visualizing Data & Distributions}
 
-\subsection{Probabilistic aggregations/information gain metrics}
-\label{sec:probabilistic}
-%%Dump Nir into here-it kinda ties into the band depth stuff that'll show up later
-\subsection{markov chain analysis}
-\subsection{clustering}
+Line, Scatter, Heat...end on the histogram
 
-\section{Let's throw Everything in one picture}                                                        
-%% Evolution of Single Task/Single Idiom to Higher Order 
-\subsection{Scatter, matrix of scatter, scagnostics}                                        
-\subsection{Quiver, Flow, Spaghetti, Lasagna}
-\subsection{BoxPlots, Functional Box Plots, Countour Box Plots}                                      
-A box plot is a really simple measure of uncertainity: there's a line and then two bars represent the uncertainity of that point. %insert image
-While this is fairly readable for small datasets, %%\cite{} argues that the upper limit is x/y/z.
+\section{Visualizing Uncertainty}
+
+While bar graphs and histograms start giving a sense of the variability in
+data, they are limited to showing the dispersion of the data only in 1
+dimension; namely the measureents themselves. To explore more dimensions,
+Tukey introduced the box and whiskers plot \cite{Tukey1970, Tukey1977}. It
+takes the distributional information of a histogram and encodes the summary
+statistics in a box and outlier information as whiskers coming out of the box,
+as seen in %make a figure that shows how a histogram gets encoded as a box and
+%whisker.
+The advantage of the boxplot is that these boxes can then be strung together,
+such that changes in a distribution over time (%show lineplot w/ box & whisker)
+or differences over time can be compared. In their survey of the history of the
+boxplot \cite{Wickham2011} Wickham and Stryjewski trace the evolution of the
+boxplot from a difinitive box to amorphous envelopes.  They start by laying out
+the core statistics of a boxplot: a box centered at the median (or mean) and
+bound by the upper and lower quartiles, and two whiskers which represent the extremes
+(often 1.5 the outer quartiles, sometimes the \alpha bounds). Sometimes there's
+also a dot to signify outliers, as seen in \ref{imaginary figure}, but one of
+the strengths of the box plot has been in it's flexibility. Many authors have
+kept the basic structure, but used different quantile \cite{Hyndman}s or
+measures of extremes \cite{Frigge, carter} or used assymetric whiskers \cite{Rousseuw}
+or otherwise exploited the structure to incorporate skewness, kurtosis, and
+other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}
+%find papers not in Wickham
+
+Inferential information was incorporated through the notched boxplot. \cite{McGill} % find/print this
+% paper
+In this plot
 

### 2016-07-25 sections/toolkits.tex
reorginazation + new outline

--- 2016-06-21 exam2.tex
@@
-\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
-\usepackage{siunitx}
-\usepackage{amsmath}
-\usepackage{url}
-\sisetup{detect-weight=true, detect-family=true}
-\newcommand{\micron}[1] {\SI{#1}{\micro\meter}}
-
-%\usepackage{setspace}
-%\onehalfspace
-
-\title{Visualizing Relationships Amongst and With Climate Ensembles}
-\author{Hannah Aizenman}
-\committee{ Dr. Michael Grossberg(Advisor), Robert Haralick, Huy Vo}
-\submitted{}
-\abstract{Climate datasets often inherently try to encapsulate complex systems.
-The variables are often high dimensional and often have long ranging interactions both temporally and spatially.
-Ensemble climate models often add another layer to the complexity because they add uncertainty to the already complex interactions.
-Scientists often try to visualize these interactions either by attempting to encode many if them in a visualization, using a machine learning algorithm to model the interaction and traditional approaches to visualize them, or building out a complex toolkit that relies on users to compare graphs of each of the sub interactions to build out the relationships.
-Tamara'Munzner's domain-task-idiom framework can be an effective way of grouping the visualization techniques that have become dominant in the ensemble evaluation community and understanding what it is they're trying to accomplish.}
-\begin{document}
-\makefrontmatter
-
-\section{Introduction}
-
-%Large datasets of climate and weather data are being produced from multiple sources.
-With cheaper and increased computer power, and cheaper disk storage, it is now possible to more easily create long model simulations at higher resolutions.
-Increasingly  high
-%resolution satellite data and reanalysis projects which blend both model prediction and measurements create readily available large uniformly sampled spatio-temporal resolution datasets.
-Measured data from networks of ground stations and proxy data from trees,
-%glaciers, oceans, corals, fossils, and historical records are also becoming more accessible, 
-%yielding sometimes irregularly sampled data that provides a host of 
-%information on climate variability.
-The great size of many of these data sets mean that just providing a download link does not really make the data accessible to the public or even other scientists.
-It is important for the scientists who produce the data to have a way to make it explorable..
-
-And that those explorations are understandable...while many papers(\cite{Munzner} \cite{Acquired Codes} \cite{...}discuss how more complex tasks inherently encourage more complex visualizations, they also warn that more complex visual idioms are often more difficult to understand and so they may not be beneficial.
-Some authors though discuss how highly specialized dashboards may be more productive once specialists are trained on the software.
-
-Organization:
-\begin{table}
-data & task & idiom  & AUDIENCE\\
-
-\end{table}
-
-\subsection{Scope}
-
-This paper focuses on the visualization of:
-\begin{itemize}
-\item climate and weather data - observational and model
-\item has both a spaitial and temporal component
-\item requires aggregation(ML or stats) before can be visualized with simple viz (line, bar, pie, scatter, heatmap)
-\end{itemize}
-
-%%some sorta transition from the generic to the specific tools-and shift this to be more tool oriented
-
-Munzner \cite{Munzner14} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day.
-She contrasts this with datasets that are amassed over time where the length of the record does not intrinsically mean the data is time varying; for example [insert something that isn't horse racing]
-
-Often the observations in these time-varying datasets are also spatially varying, spreading over the earth or the brain or another variable space where all the observations in the high dimensional space are of the same unit.
-%%note: must get some reference/notation/way to discuss this
-
-This poses a difficult visualization task because the researcher is trying to capture multiple levels of interaction:
-\begin{enumerate}
-	\item intraobservational: snowfall in New York and snowfall in New Jersey on August 9th 2015)
-	\item interobservational: global snowfall on August 9th and August 10th
-	\item a mixture thereof: does snowfall in New York on the 9th affect snowfall in New Jersey on the 10th?
-\end{enumerate}
-
-\section{Simple idioms, hard math}
-
-\subsection{Probabilistic aggregations/information gain metrics}
-\label{sec:probabilistic}
-%%Dump Nir into here-it kinda ties into the band depth stuff that'll show up later
-\subsection{markov chain analysis}
-\subsection{clustering}
-
-\section{Let's throw Everything in one picture}                                                        
-%% Evolution of Single Task/Single Idiom to Higher Order 
-\subsection{Scatter, matrix of scatter, scagnostics}                                        
-\subsection{Quiver, Flow, Spaghetti, Lasagna}
-\subsection{BoxPlots, Functional Box Plots, Countour Box Plots}                                      
-A box plot is a really simple measure of uncertainity: there's a line and then two bars represent the uncertainity of that point. %insert image
-While this is fairly readable for small datasets, %%\cite{} argues that the upper limit is x/y/z.
-
-Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
-creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
-statistics for a measure called band depth.
-Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.
-The bands are then sorted by band depth such that the median is the ensemble with about 50\% of its members withen all envelopes formed by other bands (so most centered).
-Outer bands are chosen according to the task at hand.
-In \cite{Whitaker2013}, they apply the countour boxplots to temperature visualizations.
-%%insert figures here.
-The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in (fig) because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
-Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray).
-It retains much of the information of the spagethhit, but removes the visual noise of the lines.
-\subsection{Maps & Heatmaps -> Glyphs}
-
-\subsection{Let's Jam all the simple together as Toolkits}
 Some research institutions provide a tool to explore data they created, such as the \cite{src:esrlpsd}, but since all these tools are site specific, a researcher would either have to use the datasets on the website or submit their datasets for uploading
@@
 %%somewhere scope out/differentiate between plotting and viz
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-\pagebreak
-\bibliographystyle{abbrv}
-\bibliography{exam2}
-
-\end{document}

### 2016-07-25 exam2.tex
reorginazation + new outline

--- 2016-06-21 exam2.tex
@@
 \documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+\usepackage{import}
 \usepackage{siunitx}
@@
 
-\title{Visualizing Relationships Amongst and With Climate Ensembles}
+\title{Visualizing Multivariate PDFs}
 \author{Hannah Aizenman}
@@
 \submitted{}
-\abstract{Climate datasets often inherently try to encapsulate complex systems.
-The variables are often high dimensional and often have long ranging interactions both temporally and spatially.
-Ensemble climate models often add another layer to the complexity because they add uncertainty to the already complex interactions.
-Scientists often try to visualize these interactions either by attempting to encode many if them in a visualization, using a machine learning algorithm to model the interaction and traditional approaches to visualize them, or building out a complex toolkit that relies on users to compare graphs of each of the sub interactions to build out the relationships.
-Tamara'Munzner's domain-task-idiom framework can be an effective way of grouping the visualization techniques that have become dominant in the ensemble evaluation community and understanding what it is they're trying to accomplish.}
+\abstract{
+  Problem:
+  Importance:
+  Solution:
+}
+
 \begin{document}
@@
 
-%Large datasets of climate and weather data are being produced from multiple sources.
-With cheaper and increased computer power, and cheaper disk storage, it is now possible to more easily create long model simulations at higher resolutions.
-Increasingly  high
-%resolution satellite data and reanalysis projects which blend both model prediction and measurements create readily available large uniformly sampled spatio-temporal resolution datasets.
-Measured data from networks of ground stations and proxy data from trees,
-%glaciers, oceans, corals, fossils, and historical records are also becoming more accessible, 
-%yielding sometimes irregularly sampled data that provides a host of 
-%information on climate variability.
-The great size of many of these data sets mean that just providing a download link does not really make the data accessible to the public or even other scientists.
-It is important for the scientists who produce the data to have a way to make it explorable..
-
-And that those explorations are understandable...while many papers(\cite{Munzner} \cite{Acquired Codes} \cite{...}discuss how more complex tasks inherently encourage more complex visualizations, they also warn that more complex visual idioms are often more difficult to understand and so they may not be beneficial.
-Some authors though discuss how highly specialized dashboards may be more productive once specialists are trained on the software.
-
-Organization:
-\begin{table}
-data & task & idiom  & AUDIENCE\\
-
-\end{table}
-
-\subsection{Scope}
-
-This paper focuses on the visualization of:
-\begin{itemize}
-\item climate and weather data - observational and model
-\item has both a spaitial and temporal component
-\item requires aggregation(ML or stats) before can be visualized with simple viz (line, bar, pie, scatter, heatmap)
-\end{itemize}
-
-%%some sorta transition from the generic to the specific tools-and shift this to be more tool oriented
-
-Munzner \cite{Munzner14} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day.
+Munzner \cite{Munzner14ii} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day.
 She contrasts this with datasets that are amassed over time where the length of the record does not intrinsically mean the data is time varying; for example [insert something that isn't horse racing]
@@
 
-\section{Simple idioms, hard math}
-
-\subsection{Probabilistic aggregations/information gain metrics}
-\label{sec:probabilistic}
-%%Dump Nir into here-it kinda ties into the band depth stuff that'll show up later
-\subsection{markov chain analysis}
-\subsection{clustering}
-
-\section{Let's throw Everything in one picture}                                                        
-%% Evolution of Single Task/Single Idiom to Higher Order 
-\subsection{Scatter, matrix of scatter, scagnostics}                                        
-\subsection{Quiver, Flow, Spaghetti, Lasagna}
-\subsection{BoxPlots, Functional Box Plots, Countour Box Plots}                                      
-A box plot is a really simple measure of uncertainity: there's a line and then two bars represent the uncertainity of that point. %insert image
-While this is fairly readable for small datasets, %%\cite{} argues that the upper limit is x/y/z.
-
-Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
-creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
-statistics for a measure called band depth.
-Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.
-The bands are then sorted by band depth such that the median is the ensemble with about 50\% of its members withen all envelopes formed by other bands (so most centered).
-Outer bands are chosen according to the task at hand.
-In \cite{Whitaker2013}, they apply the countour boxplots to temperature visualizations.
-%%insert figures here.
-The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in (fig) because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
-Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray).
-It retains much of the information of the spagethhit, but removes the visual noise of the lines.
-\subsection{Maps & Heatmaps -> Glyphs}
-
-\subsection{Let's Jam all the simple together as Toolkits}
-Some research institutions provide a tool to explore data they created, such as the \cite{src:esrlpsd}, but since all these tools are site specific, a researcher would either have to use the datasets on the website or submit their datasets for uploading
-(assuming their dataset meets the submission criteria for an aggregation site).
-A scientist also has the option of using the CDAT suite of libraries, developed by the \cite{WilliamsEtAl13}, GrADS, \cite{src:grads},and Ferret, \cite{src:HankinEtAl96}.
-The major critique with all of these though is that they mostly render the data as is, with limited support for statistical analysis and the application of typical meterological analysis.
-Support for even common machine learning techniques often requires extending the library; therefore for the purpose of this discussion it is somewhat useful to treat these tools more as plotting libraries than visualization systems.
-%%somewhere scope out/differentiate between plotting and viz
+\section{Seperable Dimensions}
+\import{sections/}{seperable.tex}
+%%history of the boxplot and the like
+\section{Non-Seperable Dimensions}
+%%using machine learning to find seperable distributions in data/
+%%PCA, Clustering,ICA, Linear Manifold, etc...
+\section{Interdependent but Important Dimensions}
+%% Spatio Temporal Probabilitues - hurrican tracks yield curves 
+\section{Climate Visualization Toolkits}
 

### 2016-07-25 sections/seperable.tex
reorginazation + new outline

While bar graphs and histograms start giving a sense of the variability in
data, they are limited to showing the dispersion of the data only in 1
dimension; namely the measureents themselves. To explore more dimensions,
Tukey introduced the box and whiskers plot \cite{Tukey1970, Tukey1977}. It
takes the distributional information of a histogram and encodes the summary
statistics in a box and outlier information as whiskers coming out of the box,
as seen in %make a figure that shows how a histogram gets encoded as a box and
%whisker.
The advantage of the boxplot is that these boxes can then be strung together,
such that changes in a distribution over time (%show lineplot w/ box & whisker)
or differences over time can be compared. In their survey of the history of the
boxplot \cite{Wickham2011} Wickham and Stryjewski trace the evolution of the
boxplot from a difinitive box to amorphous envelopes.  They start by laying out
the core statistics of a boxplot: a box centered at the median (or mean) and
bound by the upper and lower quartiles, and two whiskers which represent the extremes
(often 1.5 the outer quartiles, sometimes the \alpha bounds). Sometimes there's
also a dot to signify outliers, as seen in \ref{imaginary figure}, but one of
the strengths of the box plot has been in it's flexibility. Many authors have
kept the basic structure, but used different quantile \cite{Hyndman}s or
measures of extremes \cite{Frigge, carter} or used assymetric whiskers \cite{Rousseuw}
or otherwise exploited the structure to incorporate skewness, kurtosis, and
other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}
%find papers not in Wickham

Inferential information was incorporated through the notched boxplot. \cite{McGill} % find/print this
% paper
In this plot

Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
statistics for a measure called band depth. Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles. The bands are then sorted by band depth such that the median is the ensemble with about 50\% of its members withen all envelopes formed by other bands (so most centered). Outer bands are chosen according to the task at hand. In \cite{Whitaker2013}, they apply the countour boxplots to temperature visualizations. %%insert figures here. 
The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in (fig) because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns. Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray). It retains much of the information of the spagethhit, but removes the visual noise of the lines.

### 2016-08-15 exam2.tex
tukey 1975 intro

@@
 \section{Introduction}
-
-Munzner \cite{Munzner14ii} defines a time-varying dataset as one in which time is an intrinsic attribute of how the various observations are measured; for example recording snow fall every 3 hours or stock prices at the beginning or end of day.
-She contrasts this with datasets that are amassed over time where the length of the record does not intrinsically mean the data is time varying; for example [insert something that isn't horse racing]
-
-Often the observations in these time-varying datasets are also spatially varying, spreading over the earth or the brain or another variable space where all the observations in the high dimensional space are of the same unit.
-%%note: must get some reference/notation/way to discuss this
-
-This poses a difficult visualization task because the researcher is trying to capture multiple levels of interaction:
-\begin{enumerate}
-	\item intraobservational: snowfall in New York and snowfall in New Jersey on August 9th 2015)
-	\item interobservational: global snowfall on August 9th and August 10th
-	\item a mixture thereof: does snowfall in New York on the 9th affect snowfall in New Jersey on the 10th?
-\end{enumerate}
+In \cite{tukey1975m} Tukey lays out the case for exploratory data analysis when
+he says ``The picturing of data must be sensitive, not only to the multiple hypotheses
+we hold, but to the many more we have not yet thought of, regard as unlikely or
+think impossible''. He advocates breaking from a deterministic \mathbox{y=mx+b}
+best fit line approach where the b is treated as an afterthough to a more
+nuanced
+\begin{align*}
+  given = fit PLUS residual
+\end{align*}
+explicitly to bring attention to the residuals. Tukey is arguing that it's not
+good enough to just look at the path of the data but also at the variation in
+the path. In some sense this is the fundemental disctinction between
+confirmatory data analysis and exploratary data analysis; whereas the former
+intends to display a picture that supports a hypothesis (such as that best fit
+line that shows the direction of the data), the latter's purpose is to
+facilitate investigation of the data (and the hypothesis). To explore data, it
+is crucial to understand the shape of the data, which can be done using a
+number of exploratory visualzation techniques. These can be subdvided into:
 

### 2016-08-15 sections/seperable.tex
tukey 1975 intro

@@
 statistics in a box and outlier information as whiskers coming out of the box,
-as seen in %make a figure that shows how a histogram gets encoded as a box and
+as seen in. %pull in more from Tukey
+
+%make a figure that shows how a histogram gets encoded as a box and
 %whisker.
+
 The advantage of the boxplot is that these boxes can then be strung together,
@@
 
+%notched boxplot
+
 Inferential information was incorporated through the notched boxplot. \cite{McGill} % find/print this
-% paper
+% paper variations
 In this plot
 
+%letter value plots
+
+%bagplots-> 
 Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 

### 2016-08-15 sections/seperable.tex
eda

@@
 dimension; namely the measureents themselves. To explore more dimensions,
-Tukey introduced the box and whiskers plot \cite{Tukey1970, Tukey1977}. It
-takes the distributional information of a histogram and encodes the summary
-statistics in a box and outlier information as whiskers coming out of the box,
-as seen in. %pull in more from Tukey
+Tukey introduced the box and whiskers plot \cite{tukey1977e}. The box and
+whisker plot takes the distributional information of a
+histogram and encodes the summary statistics in a box and outlier information as
+whiskers coming out of the box. While this is inherently useful, the
+information could have just as easily been shown in a histogram. The power of
+the boxplot is in that it facilitates the comparison of multiple distributions:
+\begin{figure}
+%% side by side  boxplot comparsions of random data?
+\end{figure}
 
-%make a figure that shows how a histogram gets encoded as a box and
-%whisker.
+Where the box and whisker plot really shines is in comparing the extreme
+values because they are represented as lines from the center and therefore the
+length of the two lines can be compared \cite{tukey1977e}.  Also, by
+translating the distribution statistics to a visual representation, the
+outliers are proportioanlly scaled to the mean and standard, which gives a
+sense of how the summary stats relate to each other that is sometimes lost in
+numerical calculations when the numbers are small or close to each
+other. %%probably needs a citation
 
-The advantage of the boxplot is that these boxes can then be strung together,
-such that changes in a distribution over time (%show lineplot w/ box & whisker)
-or differences over time can be compared. In their survey of the history of the
+In their survey of the history of the
 boxplot \cite{Wickham2011} Wickham and Stryjewski trace the evolution of the

### 2016-08-19 sections/seperable.tex
figs and references and more pieces

--- 2016-08-15 sections/seperable.tex
@@
 data, they are limited to showing the dispersion of the data only in 1
-dimension; namely the measureents themselves. To explore more dimensions,
-Tukey introduced the box and whiskers plot \cite{Tukey1970, Tukey1977}. It
-takes the distributional information of a histogram and encodes the summary
-statistics in a box and outlier information as whiskers coming out of the box,
-as seen in. %pull in more from Tukey
-
-%make a figure that shows how a histogram gets encoded as a box and
-%whisker.
-
-The advantage of the boxplot is that these boxes can then be strung together,
-such that changes in a distribution over time (%show lineplot w/ box & whisker)
-or differences over time can be compared. In their survey of the history of the
-boxplot \cite{Wickham2011} Wickham and Stryjewski trace the evolution of the
-boxplot from a difinitive box to amorphous envelopes.  They start by laying out
+dimension; namely the measurements themselves. In their survey of the history
+of the boxplot \cite{wickham11} Wickham and Stryjewski trace the evolution of
+the boxplot from a difinitive box to amorphous envelopes.  They start by laying out
 the core statistics of a boxplot: a box centered at the median (or mean) and
-bound by the upper and lower quartiles, and two whiskers which represent the extremes
-(often 1.5 the outer quartiles, sometimes the \alpha bounds). Sometimes there's
-also a dot to signify outliers, as seen in \ref{imaginary figure}, but one of
+bound by the upper and lower quartiles (termed the hinges), and two whiskers which represent the
+extremes (often 1.5 the outer quartiles, sometimes the \alpha bounds). Sometimes there's
+also a dot to signify outliers, as seen in \ref{fig:boxplot}, but one of
 the strengths of the box plot has been in it's flexibility. Many authors have
 kept the basic structure, but used different quantile \cite{Hyndman}s or
-measures of extremes \cite{Frigge, carter} or used assymetric whiskers \cite{Rousseuw}
-or otherwise exploited the structure to incorporate skewness, kurtosis, and
-other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}
-%find papers not in Wickham
+measures of extremes \cite{Frigge, carter} or used assymetric
+whiskers \cite{Rousseuw} or otherwise exploited the structure to incorporate skewness, kurtosis, and
+other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}.
+
+\begin{figure}
+\includegraphics{figs/boxplot.png}
+\caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
+Exploratory Data Analytics\cite{tukey77e}) illustrates how the box plot shows the distribution of topology of states and
+the distribution of volcanos. The strength of the box and whisker is that the
+reader can quickly compare the two distributions and learn, for example, that
+volcanos tend to span a shorter range of average heights, but that the extremes
+are much further from the center. Essentially, the distributions act as
+references for the other displayed distributions.}
+\label{fig:boxplot}
+\end{figure}
+
+In Tukey's introduction to the boxplot \cite{tukey77e}, shown in figure~\ref{fig:boxplot}, the box and whisker plot takes the
+distributional information of a histogram and encodes the state heights between the
+25th and 75th percentiles as a box, the median height as a line cutting across
+the heights, and the extreme state heights as lines coming out of the
+boxes. While a histogram would also show that the distribution of state heights has a fat right tail,
+the power of the box and whisker plot is in encoding this information such that
+it's trivial to compare to it's neighboring distribution of volcano
+heights because they can share the Y-axis. As the number of distributions grow,
+this technique becomes invaluable because the number of histograms that can be
+placed next to each other or overlayed as pdfs can quickly become
+unweildy. And for wider distributions like the state heights, it can sometimes
+be difficult to pinpoint the median without additional annotations.
 
@@
 %bagplots->
-Countour Boxplots \cite{Whitaker2013} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
+
+\begin{figure}
+\includegraphics{figs/countour_weather.png}
+\label{fig:countour}
+\end{figure}
+Countour Boxplots \cite{whitaker13} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
 creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
@@
 Outer bands are chosen according to the task at hand.
-In \cite{Whitaker2013}, they apply the countour boxplots to temperature visualizations.
-%%insert figures here.
-The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in (fig) because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
+In \cite{whitaker13}, they apply the countour boxplots to temperature visualizations.
+%
+The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in \ref{fig:countour} because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
 Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray).

### 2016-08-24 sections/seperable.tex
reworked laundry list of plots

@@
 also a dot to signify outliers, as seen in \ref{fig:boxplot}, but one of
-the strengths of the box plot has been in it's flexibility. Many authors have
-kept the basic structure, but used different quantile \cite{Hyndman}s or
-measures of extremes \cite{Frigge, carter} or used assymetric
-whiskers \cite{Rousseuw} or otherwise exploited the structure to incorporate skewness, kurtosis, and
-other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}.
+the strengths of the box plot has been in it's flexibility.
 
@@
 \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
-Exploratory Data Analytics\cite{tukey77e}) illustrates how the box plot shows the distribution of topology of states and
+Exploratory Data Analytics\cite{tukey77e}) illustrates how the box plot shows
+the distribution of topology of states and
 the distribution of volcanos. The strength of the box and whisker is that the
@@
 
-In Tukey's introduction to the boxplot \cite{tukey77e}, shown in figure~\ref{fig:boxplot}, the box and whisker plot takes the
-distributional information of a histogram and encodes the state heights between the
+In Tukey's introduction to the boxplot \cite{tukey77e}, shown in
+figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional information of a histogram and encodes the state heights between the
 25th and 75th percentiles as a box, the median height as a line cutting across
@@
 placed next to each other or overlayed as pdfs can quickly become
-unweildy. And for wider distributions like the state heights, it can sometimes
-be difficult to pinpoint the median without additional annotations.
+unweildy.  Many others authors have kept the basic structure of the boxplot,
+but exploited the structure to convey more information about the data. Authors
+have incorporated sample size through box width \cite{mcgill1978}, used
+different quantile levels \cite{Hyndman}s or measures or extremes\cite{Frigge,
+carter}, proposed assymetric whiskers \cite{Rousseuw}, and otherwise incorporated skewness, kurtosis, and other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}.
 
-%notched boxplot
-
-Inferential information was incorporated through the notched boxplot. \cite{McGill} % find/print this
-% paper variations
-In this plot
+The box-plot was then expanded out from descriptive statistics to inferential
+information through the notched boxplot. \cite{mcgill1978}
 

### 2016-08-24 sections/seperable.tex
notched box plots

@@
 unweildy.  Many others authors have kept the basic structure of the boxplot,
-but exploited the structure to convey more information about the data. Authors
-have incorporated sample size through box width \cite{mcgill1978}, used
+but exploited the structure to convey more information about the data. Authors used
 different quantile levels \cite{Hyndman}s or measures or extremes\cite{Frigge,
@@
 
-The box-plot was then expanded out from descriptive statistics to inferential
-information through the notched boxplot. \cite{mcgill1978}
+\begin{figure}
+%%\includegraphic{Mcgill} #annottate figure with highlighted line for the notches!
+\label{fig:notched}
+\end{figure}
+The box-plot was then expanded out from descriptive statistics to facilitatting
+inferential deduction through the notched boxplot. \cite{mcgill1978} The notched boxplot
+arose as an expansion of indicating sample size through the width of the box,
+which is a technique that only works well for relatively small sample sizes (\<
+1000), at larger scales the widths are about equal. Instead, notches, as seen
+in Figure~\ref{fig:notched}, are used to indicate the confidence level of the
+median for that box. The medians are statistically different from each other at
+the 95\% confidence level if the notches do not overlap. In
+Figure~\ref{fig:notched}, all the notches overlap %%annotate figure with what
+%%overlapping region means
+and so the difference in telephone bills is not-significant, but there's a
+statistically significant difference between the 11-15 and over 15 groups. Notches may extend
+beyond the box edges, and sometimes the width technique is incorporated back
+into the notched boxpplot such that the confidence can be grounded in sample
+size. The notches in Figure~\ref{fig:notched} are computed assuming a gaussian
+distribution, but many authors \cite{x,y,z} discuss ways to compute notches for
+non-gaussian distributions. 
+\end{align}
 

### 2016-08-24 sections/seperable.tex
hist, vase, violin

@@
 non-gaussian distributions.
-\end{align}
+
+\begin{figure}
+%%two figs a and b with a shared caption...
+%%\includegraphics{benjamini}
+\label{fig:histplot}
+\label{fig:vase}
+\end{figure}
+
+Because the box-plot does not give any indication of how data is distributed
+inside the box or along the whiskers, the histplot and vaseplot were introduced to indicate
+the density of the data\cite{benjamini1988}. As shown in
+Figure~\ref{fig:histplot} the histplot seperates the box
+into  5 bins and scales the width of that section of the box based on the
+number of values in that bin; it's essentially a rotated histogram. The
+vaseplot, shown in Figure~\ref{fig:vase} removes the bins and is instead based
+on the estimated density at that point in the data. The specific shape is
+dependent on what sort of density estimation method is used, such as kernal
+estimation or nearest neighbor estimation \cite{chambers1983}. The
+violinplot \cite{hintz1998} expands this concept further by representing the
+width using the density-trace method proposed by Chambers
+et. al \cite{chambers1983}%%is density trace the same as a pdf?
 

### 2016-08-24 sections/seperable.tex
ordering

@@
 bound by the upper and lower quartiles (termed the hinges), and two whiskers which represent the
-extremes (often 1.5 the outer quartiles, sometimes the \alpha bounds). Sometimes there's
+extremes (often 1.5 the outer quartiles, sometimes the $\alpha$ bounds). Sometimes there's
 also a dot to signify outliers, as seen in \ref{fig:boxplot}, but one of
@@
 \label{fig:vase}
+\label{fig:violin}
 \end{figure}
@@
 width using the density-trace method proposed by Chambers
-et. al \cite{chambers1983}%%is density trace the same as a pdf?
+et. al \cite{chambers1983}, as seen in Figure~\ref{fig:violinplot}%%is density
+%%trace the same as a pdf?
+The letter-value plot\cite{hofmann} ....
 
-%letter value plots
+%Find roussouw 1999 bagplot paper...
 
-%bagplots->
+\begin{figure}
+%%\includegraphics{bagplot}
+\end{figure}
+The boxplot is inherently limited to 1D data; the bagplot was introduced to
+visualize 2D and higher data \cite{roussow99}.
+
+\begin{figure}
+%%\includegraphics
+\end{figure}
+While boxplots were designed to visualize discrete observations, often there
+are no discrete observations and instead the `observation` is a timeseries
+curve (where the connections are as important as the individual points) or
+spatial region. To visualize these datasets, Hyndman and Shang proposed the
+rainbow plot, functional bagplot and high density region (HDR)
+plot\cite{Hyndman2009}. These visualizations rely on the idea of ordering the
+functions such that there's some measure of how close these functions are to
+each other-functional depth\cite{fabrero2007}, bivariate score depth, and
+bivariate kernel density estimation\cite{scott92} are some examples. This
+ordering then yields functional analogs to median and other percentile
+metrics. The rainbow plot simply colors the data based on its depth, shown
+in \ref{fig:rainbow}
+
+\begin{figure}
+%%\includegraphics{abc}
+\end{figure}
+
+The functional boxplot was developed to describe the distributional nature of functional data\cite{sun2011}.
 
@@
 \end{figure}
-Countour Boxplots \cite{whitaker13} are an extension of boxplots that tries to better encapsulate outliers. Box plots inherently clip the uncertainity they show to some upper and lower band, 
-creating a somewhat bounding envelope for the function that looks fairly regular when aggregated. Countour boxplots aim instead to capture the variablity of the uncertainity by trading in descripting
-statistics for a measure called band depth.
+Countour Boxplots \cite{whitaker13} are an extension of functional boxplots that tries to better encapsulate outliers.
+Box plots inherently clip the uncertainity they show to some upper and lower band,
+creating a somewhat bounding envelope for the function that looks fairly
+regular when aggregated. Countour boxplots aim instead to capture the
+variablity of the uncertainity via a band depth ordering metric.
 Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.

### 2016-08-24 exam2.tex
ordering

@@
 is crucial to understand the shape of the data, which can be done using a
-number of exploratory visualzation techniques. These can be subdvided into:
+number of exploratory visualzation techniques. Because these techniques are
+highly dependent on the structure of the data, this paper explores techniques suitable to different classes of data-seperable, non-seperable, and
+possibly seperable but interdependent-and then uses this framework to evaluate
+distrubitional visualziation in recent climate visualization toolkits.
 

### 2016-08-24 sections/seperable.tex
functional box

@@
 %%trace the same as a pdf?
-The letter-value plot\cite{hofmann} ....
+%%The letter-value plot\cite{hofmann} ....
 
@@
 The boxplot is inherently limited to 1D data; the bagplot was introduced to
-visualize 2D and higher data \cite{roussow99}.
+visualize 2D and higher data \cite{roussow1999}.
+Bivariate HDR plot \cite{Hyndman1996}
 
-\begin{figure}
-%%\includegraphics
-\end{figure}
 While boxplots were designed to visualize discrete observations, often there
@@
 curve (where the connections are as important as the individual points) or
-spatial region. To visualize these datasets, Hyndman and Shang proposed the
-rainbow plot, functional bagplot and high density region (HDR)
-plot\cite{Hyndman2009}. These visualizations rely on the idea of ordering the
+spatial region. These visualizations rely on the idea of ordering the
 functions such that there's some measure of how close these functions are to
-each other-functional depth\cite{fabrero2007}, bivariate score depth, and
+each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{Hyndman2009}, and
 bivariate kernel density estimation\cite{scott92} are some examples. This
 ordering then yields functional analogs to median and other percentile
-metrics. The rainbow plot simply colors the data based on its depth, shown
-in \ref{fig:rainbow}
+metrics. For small values, the ordered curves can then be colored based on the
+ordering metric \cite{Hyndman2009}, but this method is unweildy for larger
+ensembles.
 
 \begin{figure}
-%%\includegraphics{abc}
+%%\includegraphics{funcbaghdr}
+\label{fig:funcbag}
 \end{figure}
+Instead, Hyndmand and Shang proposed functional versions of the bag and HDR
+plots\cite{Hyndman2009}, as shown in Figure~\ref{fig:funcbag}. In these graphs,
+what was traditionally the box part of a box plot is now a region bounded by
+the curves (obtained by using one of the previously mentioned ordering
+metrics). The median curve is defined as a blackline and the exceptional
+outlines are shown. By using regions defined by the curves, the functional bag
+and HDR plots display the variance in the ensembles while simulatanously
+flattening the similaraties so that the outliers are strongly highlighted. The
+HDR method better highlights outliers closer to the median, whereas the bagplot
+method better highlights the outliers outside the envelopes of functions.
 
@@
 \end{figure}
-Countour Boxplots \cite{whitaker13} are an extension of functional boxplots that tries to better encapsulate outliers.
-Box plots inherently clip the uncertainity they show to some upper and lower band,
-creating a somewhat bounding envelope for the function that looks fairly
-regular when aggregated. Countour boxplots aim instead to capture the
-variablity of the uncertainity via a band depth ordering metric.
+While functional bag, box, and HDR plots all show outliers, each method is most
+successful visualizing a specific type of outlier. Countour
+boxplots\cite{whitaker3} aim instead to capture the variablity of the uncertainity via a band depth ordering metric.
 Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.

### 2016-08-24 sections/seperable.tex
bibliography

@@
 dimension; namely the measurements themselves. In their survey of the history
-of the boxplot \cite{wickham11} Wickham and Stryjewski trace the evolution of
+of the boxplot \cite{wickham2011} Wickham and Stryjewski trace the evolution of
 the boxplot from a difinitive box to amorphous envelopes.  They start by laying out
@@
 \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
-Exploratory Data Analytics\cite{tukey77e}) illustrates how the box plot shows
+Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot shows
 the distribution of topology of states and
@@
 
-In Tukey's introduction to the boxplot \cite{tukey77e}, shown in
+In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
 figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional information of a histogram and encodes the state heights between the
@@
 arose as an expansion of indicating sample size through the width of the box,
-which is a technique that only works well for relatively small sample sizes (\<
-1000), at larger scales the widths are about equal. Instead, notches, as seen
+which is a technique that only works well for relatively small sample sizes ($$\<1000$$), at larger scales the widths are about equal. Instead, notches, as seen
 in Figure~\ref{fig:notched}, are used to indicate the confidence level of the
@@
 violinplot \cite{hintz1998} expands this concept further by representing the
-width using the density-trace method proposed by Chambers
-et. al \cite{chambers1983}, as seen in Figure~\ref{fig:violinplot}%%is density
+width using the density-trace method (which is a probability density estimation
+via moving windowed average) proposed by Chambers et. al \cite{chambers1983}, as seen in Figure~\ref{fig:violinplot}%%is density
 %%trace the same as a pdf?
@@
 functions such that there's some measure of how close these functions are to
-each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{Hyndman2009}, and
+each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{hyndman2009}, and
 bivariate kernel density estimation\cite{scott92} are some examples. This
@@
 metrics. For small values, the ordered curves can then be colored based on the
-ordering metric \cite{Hyndman2009}, but this method is unweildy for larger
+ordering metric \cite{hyndman2009}, but this method is unweildy for larger
 ensembles.
@@
 Instead, Hyndmand and Shang proposed functional versions of the bag and HDR
-plots\cite{Hyndman2009}, as shown in Figure~\ref{fig:funcbag}. In these graphs,
+plots\cite{hyndman2009}, as shown in Figure~\ref{fig:funcbag}. In these graphs,
 what was traditionally the box part of a box plot is now a region bounded by
@@
 \begin{figure}
-\includegraphics{figs/countour_weather.png}
+%%\includegraphics{figs/countour_weather.png}
 \label{fig:countour}
@@
 successful visualizing a specific type of outlier. Countour
-boxplots\cite{whitaker3} aim instead to capture the variablity of the uncertainity via a band depth ordering metric.
+boxplots\cite{whitaker2013} aim instead to capture the variablity of the uncertainity via a band depth ordering metric.
 Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.
@@
 Outer bands are chosen according to the task at hand.
-In \cite{whitaker13}, they apply the countour boxplots to temperature visualizations.
+In \cite{whitaker2013}, they apply the countour boxplots to temperature visualizations.
 %

### 2016-08-24 exam2.tex
bibliography

@@
 \usepackage{amsmath}
+\usepackage{graphicx}
 \usepackage{url}
@@
 \submitted{}
-\abstract{
-  Problem:
-  Importance:
-  Solution:
-}
+\abstract{}
 
@@
 \section{Introduction}
-In \cite{tukey1975m} Tukey lays out the case for exploratory data analysis when
-he says ``The picturing of data must be sensitive, not only to the multiple hypotheses
-we hold, but to the many more we have not yet thought of, regard as unlikely or
-think impossible''. He advocates breaking from a deterministic \mathbox{y=mx+b}
+Tukey lays out the case for exploratory data analysis when
+he says ``The picturing of data must be sensitive, not only to the multiple
+hypotheses we hold, but to the many more we have not yet thought of, regard as unlikely or
+think impossible'' \cite{tukey1975}. He advocates breaking from a deterministic $$y=mx+b$$
 best fit line approach where the b is treated as an afterthough to a more

### 2016-08-25 sections/seperable.tex
figs compiles using standalone

@@
+\documentclass[../main.tex]{subfiles}
+
+\begin{document}
+\section{Seperable Dimensions}
+\label{sec:sepdim}
+
 While bar graphs and histograms start giving a sense of the variability in
@@
 
+\subsection{Discrete observations}
+
 \begin{figure}
-\includegraphics{figs/boxplot.png}
+\includegraphics{boxplot}
 \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
@@
 In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
-figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional information of a histogram and encodes the state heights between the
+Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional information of a histogram and encodes the state heights between the
 25th and 75th percentiles as a box, the median height as a line cutting across
@@
 
+\subsubsection{Incoporating Density}
+
 \begin{figure}
@@
 \end{figure}
-The box-plot was then expanded out from descriptive statistics to facilitatting
+The box-plot was then expanded out from descriptive statistics to facilitating
 inferential deduction through the notched boxplot. \cite{mcgill1978} The notched boxplot
 arose as an expansion of indicating sample size through the width of the box,
-which is a technique that only works well for relatively small sample sizes ($$\<1000$$), at larger scales the widths are about equal. Instead, notches, as seen
+which is a technique that only works well for relatively small sample sizes ($1\<1000$), at larger scales the widths are about equal. Instead, notches, as seen
 in Figure~\ref{fig:notched}, are used to indicate the confidence level of the
@@
 
+\subsubsection{2D Distributions}
+
 %Find roussouw 1999 bagplot paper...
@@
 
+\subsection{Functional `observations`}
+
 While boxplots were designed to visualize discrete observations, often there
@@
 curve (where the connections are as important as the individual points) or
-spatial region. These visualizations rely on the idea of ordering the
-functions such that there's some measure of how close these functions are to
-each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{hyndman2009}, and
+spatial region. Understanding the distributions of functional observations
+often relies first ordering the functions such that there's some measure of how
+close these functions are to each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{hyndman2009}, and
 bivariate kernel density estimation\cite{scott92} are some examples. This
@@
 \begin{figure}
-%%\includegraphics{figs/countour_weather.png}
+\includegraphics{contour_weather}
 \label{fig:countour}
@@
 It retains much of the information of the spagethhit, but removes the visual noise of the lines.
+\end{document}

### 2016-08-26 sections/seperable.tex
chasing down citations

@@
 it's trivial to compare to it's neighboring distribution of volcano
-heights because they can share the Y-axis. As the number of distributions grow,
+heights because they can share the Y axis. As the number of distributions grow,
 this technique becomes invaluable because the number of histograms that can be
@@
 unweildy.  Many others authors have kept the basic structure of the boxplot,
-but exploited the structure to convey more information about the data. Authors used
-different quantile levels \cite{Hyndman}s or measures or extremes\cite{Frigge,
-carter}, proposed assymetric whiskers \cite{Rousseuw}, and otherwise incorporated skewness, kurtosis, and other descriptive distributional statsics \cite{ Aslam, choon, Marmelejo}.
+but exploited the structure to convey more information about the data. Authors
+used different quantile levels \cite{hyndman1996}s or measures of outliers
+\cite{frigge1989}, or ways of identifying outliers\cite{carter2009, schwertman2004, schwertman2007}, or otherwise incorporated skewness,
+kurtosis, and other descriptive distributional statistics \cite{kim2004, hubert2008, marmolejo2015}.
 
@@
 \end{figure}
-The box-plot was then expanded out from descriptive statistics to facilitating
+The boxplot was then expanded out from descriptive statistics to facilitating
 inferential deduction through the notched boxplot. \cite{mcgill1978} The notched boxplot
 arose as an expansion of indicating sample size through the width of the box,
-which is a technique that only works well for relatively small sample sizes ($1\<1000$), at larger scales the widths are about equal. Instead, notches, as seen
+which is a technique that only works well for relatively small sample sizes (N
+less than 1000), at larger scales the widths are about equal. Instead, notches, as seen
 in Figure~\ref{fig:notched}, are used to indicate the confidence level of the
@@
 %%overlapping region means
-and so the difference in telephone bills is not-significant, but there's a
+and so the difference in telephone bills is not significant, but there's a
 statistically significant difference between the 11-15 and over 15 groups. Notches may extend
@@
 width using the density-trace method (which is a probability density estimation
-via moving windowed average) proposed by Chambers et. al \cite{chambers1983}, as seen in Figure~\ref{fig:violinplot}%%is density
-%%trace the same as a pdf?
-%%The letter-value plot\cite{hofmann} ....
+via moving windowed average) proposed by Chambers et. al \cite{chambers1983},
+as seen in Figure~\ref{fig:violinplot}
 
 \subsubsection{2D Distributions}
-
-%Find roussouw 1999 bagplot paper...
-
 \begin{figure}
@@
 The boxplot is inherently limited to 1D data; the bagplot was introduced to
-visualize 2D and higher data \cite{roussow1999}.
-Bivariate HDR plot \cite{Hyndman1996}
+visualize 2D and higher data \cite{rousseeuw1999}
+Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion of boxplot/HDR plot}
 
@@
 often relies first ordering the functions such that there's some measure of how
-close these functions are to each other-functional depth\cite{fabrero2007}, bivariate score depth\cite{hyndman2009}, and
-bivariate kernel density estimation\cite{scott92} are some examples. This
+close these functions are to each other--functional depth\cite{febrero2007},
+bivariate score depth\cite{hyndman2009}, and
+bivariate kernel density estimation\cite{scott1992} are some examples. This
 ordering then yields functional analogs to median and other percentile
@@
 
-The functional boxplot was developed to describe the distributional nature of functional data\cite{sun2011}.
+The functional boxplot was developed to describe the distributional nature of
+functional data\cite{sun2011}. \textbf{expand this section with discusson of Sun's paper}
 
 \begin{figure}
-\includegraphics{contour_weather}
+  \includegraphics[width=.75\linewidth]{contour_weather}
+  \caption{}
 \label{fig:countour}
@@
 successful visualizing a specific type of outlier. Countour
-boxplots\cite{whitaker2013} aim instead to capture the variablity of the uncertainity via a band depth ordering metric.
-Each ensemble members band depth is computed as sum of the probabilities that the observations in any given ensemble fall within the max-min envelope defined by any two other ensembles.
-The bands are then sorted by band depth such that the median is the ensemble with about 50\% of its members withen all envelopes formed by other bands (so most centered).
-Outer bands are chosen according to the task at hand.
-In \cite{whitaker2013}, they apply the countour boxplots to temperature visualizations.
-%
-The authors argue that countour boxplots are an improved visual idiom over the traditional spagehttie plots seen in \ref{fig:countour} because as seen in (fig), the spagehetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
-Instead in (fig3), the authors remove most of the bands and instead visualize an outliser envelope(light gray) and a more central envelope(dark gray).
-It retains much of the information of the spagethhit, but removes the visual noise of the lines.
+boxplots\cite{whitaker2013} aim instead to capture the variablity of the
+uncertainity via a band depth ordering metric. Each ensemble members band depth
+is computed as sum of the probabilities that the observations in any given
+ensemble fall within the max\-min envelope defined by any two other
+ensembles. The bands are then sorted by band depth such that the median is the
+ensemble with about 50\% of its members withen all envelopes formed by other
+bands (so most centered). Outer bands are chosen according to the task at
+hand. In \cite{whitaker2013}, they apply the countour boxplots to temperature
+visualizations. The authors argue that countour boxplots are an improved visual
+idiom over the traditional plots seen in Figure~\ref{fig:spag} because
+as seen in , the spagehetti plots get nosy as the number of plots increase
+and so it's hard to tease out specific patterns. Instead in Figure~\ref{fig:countour}, the authors
+remove most of the bands and instead visualize an outlier envelope(light gray)
+and a more central envelope(dark gray). It retains much of the information of
+the spagethhit, but removes the visual noise of the lines.
+
 \end{document}

### 2016-09-22 sections/draft.tex
papers, rearranging, etc

--- 2016-07-25 sections/toolkits.tex
@@
+\subsection{Structure & Seperability}
+
+\subsection{Outlier Detection}
 Some research institutions provide a tool to explore data they created, such as the \cite{src:esrlpsd}, but since all these tools are site specific, a researcher would either have to use the datasets on the website or submit their datasets for uploading

### 2016-09-22 sections/section1.tex
papers, rearranging, etc

--- 2016-08-26 sections/seperable.tex
@@
 \begin{document}
-\section{Seperable Dimensions}
-\label{sec:sepdim}
 
+\section{Independent Observations}
+\label{sec:independent}
+
+\subsection{Univariate Observations}
+
+%%Histogram paper + chambers + kde/pdf/density estimaton
+
+\subsection{Multivariate Observations}
 While bar graphs and histograms start giving a sense of the variability in
@@
 the boxplot from a difinitive box to amorphous envelopes.  They start by laying out
-the core statistics of a boxplot: a box centered at the median (or mean) and
-bound by the upper and lower quartiles (termed the hinges), and two whiskers which represent the
-extremes (often 1.5 the outer quartiles, sometimes the $\alpha$ bounds). Sometimes there's
-also a dot to signify outliers, as seen in \ref{fig:boxplot}, but one of
-the strengths of the box plot has been in it's flexibility.
+the core statistics of a boxplot, which can also be seen in Figure~\ref{fig:boxplot}.
+\begin{description}
+  \item[Median (Q2)] 50\% quartile
+  \item[Inter Quartile Range (IQR)]
+\end{description}
 
-\subsection{Discrete observations}
+\begin{tabular}{|r| r |r|r|}
+ & Lower & Upper\\
+  \multicolumn{2}{Hinge} & Q1: 25\%  & Q3: 75\% \\
+  \multicolumn{2}{Interquartile range} & \multicolumn{2}{IQR = Q3-Q1}\\
+  \multirow{2}{Fence} & Inner & Q1-1.5*IQR & Q3-1.5*IQR \\
+  & outer & Q1-3*IQR & Q3+3*IQR \\
+\end{tabular}
+
+%%Insert a labeled boxplot with fences! (this should probably just be an
+%%annotated figure
+
+\subsection{Multivariate Observations}
 
@@
   \includegraphics[width=.75\linewidth]{contour_weather}
-  \caption{}
-\label{fig:countour}
+  \caption{\label{fig:countour}}
 \end{figure}
+
 While functional bag, box, and HDR plots all show outliers, each method is most
@@
 is computed as sum of the probabilities that the observations in any given
-ensemble fall within the max\-min envelope defined by any two other
+ensemble fall within the max-min envelope defined by any two other
 ensembles. The bands are then sorted by band depth such that the median is the
@@
 visualizations. The authors argue that countour boxplots are an improved visual
-idiom over the traditional plots seen in Figure~\ref{fig:spag} because
-as seen in , the spagehetti plots get nosy as the number of plots increase
+idiom over the traditional spagehtti plots (%insert citation for them...)
+because the spagehetti plots get nosy as the number of plots increase
 and so it's hard to tease out specific patterns. Instead in Figure~\ref{fig:countour}, the authors

### 2016-09-22 main.tex
figs compiles using standalone
chasing down citations
papers, rearranging, etc

--- 2016-08-15 exam2.tex
@@
 \documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
-\usepackage{import}
-\usepackage{siunitx}
+
+\usepackage{graphicx}
+\graphicspath{{sections/figs/}{.figs/}}
+
+\usepackage[backend=bibtex, style=numeric-comp]{biblatex}
+\bibliography{main}
+
+\usepackage{subfiles}
+\usepackage{url}
 \usepackage{amsmath}
-\usepackage{url}
-\sisetup{detect-weight=true, detect-family=true}
-\newcommand{\micron}[1] {\SI{#1}{\micro\meter}}
 
-%\usepackage{setspace}
-%\onehalfspace
-
-\title{Visualizing Multivariate PDFs}
+\title{Visualizing Functional Data}
 \author{Hannah Aizenman}
@@
 \section{Introduction}
-In \cite{tukey1975m} Tukey lays out the case for exploratory data analysis when
-he says ``The picturing of data must be sensitive, not only to the multiple hypotheses
-we hold, but to the many more we have not yet thought of, regard as unlikely or
-think impossible''. He advocates breaking from a deterministic \mathbox{y=mx+b}
-best fit line approach where the b is treated as an afterthough to a more
-nuanced
-\begin{align*}
-  given = fit PLUS residual
-\end{align*}
-explicitly to bring attention to the residuals. Tukey is arguing that it's not
-good enough to just look at the path of the data but also at the variation in
-the path. In some sense this is the fundemental disctinction between
-confirmatory data analysis and exploratary data analysis; whereas the former
-intends to display a picture that supports a hypothesis (such as that best fit
-line that shows the direction of the data), the latter's purpose is to
-facilitate investigation of the data (and the hypothesis). To explore data, it
-is crucial to understand the shape of the data, which can be done using a
-number of exploratory visualzation techniques. These can be subdvided into:
 
-\section{Seperable Dimensions}
-\import{sections/}{seperable.tex}
-%%history of the boxplot and the like
-\section{Non-Seperable Dimensions}
-%%using machine learning to find seperable distributions in data/
-%%PCA, Clustering,ICA, Linear Manifold, etc...
-\section{Interdependent but Important Dimensions}
-%% Spatio Temporal Probabilitues - hurrican tracks yield curves 
-\section{Climate Visualization Toolkits}
+High dimensional data is catch-all term, typically used to describe data that's highly
+multivariate or data that's structurally complex, or even data that simply has
+a large number of observations. The lack of specificity of the phrase
+`high-demensional` can make it difficult to parse out which high-dimensional
+analysis and visaualization technique is suited to the research question at
+hand. To answer that question, it's crucial to first have a set of working
+definitions for discussing the data:
+
+\begin{definition}
+\item[measurements] criteria based assignment of numbers to objects or events \cite{grimm1993}
+\item[variables] ``any characteristics, number, or quantity that can be measured
+  or counted''\cite{ABS}
+
+\item[field] dataset type where each cell has associated information \cite{munzner14ii}    
+\end{defintion}
+
+\subfile{sections/independent.tex}
 
@@
 
+\begin{tabular}{|r|r|r|r|}
+  \hline
+              & univariate & bivariate & multivariate\\
+  \hline
+   0D         &            &           &             \\
+  \hline
+   1D         &            &           &             \\
+   \hline
+   ND         &            &           &             \\
+\end{tabular}
+
 \pagebreak
-\bibliographystyle{abbrv}
-\bibliography{exam2}
-
+\printbibliography
 \end{document}

### 2016-09-29 sections/section0d.tex
new outline

\documentclass[../main.tex]{subfiles}

\begin{document}
\section{Independent Observations (0D fields)}

\subsection{Univariate Observations}
%%histogram and pdfs/distribution fitting

\subsection{Bivariate Observations}
%%introduce box plot

\subsection{Multivariate Observations}
%% large scale box: letter, spark

\subsection{Relationships Between Variables}
\subsubsection{Bivariate}
%%scatter segue into bagplot and HDR?

\subsection{Multivariate Observations}

%% radial, parallel coordinates, factor plotl
%%high d-PCA, ICA, markov

### 2016-09-29 sections/section1d.tex
new outline

\documentclass[../main.tex]{subfiles}

\begin{document}
\section{1D Fields}

\subsection{Independent Variables}

\subsection{Univariate Observations}
%%functional box + derivatives,

\subsection{Bivariate Observations}
%%hurricane tracks (x+y dept)

\subsection{Multivariate Observations}
LDA, CCA, 
\subsection{Relationships Between Variables}

\subsubsection{Bivariate}
surface map, functional correlations, planes?
\subsection{Multivariate Observations}
PLS, functional PCA

### 2016-11-01 main.tex
abstract, intro

@@
 \submitted{}
-\abstract{}
-
+\abstract{There are an almost overwhelming number of machine learning and
+  visualization techniques, but each technique makes some underlying
+  assumptions about how the data is structured. While sometimes the
+  observations are completely independent from each other, in many instances
+  the observations are sampled from 1D continuum like time or 2d surface such
+  as space.This survey uses functional data analysis to classify visualizations based on
+  the number of variables, whether the variables are independent, and how the
+  observations are related to each other}
 \begin{document}
@@
 analysis and visaualization technique is suited to the research question at
-hand. To answer that question, it's crucial to first have a set of working
-definitions for discussing the data:
+hand. Datasets inherently have two components, the variables and the
+observations of those variables. Functional data analysis provides a language
+for formalizing the structure of the data in such a way that it clarifies the
+possible interactions in the data.
 
-\begin{definition}
-\item[measurements] criteria based assignment of numbers to objects or events \cite{grimm1993}
-\item[variables] ``any characteristics, number, or quantity that can be measured
-  or counted''\cite{ABS}
+%%paragraph on functional data analysis
 
-\item[field] dataset type where each cell has associated information \cite{munzner14ii}    
-\end{defintion}
+Observations can often be reduced to one of the following forms:
+
+\begin{tabular}
+   & univarate   & bivariate                & multivariate \\
+0D & v_i         & v0_i, v1_i               & v0_i, \dots, vn_i \\
+1D & v_i(x_i)     & v0_i(x_i), v1_i(x_i)    & v0_i(x_i), \dots, vn_i(x_i)\\
+ND & v_i(x0_i,\dots,xn_i)     & v0_i(x0_i,\dots,xn_i), v1_i(x0_i,\dots,xn_i)    & v0_i(x0_i,\dots,xn_i), \dots, vn_i(x0_i,\dots,xn_i)\\
+\end{tabular}
+
+The variables $v0,\dots, vn$ can also potentially be dependent on each other, adding in a third level of interactions that a visualization may want to caputure.
 

### 2016-11-03 main.tex
moved section text into proper sections

@@
 hand. Datasets inherently have two components, the variables and the
-observations of those variables. Functional data analysis provides a language
-for formalizing the structure of the data in such a way that it clarifies the
+observations of those variables. These oberservations are sometimes dependent
+on information rich parameters such as time and space, and often it is as
+important to visualize these underlying parametric dependencys as the
+relationships between the variables in given observations. Functional data
+analysis provides a language for formalizing the structure of the data in such a way that it clarifies the
 possible interactions in the data.
@@
 %%paragraph on functional data analysis
+The underlying assumption of functional data analysis is that while sometimes
+samples are discrete, often they are drawn from a continuos sample of
+obersvations \cite{ramsey, hook, meuller}. Independent observations
+by definition rely on zero paramters, whereas dependent observations can be
+paramterized on one dimension (such as as time), or multiple (such as time,
+latitude, longitude, altitude). %%throw some math here to transition to my table
 
-Observations can often be reduced to one of the following forms:
-
+\begin{table}
 \begin{tabular}
    & univarate   & bivariate                & multivariate \\
-0D & v_i         & v0_i, v1_i               & v0_i, \dots, vn_i \\
-1D & v_i(x_i)     & v0_i(x_i), v1_i(x_i)    & v0_i(x_i), \dots, vn_i(x_i)\\
-ND & v_i(x0_i,\dots,xn_i)     & v0_i(x0_i,\dots,xn_i), v1_i(x0_i,\dots,xn_i)    & v0_i(x0_i,\dots,xn_i), \dots, vn_i(x0_i,\dots,xn_i)\\
+0 parameter & RV_i         & v0_i, v1_i               & RV0_i, \dots, RVn_i \\
+1 paramter & RV_i(x_i)     & RV0_i(x_i), RV1_i(x_i)    & RV0_i(x_i), \dots, RVn_i(x_i)\\
+N parameters & RV_i(x0_i,\dots,xn_i)     & RV0_i(x0_i,\dots,xn_i), RV1_i(x0_i,\dots,xn_i)    & RV0_i(x0_i,\dots,xn_i), \dots, RVn_i(x0_i,\dots,xn_i)\\
 \end{tabular}
+\caption{Each $vi$ is a distinct variable such as precipitation or temperature,
+  each $xi$ is a distinct parameter such as time or space, and i is the
+  observation identifier.}
+\end{table}
 
-The variables $v0,\dots, vn$ can also potentially be dependent on each other, adding in a third level of interactions that a visualization may want to caputure.
+The variables $v0,\dots, vn$ can also potentially be dependent on each other,
+adding in a third level of interactions that a visualization may want to
+caputure. While this survey focuses on data structured in a linear, tabular, or
+data cube way, it could potentially be expanded to network data where the xi
+parameters represent edges in the graph.
 

### 2016-11-03 sections/section1d.tex
moved section text into proper sections

@@
 \section{1D Fields}
-
-\subsection{Independent Variables}
+While boxplots were designed to visualize discrete observations, often there
+are no discrete observations and instead the `observation` is a timeseries
+curve (where the connections are as important as the individual points) or
+spatial region. Understanding the distributions of functional observations
+often relies first ordering the functions such that there's some measure of how
+close these functions are to each other--functional depth\cite{febrero2007},
+bivariate score depth\cite{hyndman2009}, and
+bivariate kernel density estimation\cite{scott1992} are some examples. This
+ordering then yields functional analogs to median and other percentile
+metrics. For small values, the ordered curves can then be colored based on the
+ordering metric \cite{hyndman2009}, but this method is unweildy for larger
+ensembles.
 
 \subsection{Univariate Observations}
-%%functional box + derivatives,
+
+\begin{figure}
+  %%\includegraphics{funcbaghdr}
+  \label{fig:funcbag}
+\end{figure}
+Instead, Hyndmand and Shang proposed functional versions of the bag and HDR
+plots\cite{hyndman2009}, as shown in Figure~\ref{fig:funcbag}. In these graphs,
+what was traditionally the box part of a box plot is now a region bounded by
+the curves (obtained by using one of the previously mentioned ordering
+metrics). The median curve is defined as a blackline and the exceptional
+outlines are shown. By using regions defined by the curves, the functional bag
+and HDR plots display the variance in the ensembles while simulatanously
+flattening the similaraties so that the outliers are strongly highlighted. The
+HDR method better highlights outliers closer to the median, whereas the bagplot
+method better highlights the outliers outside the envelopes of functions.
+
+The functional boxplot was developed to describe the distributional nature of
+functional data\cite{sun2011}. \textbf{expand this section with discusson of
+  Sun's paper}
+
+\begin{figure}
+  \includegraphics[width=.75\linewidth]{contour_weather}
+  \caption{\label{fig:countour}}
+\end{figure}
+While functional bag, box, and HDR plots all show outliers, each method is most
+successful visualizing a specific type of outlier. Countour
+boxplots\cite{whitaker2013} aim instead to capture the variablity of the
+uncertainity via a band depth ordering metric. Each ensemble members band depth
+is computed as sum of the probabilities that the observations in any given
+ensemble fall within the max-min envelope defined by any two other
+ensembles. The bands are then sorted by band depth such that the median is the
+ensemble with about 50\% of its members withen all envelopes formed by other
+bands (so most centered). Outer bands are chosen according to the task at
+hand. In \cite{whitaker2013}, they apply the countour boxplots to temperature
+visualizations. The authors argue that countour boxplots are an improved visual
+idiom over the traditional spagehtti plots (%insert citation for them...)
+because the spagehetti plots get nosy as the number of plots increase
+and so it's hard to tease out specific patterns. Instead in
+Figure~\ref{fig:countour}, the aut\
+hors
+remove most of the bands and instead visualize an outlier envelope(light gray)
+and a more central envelope(dark gray). It retains much of the information of
+the spagethhit, but removes the visual noise of the lines.
 

### 2016-11-03 sections/section0d.tex
moved section text into proper sections

@@
 
+The specific shape is
+dependent on what sort of density estimation method is used, such as kernal
+estimation or nearest neighbor estimation \cite{chambers1983}. 
+width using the density-trace method-which is a probability density estimation
+via moving windowed average) proposed by Chambers et. al \cite{chambers1983},.
+
 \subsection{Bivariate Observations}
-%%introduce box plot
+\subsection{Multivariate Observations}
+While bar graphs and histograms start giving a sense of the variability in
+data, they are limited to showing the dispersion of the data only in 1
+dimension; namely the measurements themselves. In their survey of the history
+of the boxplot \cite{wickham2011} Wickham and Stryjewski trace the evolution of
+the boxplot from a difinitive box to amorphous envelopes.  They start by laying
+out
+the core statistics of a boxplot, which can also be seen in
+Figure~\ref{fig:boxplot}.
+\begin{description}
+\item[Median (Q2)] 50\% quartile
+\item[Inter Quartile Range (IQR)]
+\end{description}
+
+\begin{tabular}{|r| r |r|r|}
+  & Lower & Upper\\
+  \multicolumn{2}{Hinge} & Q1: 25\%  & Q3: 75\% \\
+  \multicolumn{2}{Interquartile range} & \multicolumn{2}{IQR = Q3-Q1}\\
+  \multirow{2}{Fence} & Inner & Q1-1.5*IQR & Q3-1.5*IQR \\
+  & outer & Q1-3*IQR & Q3+3*IQR \\
+  \end{tabular}
+
+\begin{figure}
+  \includegraphics{boxplot}
+  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
+    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
+    shows
+    the distribution of topology of states and
+    the distribution of volcanos. The strength of the box and whisker is that
+    the
+    reader can quickly compare the two distributions and learn, for example,
+    that
+    volcanos tend to span a shorter range of average heights, but that the
+    extremes
+    are much further from the center. Essentially, the distributions act as
+    references for the other displayed distributions.}
+  \label{fig:boxplot}
+\end{figure}
+
+In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
+Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional
+information of a h\
+istogram and encodes the state heights between the
+25th and 75th percentiles as a box, the median height as a line cutting across
+the heights, and the extreme state heights as lines coming out of the
+boxes. While a histogram would also show that the distribution of state heights
+has a fat righ\
+t tail,
+the power of the box and whisker plot is in encoding this information such that
+it's trivial to compare to it's neighboring distribution of volcano
+heights because they can share the Y axis. As the number of distributions grow,
+this technique becomes invaluable because the number of histograms that can be
+placed next to each other or overlayed as pdfs can quickly become
+unweildy.  Many others authors have kept the basic structure of the boxplot,
+but exploited the structure to convey more information about the data. Authors
+used different quantile levels \cite{hyndman1996}s or measures of outliers
+\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
+  schwertman2004, schwertman\
+  2007}, or otherwise incorporated skewness,
+kurtosis, and other descriptive distributional statistics \cite{kim2004,
+  hubert2008, marmolejo\
+  2015}.
+
+\begin{figure}
+  \includegraphics{boxplot}
+  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
+    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
+    shows
+    the distribution of topology of states and
+    the distribution of volcanos. The strength of the box and whisker is that
+    the
+    reader can quickly compare the two distributions and learn, for example,
+    that
+    volcanos tend to span a shorter range of average heights, but that the
+    extremes
+    are much further from the center. Essentially, the distributions act as
+    references for the other displayed distributions.}
+  \label{fig:boxplot}
+\end{figure}
+
+In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
+Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional
+information of a h\
+istogram and encodes the state heights between the
+25th and 75th percentiles as a box, the median height as a line cutting across
+the heights, and the extreme state heights as lines coming out of the
+boxes. While a histogram would also show that the distribution of state heights
+has a fat righ\
+t tail,
+the power of the box and whisker plot is in encoding this information such that
+it's trivial to compare to it's neighboring distribution of volcano
+heights because they can share the Y axis. As the number of distributions grow,
+this technique becomes invaluable because the number of histograms that can be
+placed next to each other or overlayed as pdfs can quickly become
+unweildy.  Many others authors have kept the basic structure of the boxplot,
+but exploited the structure to convey more information about the data. Authors
+used different quantile levels \cite{hyndman1996}s or measures of outliers
+\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
+  schwertman2004, schwertman\
+  2007}, or otherwise incorporated skewness,
+kurtosis, and other descriptive distributional statistics \cite{kim2004,
+  hubert2008, marmolejo\
+  2015}.
 
@@
 %%scatter segue into bagplot and HDR?
+\begin{figure}
+  %%\includegraphics{bagplot}
+\end{figure}
+The boxplot is inherently limited to 1D data; the bagplot was introduced to
+visualize 2D and higher data \cite{rousseeuw1999}
+Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion of
+  boxplot/HDR plot}
 

### 2016-11-16 main.tex
table and conditionals

@@
 \author{Hannah Aizenman}
-\committee{ Dr. Michael Grossberg(Advisor), Robert Haralick, Huy Vo}
+\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Huy Vo}
 \submitted{}
@@
   the observations are sampled from 1D continuum like time or 2d surface such
-  as space.This survey uses functional data analysis to classify visualizations based on
-  the number of variables, whether the variables are independent, and how the
-  observations are related to each other}
+  as space.This survey uses functional data analysis to classify visualizations
+  based on the number of variables in and level of paramterization of the data.}
 \begin{document}
@@
 \section{Introduction}
-
-High dimensional data is catch-all term, typically used to describe data that's highly
-multivariate or data that's structurally complex, or even data that simply has
+High dimensional data is catch-all term, typically used to describe data that
+highly multivariate or data that's structurally complex, or even data that simply has
 a large number of observations. The lack of specificity of the phrase
@@
 relationships between the variables in given observations. Functional data
-analysis provides a language for formalizing the structure of the data in such a way that it clarifies the
-possible interactions in the data.
+analysis provides a language for formalizing the structure of the data in such
+a way that it clarifies the possible interactions in the data.
 
@@
 samples are discrete, often they are drawn from a continuos sample of
-obersvations \cite{ramsey, hook, meuller}. Independent observations
-by definition rely on zero paramters, whereas dependent observations can be
-paramterized on one dimension (such as as time), or multiple (such as time,
-latitude, longitude, altitude). %%throw some math here to transition to my table
+obersvations \cite{ramsey, hook, meuller}. That continuous sample is often over
+space or time or another attribute on which the observation is conditional.
 
-\begin{table}
+\begin{table}%%use conditional notation since it's the same
 \begin{tabular}
    & univarate   & bivariate                & multivariate \\
-0 parameter & RV_i         & v0_i, v1_i               & RV0_i, \dots, RVn_i \\
-1 paramter & RV_i(x_i)     & RV0_i(x_i), RV1_i(x_i)    & RV0_i(x_i), \dots, RVn_i(x_i)\\
-N parameters & RV_i(x0_i,\dots,xn_i)     & RV0_i(x0_i,\dots,xn_i), RV1_i(x0_i,\dots,xn_i)    & RV0_i(x0_i,\dots,xn_i), \dots, RVn_i(x0_i,\dots,xn_i)\\
+0 & RV         & RV_0, RV_1               & RV_0, \dots, RV_n \\
+1 & RV(x)     & RV_0(x), RV_1(x)    & RV_0(x), \dots, RV_n(x)\\
+N & RV(x_0,\dots,x_n)     & RV_0(x_0,\dots,x_n), RV_1(x_0,\dots,x_n)    & RV_0(x_0,\dots,xn_i), \dots, RV_n(x_0,\dots,x_n)\\
 \end{tabular}
-\caption{Each $vi$ is a distinct variable such as precipitation or temperature,
-  each $xi$ is a distinct parameter such as time or space, and i is the
-  observation identifier.}
+\caption{Each RV is a distinct variable such as precipitation or temperature,
+  each x is a parameter, such as time or space, on which RV is conditionally
+  dependent. Each cell in the table is distinct observation in the dataset}
 \end{table}

### 2016-12-23 sections/section0d.tex
table and conditionals
histograms and bars

@@
 \subsection{Univariate Observations}
-%%histogram and pdfs/distribution fitting
 
-The specific shape is
-dependent on what sort of density estimation method is used, such as kernal
-estimation or nearest neighbor estimation \cite{chambers1983}. 
-width using the density-trace method-which is a probability density estimation
-via moving windowed average) proposed by Chambers et. al \cite{chambers1983},.
+The bar graph is arguable the simplest way to visualize data that is all of the
+same type drawn from a sample at random. In this visualization technique, the
+heights of the bar are representative of the values, the x valuesare each
+invidual observation and the positioning of the x values is somewhat
+arbitrary. \cite{TufteVisualDIsplay, Friendly, Playfair}. Since bar
+graphs become somewhat useless as the number of samples increase, it is common
+to group (bin) the data and then count the number of elements in each
+been. Both this binning and counting and the subsequent bar chart of the
+frequency are referred to as histograms \cite{ioannidis2003}. While histograms
+give an approximation of the distribution of the data, sometimes the bin widths
+can be too wide and thereby fail to catch naunces in the data such as
+bimodality. A way to counteract this is to build histograms with very thin
+buckets, which is one of the less sophisticated methods of doing kernal density
+estimation. Other methods of approximating the distribution include nearest
+neighbor estimation or the  moving windowed average density trace method \cite{chambers1983}.
 

### 2017-02-07 sections/section0d.tex
scatterplot paper

@@
 same type drawn from a sample at random. In this visualization technique, the
-heights of the bar are representative of the values, the x valuesare each
+heights of the bar are representative of the values, the x values are each
 invidual observation and the positioning of the x values is somewhat
@@
 \subsection{Bivariate Observations}
-\subsection{Multivariate Observations}
+
 While bar graphs and histograms start giving a sense of the variability in
@@
 of the boxplot \cite{wickham2011} Wickham and Stryjewski trace the evolution of
-the boxplot from a difinitive box to amorphous envelopes.  They start by laying
-out
-the core statistics of a boxplot, which can also be seen in
-Figure~\ref{fig:boxplot}.
+the boxplot from definitive boxes to amorphous envelopes.  They start by laying
+out the core statistics of a boxplot, which can also be seen in Figure~\ref{fig:boxplot}.
 \begin{description}
+
 \item[Median (Q2)] 50\% quartile
@@
 boxes. While a histogram would also show that the distribution of state heights
-has a fat righ\
-t tail,
+has a fat right tail,
 the power of the box and whisker plot is in encoding this information such that
@@
 
-\begin{figure}
-  \includegraphics{boxplot}
-  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
-    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
-    shows
-    the distribution of topology of states and
-    the distribution of volcanos. The strength of the box and whisker is that
-    the
-    reader can quickly compare the two distributions and learn, for example,
-    that
-    volcanos tend to span a shorter range of average heights, but that the
-    extremes
-    are much further from the center. Essentially, the distributions act as
-    references for the other displayed distributions.}
-  \label{fig:boxplot}
-\end{figure}
+\subsubsection{Dependent variables}
+%%scatter segue into bagplot and HDR?
+Arguably the simplest and most well known method of visualizing two
+variables against each other is the scatterplot\cite{friendly2005}. In
+a scatter plot, one variable is mapped to the x coordinate and a
+second variable is mapped to the y coordinate,
 
-In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
-Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional
-information of a h\
-istogram and encodes the state heights between the
-25th and 75th percentiles as a box, the median height as a line cutting across
-the heights, and the extreme state heights as lines coming out of the
-boxes. While a histogram would also show that the distribution of state heights
-has a fat righ\
-t tail,
-the power of the box and whisker plot is in encoding this information such that
-it's trivial to compare to it's neighboring distribution of volcano
-heights because they can share the Y axis. As the number of distributions grow,
-this technique becomes invaluable because the number of histograms that can be
-placed next to each other or overlayed as pdfs can quickly become
-unweildy.  Many others authors have kept the basic structure of the boxplot,
-but exploited the structure to convey more information about the data. Authors
-used different quantile levels \cite{hyndman1996}s or measures of outliers
-\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
-  schwertman2004, schwertman\
-  2007}, or otherwise incorporated skewness,
-kurtosis, and other descriptive distributional statistics \cite{kim2004,
-  hubert2008, marmolejo\
-  2015}.
-
-\subsection{Multivariate Observations}
-%% large scale box: letter, spark
-
-\subsection{Relationships Between Variables}
-\subsubsection{Bivariate}
-%%scatter segue into bagplot and HDR?
 \begin{figure}
@@
   \end{figure}
-The boxplot is inherently limited to 1D data; the bagplot was introduced to
+  The boxplot is inherently limited to 1D data; the bagplot was
+  introduced to
   visualize 2D and higher data \cite{rousseeuw1999}
-Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion of
+  Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion
+  of
     boxplot/HDR plot}
@@
 
+%%bubble plot? 
+%% large scale box: letter, spark
+
 %% radial, parallel coordinates, factor plotl

### 2017-02-08 sections/section0d.tex
more fleshing of section 0d bi/multi

@@
 
-The bar graph is arguable the simplest way to visualize data that is all of the
+The bar graph is arguably the simplest way to visualize data that is all of the
 same type drawn from a sample at random. In this visualization technique, the
@@
 a scatter plot, one variable is mapped to the x coordinate and a
-second variable is mapped to the y coordinate,
+second variable is mapped to the y coordinate, and the distribution of the
+points in the x/y plane indicates whether the x and y variables are related to
+each other. While this plot gives a sense of direction, it can fail for larger
+datesets (% insert citations for scatter for larger datasets) and doesn't
+capture the uncertainity of the relationship between the x and y variables. To
+display the distributional variance, the bagplot and bivariate HDR plot were
+created. These plots work in X/Y/Z - bag/box papers tomorrow morning
 
@@
 \subsection{Multivariate Observations}
+If the variables are independent, then small multiples (insert citation) or
+visualizations for a smaller number of variables can be used to visualize
+them. But this technique is limited to a somewhat small number of variables
+(maybe 20 (find paper)) before the number of multiples is overwhelming and
+impedes decision making. Instead, there are some techniques to look at a large
+set of independent variables at once, such as ....
 
-%%bubble plot? 
 %% large scale box: letter, spark
 
+%%bubble plot
 %% radial, parallel coordinates, factor plotl

### 2017-02-15 main.tex
radial/parallel/etc

@@
 \section{Introduction}
+
 High dimensional data is catch-all term, typically used to describe data that

### 2017-05-09 sections/sectionnd.tex
new outline
somehow lost my sections...

--- 2016-09-29 sections/section0d.tex
@@
 \begin{document}
-\section{Independent Observations (0D fields)}
+\section{ND Fields}
+Toolkits, all the Toolkits.
 
+overview-zoom/filter-details-on-demand is the guiding mantra for many of the
+tools built to visiaulzie demnsionally complex data \cite{shneiderman1996eyes}
+because these tools facilitate engagment with the data at the aggregate and
+encourage drilling down to the core. Some of the most widley used of these
+tools are:
+*parview
+*vistrails
+*...
+
+\subsection{Independent Variables}
 \subsection{Univariate Observations}
-%%histogram and pdfs/distribution fitting
+Toolkits, all the Toolkits.
 
+%cite my calender paper here!
+subspace decomposition, projection persuite, manifolds,
+spation temporal viz, yield curve
 \subsection{Bivariate Observations}
-%%introduce box plot
-
-\subsection{Multivariate Observations}
-%% large scale box: letter, spark
 
@@
 \subsubsection{Bivariate}
-%%scatter segue into bagplot and HDR?
-
 \subsection{Multivariate Observations}
-
-%% radial, parallel coordinates, factor plotl
-%%high d-PCA, ICA, markov

### 2017-05-09 paper/sections/section0d.tex
sections are awesome

--- 2017-02-08 sections/section0d.tex
@@
 (maybe 20 (find paper)) before the number of multiples is overwhelming and
-impedes decision making. Instead, there are some techniques to look at a large
+impedes decision making. Instead, there are some techniques to look at a larger
 set of independent variables at once, such as ....
 
+The simiplest of these is the bubble plot (insert citation), which is a simple
+extension of the scatter plot where the size of the scatter point varies
+relative to some ordinal variable. Varying the marker style and colors adds two
+more levels of (insert citation on cramming in all the viz on scatters), and
+the scatter can be 3d, but the maximum number of variables tops out at about 6
+and the graph has a lot of visual noise.
+
+Instead a typical approach is to visualize the piecewise relationships. The
+simpliest of these are a matrix of scatterplots (insert scatter matrix
+citation), but the relationships can get drowned out when there are a large
+number of observations. So instead, a common approach is to use parralel
+coordinates (insert citation) wherein each variable is an axis and lines are
+drawn between the value of variable 1 and the value of variable 2. Bundles of
+lines indicate very common trends between the two variables, and many
+implementations of parallel coordianets have movable axis so users can explore
+each pairwise relationship (insert citation). Radial graphs are a type of 
+parallel coordinates where all the variables are linked to each other; this
+forms a circular type graph wherin the area of the graph is used as an
+indicator of the relationship amongst these variables (insert citation). There
+are a familiy of these sorts of radial visualizations...
+
+%%show picture parallel coordinate
+
+There are more streamlined approaches
+such as  %% radial, parallel coordinates, factor plot, but
+
+Even the best of these visualization techniques are limited to about 5 or 6 variables
 %% large scale box: letter, spark

### 2017-05-09 paper/sections/section1d.tex
sections are awesome

--- 2016-11-03 sections/section1d.tex
@@
 and so it's hard to tease out specific patterns. Instead in
-Figure~\ref{fig:countour}, the aut\
-hors
-remove most of the bands and instead visualize an outlier envelope(light gray)
+Figure~\ref{fig:countour}, the authorsremove most of the bands and instead visualize an outlier envelope(light gray)
 and a more central envelope(dark gray). It retains much of the information of
@@
 
+surface map, functional correlations, planes?
+
 \subsection{Multivariate Observations}
 LDA, CCA, 
-\subsection{Relationships Between Variables}
-
-\subsubsection{Bivariate}
-surface map, functional correlations, planes?
-\subsection{Multivariate Observations}
 PLS, functional PCA

### 2017-05-09 paper/main.tex
intro problem

--- 2017-02-15 main.tex
@@
-\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+==\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
 
@@
 
-\section{Introduction}
-
-High dimensional data is catch-all term, typically used to describe data that
-highly multivariate or data that's structurally complex, or even data that simply has
-a large number of observations. The lack of specificity of the phrase
-`high-demensional` can make it difficult to parse out which high-dimensional
-analysis and visaualization technique is suited to the research question at
-hand. Datasets inherently have two components, the variables and the
-observations of those variables. These oberservations are sometimes dependent
-on information rich parameters such as time and space, and often it is as
-important to visualize these underlying parametric dependencys as the
-relationships between the variables in given observations. Functional data
-analysis provides a language for formalizing the structure of the data in such
-a way that it clarifies the possible interactions in the data.
-
-%%paragraph on functional data analysis
-The underlying assumption of functional data analysis is that while sometimes
-samples are discrete, often they are drawn from a continuos sample of
-obersvations \cite{ramsey, hook, meuller}. That continuous sample is often over
-space or time or another attribute on which the observation is conditional.
-
-\begin{table}%%use conditional notation since it's the same
-\begin{tabular}
-   & univarate   & bivariate                & multivariate \\
-0 & RV         & RV_0, RV_1               & RV_0, \dots, RV_n \\
-1 & RV(x)     & RV_0(x), RV_1(x)    & RV_0(x), \dots, RV_n(x)\\
-N & RV(x_0,\dots,x_n)     & RV_0(x_0,\dots,x_n), RV_1(x_0,\dots,x_n)    & RV_0(x_0,\dots,xn_i), \dots, RV_n(x_0,\dots,x_n)\\
-\end{tabular}
-\caption{Each RV is a distinct variable such as precipitation or temperature,
-  each x is a parameter, such as time or space, on which RV is conditionally
-  dependent. Each cell in the table is distinct observation in the dataset}
-\end{table}
-
-The variables $v0,\dots, vn$ can also potentially be dependent on each other,
-adding in a third level of interactions that a visualization may want to
-caputure. While this survey focuses on data structured in a linear, tabular, or
-data cube way, it could potentially be expanded to network data where the xi
-parameters represent edges in the graph.
-
+\subfile{sections/intro.tex}
+\subfilt{sections/problem.tex}
 \subfile{sections/independent.tex}

### 2017-05-09 paper/sections/problem.tex
intro problem

--- 2017-02-15 main.tex
@@
-\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+\documentclass[../main.text]{subfiles}
+\begin{document}
 
-\usepackage{graphicx}
-\graphicspath{{sections/figs/}{.figs/}}
-
-\usepackage[backend=bibtex, style=numeric-comp]{biblatex}
-\bibliography{main}
-
-\usepackage{subfiles}
-\usepackage{url}
-\usepackage{amsmath}
-
-\title{Visualizing Functional Data}
-\author{Hannah Aizenman}
-\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Huy Vo}
-\submitted{}
-\abstract{There are an almost overwhelming number of machine learning and
-  visualization techniques, but each technique makes some underlying
-  assumptions about how the data is structured. While sometimes the
-  observations are completely independent from each other, in many instances
-  the observations are sampled from 1D continuum like time or 2d surface such
-  as space.This survey uses functional data analysis to classify visualizations
-  based on the number of variables in and level of paramterization of the data.}
-\begin{document}
-\makefrontmatter
-
-\section{Introduction}
-
-High dimensional data is catch-all term, typically used to describe data that
-highly multivariate or data that's structurally complex, or even data that simply has
-a large number of observations. The lack of specificity of the phrase
-`high-demensional` can make it difficult to parse out which high-dimensional
-analysis and visaualization technique is suited to the research question at
-hand. Datasets inherently have two components, the variables and the
-observations of those variables. These oberservations are sometimes dependent
-on information rich parameters such as time and space, and often it is as
-important to visualize these underlying parametric dependencys as the
-relationships between the variables in given observations. Functional data
-analysis provides a language for formalizing the structure of the data in such
-a way that it clarifies the possible interactions in the data.
-
-%%paragraph on functional data analysis
 The underlying assumption of functional data analysis is that while sometimes
 samples are discrete, often they are drawn from a continuos sample of
-obersvations \cite{ramsey, hook, meuller}. That continuous sample is often over
-space or time or another attribute on which the observation is conditional.
+obersvations \cite{ramsey2006, ramsey2002, muller2006}. That continuous sample is often over
+space or time or another attribute on which the observation is conditional. In
+functional data analysis,
 
@@
 
-\subfile{sections/independent.tex}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-\begin{tabular}{|r|r|r|r|}
-  \hline
-              & univariate & bivariate & multivariate\\
-  \hline
-   0D         &            &           &             \\
-  \hline
-   1D         &            &           &             \\
-   \hline
-   ND         &            &           &             \\
-\end{tabular}
-
-\pagebreak
-\printbibliography
 \end{document}

### 2017-05-09 paper/sections/intro.tex
lost my intro

\documentclass[../main.text]{subfiles}
\begin{document}

A high resolution spatio-temporal temperature model and the academic
transcripts of all students at The Graduate Center are both high dimensional
data sets, but they pose intrinsically different data visualization
challenges. The former necessitates visualization techniques that preserve the
spatio-temporal structure, whereas the latter requires a technique for
highlighting the important information in a sea of variables.

%% include retinal variables figure
Bertin's retinal variables codified the language of visualizaton. As
figure~\ref{fig:bertin} describes, qualitiave aspects of a visual symbol -
it's shape, hue, intensity, texture-are varied to display qualitative
differences in the data, and the quantatitive attributes - size, color value -
are manipulated to show quantatiative differences in the data
\cite{bertin1983, krygier2016}. Figure~\ref{fig:bertin} illustrates the
underlying implict distinction between variables and the components of the data
that inform the structure of the data (the geographic plane or the time series
line). In this system, the data's internal structure informs whether it will be
represented as a point, a line, or a plane. While more structural components
can be visualized using shape, size, hue, etc. there's no assumed
representation outside what can be easily drawn on paper.

\cite{wilkinson2006}

### 2017-06-23 paper/sections/discrete.tex
scraps of start of categorical

\documentclass[../main.text]{subfiles}
\begin{document}

%%% throw up a visualization of what this sort of data means

Demographic data is some of the most commonly analyzed in any number of fields,
and often researchers are interested in the joint distributions of different
demographic categories (such as race, income level, and education level). While
a table is encouraged for a small number of variables
\cite{munznertablesgood}), it can get unweildy rather quickly.

Almost regardless of the number of categories, a single categorical variable
is visualized as a bar graph of the frequency of each catgory \cite{friendly:...}

More frequencies? Fourfold display \cite{Fienberg, 1975, Friendly, 1994a,c},
and radial plots \cite{...,....}

Correlational maps....

\cite{friendly:visualizingcategoricaldata}

Categorical attributes can also be mapped to mathematical sets such that an
observation's categorical attributes can also be a set of sets and visualized
as the distribution of set intersections \cite{UpSet, parrallel sets paper}

\end{document}

### 2017-06-23 paper/sections/intro.tex
scraps of start of categorical

@@
 
-A high resolution spatio-temporal temperature model and the academic
-transcripts of all students at The Graduate Center are both high dimensional
-data sets, but they pose intrinsically different data visualization
-challenges. The former necessitates visualization techniques that preserve the
-spatio-temporal structure, whereas the latter requires a technique for
-highlighting the important information in a sea of variables.
+p(X|Y) is normalized so that holding Y fixed and summing over X it sums to 1 oh
+actually they write it P(X|Y)
 
-%% include retinal variables figure
-Bertin's retinal variables codified the language of visualizaton. As
-figure~\ref{fig:bertin} describes, qualitiave aspects of a visual symbol -
-it's shape, hue, intensity, texture-are varied to display qualitative
-differences in the data, and the quantatitive attributes - size, color value -
-are manipulated to show quantatiative differences in the data
-\cite{bertin1983, krygier2016}. Figure~\ref{fig:bertin} illustrates the
-underlying implict distinction between variables and the components of the data
-that inform the structure of the data (the geographic plane or the time series
-line). In this system, the data's internal structure informs whether it will be
-represented as a point, a line, or a plane. While more structural components
-can be visualized using shape, size, hue, etc. there's no assumed
-representation outside what can be easily drawn on paper.
+P is discrete and p is density continuous
 
-\cite{wilkinson2006}
+P is a probability distribution function p is a propability density function
+
+Sheldon M. Ross .... intro to prob models
+Discrete Random Variable Continuous Random Variables
+Categorical Parameters
+Continuous Parameters
+
+Continuous: time, space, time+space
+
+\begin{table}%%use conditional notation since it's the same
+\begin{tabular}
+   & univarate   & bivariate                & multivariate \\
+0 & RV         & RV_0, RV_1               & RV_0, \dots, RV_n \\
+1 & RV(x)     & RV_0(x), RV_1(x)    & RV_0(x), \dots, RV_n(x)\\
+N & RV(x_0,\dots,x_n)     & RV_0(x_0,\dots,x_n), RV_1(x_0,\dots,x_n)    & RV_0(x_0,\dots,xn_i), \dots, RV_n(x_0,\dots,x_n)\\
+\end{tabular}
+\caption{Each RV is a distinct variable such as precipitation or temperature,
+  each x is a parameter, such as time or space, on which RV is conditionally
+  dependent. Each cell in the table is distinct observation in the dataset}
+\end{table}
+
+The variables $v0,\dots, vn$ can also potentially be dependent on each other,
+adding in a third level of interactions that a visualization may want to
+caputure. While this survey focuses on data structured in a linear, tabular, or
+data cube way\cite{munzner data structure}, it could potentially be expanded to network data where the xi
+parameters represent edges in the graph.
+
+\end{document}

### 2017-08-14 paper/sections/categorical.tex
Upset

\documentclass[../main.text]{subfiles}
\begin{document}

%%find a citation that talks about categorical data in set theoretic terms

Categorical data can be likened to set measurment, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets. It is the latter that is analogous to a
conditional probability, the question of the likelihood of membership in one
set given membership in another. And sometimes there is also a quantative
component involved too. For example, in a study of how NEH funding has changed
over time, there are questions of the likelihoods of different disciplines
being funded (\{d_0,\dots,\d_i}\in \D), and if that is conditioned on
  institutione
  type \{i_0, \dots,\i_n}\in I}, and if either of
those are then conditional on time \{t_0, \dots,t_n} \cite{wired} 
%% this can as easily be 311 data, but I don't have a citation for that
 The most common approach is usually to create a list of frequncies, sometimes of
 sets of one category, such as \D, or a table wherein each cell is a an element
 in the product of two (or more sets), such as \D\cross \I creates a table
 wherein cell one is the frequency of elements that are in both $\d_0$ and
 $\i_0$ \cite{find_pivot_table_citation, set theoreitoc categorical data citation}.

 While a table is encouraged for a small number of variables
 \cite{munznertablesgood}), bar graphs are typically used to display
 differences between subsets such as \d_0,\dots, \d_n \cite{riendly_visualizing_2000,,
   ioannidis_history_2003-1, friendly_brief_2006}. They are somewhat limited to
 showing singular frequencies, in that multiple sets can be shown using
 different colors \cite{semiology} but the bars can not really be used to show
 joint set membership.

 \begin{figure}
   \includegraphics{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie
     genres. On the left is the set visualiztion; the columns are the sets and
     the rows are the exclusive intersections or aggregates. On the right is
     the element view, and the scatter plot is a comparison of two sets with
     each dot equaling one element in the set.}
   \label{fig:upsetfig1}
 \end{figure}

The UpSet tool is designed to display set intersections and the aggregate
information of quantative attributes and set membership.
As shown in \ref{fig:upsetfig4}, the Upset tool provides linked set and element views. The set view provides the intersections and their
aggregates, the frequencies of each set and intersection, and the aggregate
statistics of the associated quantative variables, while the element view shows
selected elements and information about their set membership. The Upset tool
also faciliatates comparison of filtered sets via scatterplot, which acts as a
visualization of the attributes of the data conditioned on membership in the
filtered set.

\end{document}

### 2017-08-14 paper/sections/categorical.tex
parallel sets

@@
 over time, there are questions of the likelihoods of different disciplines
-being funded (\{d_0,\dots,\d_i}\in \D), and if that is conditioned on
-  institutione
-  type \{i_0, \dots,\i_n}\in I}, and if either of
-those are then conditional on time \{t_0, \dots,t_n} \cite{wired} 
+being funded (\{d_0,\dots,\d_i}\in \D), and if that is conditioned on institutione
+  type \{i_0, \dots,\i_n}\in I}, and if either of those are then conditional on time \{t_0, \dots,t_n} \cite{wired} 
 %% this can as easily be 311 data, but I don't have a citation for that
  The most common approach is usually to create a list of frequncies, sometimes of
- sets of one category, such as \D, or a table wherein each cell is a an element
+ sets of one category, such as \D, or a crosstable wherein each cell is a an element
  in the product of two (or more sets), such as \D\cross \I creates a table
  wherein cell one is the frequency of elements that are in both $\d_0$ and
- $\i_0$ \cite{find_pivot_table_citation, set theoreitoc categorical data citation}.
+ $\i_0$ \cite{what is cross tabulation?}
 
  While a table is encouraged for a small number of variables
- \cite{munznertablesgood}), bar graphs are typically used to display
- differences between subsets such as \d_0,\dots, \d_n \cite{riendly_visualizing_2000,,
-   ioannidis_history_2003-1, friendly_brief_2006}. They are somewhat limited to
- showing singular frequencies, in that multiple sets can be shown using
- different colors \cite{semiology} but the bars can not really be used to show
- joint set membership.
+ \cite{munznertablesgood}), bar graphs and histograms are typically used to display
+ differences between subsets such as \d_0,\dots, \d_n
+ \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
+ include the hanging rootgram, ord plot, and other distribution oriented
+ plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}.
+
+ While conditional  distributions can somewhat be visualized using Fourfold displays,
+ Mosiac displays, and using logit models, these plots have somewhat limited
+ dimensionality and are somewhat limited to 3 sets of categories
+ \cite{friendly_visualizing_2000}. Another approach is to [insert correlation
+   paper here...]
 
@@
 
-The UpSet tool is designed to display set intersections and the aggregate
-information of quantative attributes and set membership.
-As shown in \ref{fig:upsetfig4}, the Upset tool provides linked set and element views. The set view provides the intersections and their
+The UpSet tool is designed to
+display a much larger number of set intersections and the aggregate
+information of quantative attributes and set membership. \cite{lex_upset:_2014}
+As shown in figure~\ref{fig:upsetfig4}, the Upset tool provides linked set and element views. The set view provides the intersections and their
 aggregates, the frequencies of each set and intersection, and the aggregate
@@
 
+\begin{figure}
+  \includegraphics{parallelsets.png}
+  \caption{On the left is the categorical view, where users can define
+    categories based on other categorical information. For example, in this
+    veiw working family with children is defined as a household with one or
+    more children. On the right is a parallel coordinate graph wherin the
+    dominiant category informs the colors, the size of the box is relative to
+    ghe category size, and the historgrams show the freqiencies of those
+    categories relative to the connecting catgories (market type and income
+    respectively). }
+  \label{fig:upsetfig1}
+\end{figure}
+While UpSet is designed with all sorts of sets in mind, the Parallel Sets tool
+showin in figure~\ref{fig:upsetfig} is expicilitly designed to show frequencies of set membership in heirarichal
+ordered categorical sets, so the \{d_0, \dots, d_n} \in D
+  \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
+  the parallel cooridnates plot \cite{flexible linked axis}, Parallel Sets
+  treats each categorical set independantly but indicates conditional
+  dependency through cross tabulations and by grouping the data, indicated via
+  color, based on an active dimension that then becomes the pivot point for the
+  cross tabulations. In Parallel Sets, quantative data is converted to
+  categorical via binning.
+
 \end{document}

### 2017-08-14 paper/sections/categorical.tex
zhang correlation maps, start

@@
  plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}.
-
- While conditional  distributions can somewhat be visualized using Fourfold displays,
- Mosiac displays, and using logit models, these plots have somewhat limited
- dimensionality and are somewhat limited to 3 sets of categories
- \cite{friendly_visualizing_2000}. Another approach is to [insert correlation
-   paper here...]
+While conditional  distributions can somewhat be visualized using Fourfold displays, Mosiac displays, and using logit models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
+ \cite{friendly_visualizing_2000}.
 
@@
 \end{figure}
-While UpSet is designed with all sorts of sets in mind, the Parallel Sets tool
-showin in figure~\ref{fig:upsetfig} is expicilitly designed to show frequencies of set membership in heirarichal
-ordered categorical sets, so the \{d_0, \dots, d_n} \in D
+
+Another important factor in visualizing categorical data is retaining the
+inherent heirachy in many datasets with categorical measurements
+\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
+showin in figure~\ref{fig:upsetfig} is expicilitly designed to show frequencies
+of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
   \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
@@
 
+\begin{figure}
+
+\end{figure}
+Often though it's important to retain the internel structure of the
+quantative data, such as when it has internal structucture such as time and
+space. Zhang et al propose a method that retains data and allows for
+visualizing more than just pairwise relationships\cite{zhang_visual_2015}. Thei
+
 \end{document}

### 2017-08-14 paper/sections/categorical.tex
close of rough categorical

@@
 
-%%find a citation that talks about categorical data in set theoretic terms
-
 Categorical data can be likened to set measurment, where each category is a set
 and researchers often want to know how many members each set has and what
-intersections are common between sets. It is the latter that is analogous to a
+intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
+  It is the latter that is analogous to a
 conditional probability, the question of the likelihood of membership in one
 set given membership in another. And sometimes there is also a quantative
-component involved too. For example, in a study of how NEH funding has changed
-over time, there are questions of the likelihoods of different disciplines
-being funded (\{d_0,\dots,\d_i}\in \D), and if that is conditioned on institutione
-  type \{i_0, \dots,\i_n}\in I}, and if either of those are then conditional on time \{t_0, \dots,t_n} \cite{wired} 
-%% this can as easily be 311 data, but I don't have a citation for that
- The most common approach is usually to create a list of frequncies, sometimes of
- sets of one category, such as \D, or a crosstable wherein each cell is a an element
+component involved too. The most common approach is usually to create a list of frequncies, sometimes of sets of one category, such as \D, or a crosstable wherein each cell is a an element
  in the product of two (or more sets), such as \D\cross \I creates a table
  wherein cell one is the frequency of elements that are in both $\d_0$ and
- $\i_0$ \cite{what is cross tabulation?}
+ $\i_0$.
 
@@
   \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
-  the parallel cooridnates plot \cite{flexible linked axis}, Parallel Sets
+  the parallel cooridnates plot \cite{claessen_flexible_2011}, Parallel Sets
   treats each categorical set independantly but indicates conditional
@@
 \begin{figure}
+  \includegraphics{corr.png}
+  \label{caption} Zhang et al create an initial map (a) that they then apply an
+  edge filter to (b). They then allow the user to filter out a variable and
+  either highlight only the vertices of that variable (c) or the edges incident
+  on that variable(d).
+  \label{fig:corr}
+\end{figure}
 
-\end{figure}
 Often though it's important to retain the internel structure of the
@@
 space. Zhang et al propose a method that retains data and allows for
-visualizing more than just pairwise relationships\cite{zhang_visual_2015}. Thei
+visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
+introduce a concept of a correlation map, as shown in
+figure~\ref{fig:corr}. The authors build out a graph where every variable is a
+vertex and the edges are correlations between the connecting vertices. The
+vertices are sized relative to the variance in the data. The edges are colored
+red for negative correlations and green for positive correlations. To convert
+the categorical data to numerical space, they developed an algorithm where when
+mapped to numbers, the residual sum of squares of a logostic regression of the
+categorical variables  would be minimized. They then filter edges with low
+correlations and on variables of interest.
 

### 2017-08-19 paper/main.tex
title + consolidation

@@
 
-\title{Visualizing Functional Data}
+\title{Visualizing Conditional Relationshops}
 \author{Hannah Aizenman}
@@
 \submitted{}
-\abstract{There are an almost overwhelming number of machine learning and
-  visualization techniques, but each technique makes some underlying
-  assumptions about how the data is structured. While sometimes the
-  observations are completely independent from each other, in many instances
-  the observations are sampled from 1D continuum like time or 2d surface such
-  as space.This survey uses functional data analysis to classify visualizations
-  based on the number of variables in and level of paramterization of the data.}
+\abstract{The context surrounding data is often as important as the
+  measurements themselves. Income data provides a lens into demographic
+  inequalities when seperated by race and gender; temperature data indicates
+  seasons when plotted by time. These are relatively simple relationships, but
+ many datasets are embedded with rather complex conditional relationships. This
+ survey gives an overview of techniques showing conditional dependencies
+ amongst distributions and densities. 
+}
 \begin{document}
@@
 
-\subfile{sections/intro.tex}
+\section{Introduction}
+\label{sec:introduction}
+p(X|Y) is normalized so that holding Y fixed and summing over X it sums to 1 oh
+actually they write it P(X|Y)
+
+P is discrete and p is density continuous
+
+P is a probability distribution function p is a propability density function
+
+Sheldon M. Ross .... intro to prob models
+Discrete Random Variable Continuous Random Variables
+Categorical Parameters
+Continuous Parameters
+
+Continuous: time, space, time+space
+
+The underlying assumption of functional data analysis is that while sometimes
+samples are discrete, often they are drawn from a continuos sample of
+obersvations \cite{ramsey2006, ramsey2002, muller2006}. That continuous sample is often over
+space or time or another attribute on which the observation is conditional. In
+functional data analysis,
+
+\begin{table}%%use conditional notation since it's the same
+\begin{tabular}
+                & Distribution & Density 
+   Distribution &
+   Density      & 
+\end{tabular}
+
+\begin{table}%%use conditional notation since it's the same
+\begin{tabular}
+   & univarate   & bivariate                & multivariate \\
+0 & RV         & RV_0, RV_1               & RV_0, \dots, RV_n \\
+1 & RV(x)     & RV_0(x), RV_1(x)    & RV_0(x), \dots, RV_n(x)\\
+N & RV(x_0,\dots,x_n)     & RV_0(x_0,\dots,x_n), RV_1(x_0,\dots,x_n)    & RV_0(x_0,\dots,xn_i), \dots, RV_n(x_0,\dots,x_n)\\
+\end{tabular}
+
+\caption{Each RV is a distinct variable such as precipitation or temperature,
+  each x is a parameter, such as time or space, on which RV is conditionally
+  dependent. Each cell in the table is distinct observation in the dataset}
+\end{table}
+
+The variables $v0,\dots, vn$ can also potentially be dependent on each other,
+adding in a third level of interactions that a visualization may want to
+caputure. While this survey focuses on data structured in a linear, tabular, or
+data cube way\cite{munzner data structure}, it could potentially be expanded to network data where the xi
+parameters represent edges in the graph.
+
 \subfilt{sections/problem.tex}
@@
 
-\begin{tabular}{|r|r|r|r|}
-  \hline
-              & univariate & bivariate & multivariate\\
-  \hline
-   0D         &            &           &             \\
-  \hline
-   1D         &            &           &             \\
-   \hline
-   ND         &            &           &             \\
-\end{tabular}
-
 \pagebreak

### 2017-08-19 paper/sections/multivariate.tex
title + consolidation

\begin{figure}

\end{figure}
The parallel coordinates plot \cite{islenberg} shows how different quantative
values are related to each other conditioned on a categorical variable whose
categories are mapped to different colors. As shown in figure~\ref{fig:

### 2017-08-28 paper/sections/multivariate.tex
radial/pcp

@@
+\section{Multivariate Densities Conditioned on Categorical Distributions}
+
 \begin{figure}
+  \includegraphics{flexaxis.png}
+  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates common
+    methods for visualizing multivariate datasets.
+  \label{fig:flexlink}
+\end{figure}
 
-\end{figure}
-The parallel coordinates plot \cite{islenberg} shows how different quantative
-values are related to each other conditioned on a categorical variable whose
-categories are mapped to different colors. As shown in figure~\ref{fig:
+There are a number of methods for showing how measurmeents of different
+variables interact, and as showing figure~\ref{fig:flexlink} color is often
+used to illustrate how these relationships are different for each category of
+data. The colors can also change gradually to indicate changes in a quanitative
+variable %%find citation for pca? with qualitative lines? but it's far less
+%%ocmmon.
+As seen in figure~\ref{fig:flexlink}, the common methods for showing
+visualiztions are matrix of scatters \cite{matric of scatters paper}, parallel
+coordinates \cite{islenberg, wegman}

### 2017-08-30 paper/sections/radial.tex
?

\section{Categorical Distributions from Densities}

Often it is useful to find representative patterns (clusters) from sets of
categorical data. These clusters can in turn be considered categories of the
groupings of quantiative data and the cluster membership yields the
distribution of these categories.

%% PCP's can go here too - relative widths of bands of colors == relative
%% distributions of categories

\begon{figure}
\includegraphics{radial}
\caption{Figrue from \cite{Albo_Bak_}. Rader chart, flower chart, and circle
  charts}
\label{fig:radial}
\end{figure}
Radial visualizations can also be used to show representative
categories from multivariate vectors. For example, as seen in figure~\ref{fig:radial}
rader plot, the clusterings of the area types indicates general categories for
the data, and the number of areas in each cluster indicates the relative
frequency. Flower charts can also be grouped to find these relative
categories, as can the rings in a circle chart.

### 2017-08-30 paper/sections/densdens.tex
spell checked

--- 2017-02-07 sections/section0d.tex
@@
 \begin{document}
-\section{Independent Observations (0D fields)}
+\section{Densities from Densities}
 
-\subsection{Univariate Observations}
-
-The bar graph is arguable the simplest way to visualize data that is all of the
-same type drawn from a sample at random. In this visualization technique, the
-heights of the bar are representative of the values, the x values are each
-invidual observation and the positioning of the x values is somewhat
-arbitrary. \cite{TufteVisualDIsplay, Friendly, Playfair}. Since bar
-graphs become somewhat useless as the number of samples increase, it is common
-to group (bin) the data and then count the number of elements in each
-been. Both this binning and counting and the subsequent bar chart of the
-frequency are referred to as histograms \cite{ioannidis2003}. While histograms
-give an approximation of the distribution of the data, sometimes the bin widths
-can be too wide and thereby fail to catch naunces in the data such as
-bimodality. A way to counteract this is to build histograms with very thin
-buckets, which is one of the less sophisticated methods of doing kernal density
-estimation. Other methods of approximating the distribution include nearest
-neighbor estimation or the  moving windowed average density trace method \cite{chambers1983}.
-
-\subsection{Bivariate Observations}
-
-While bar graphs and histograms start giving a sense of the variability in
-data, they are limited to showing the dispersion of the data only in 1
-dimension; namely the measurements themselves. In their survey of the history
-of the boxplot \cite{wickham2011} Wickham and Stryjewski trace the evolution of
-the boxplot from definitive boxes to amorphous envelopes.  They start by laying
-out the core statistics of a boxplot, which can also be seen in Figure~\ref{fig:boxplot}.
-\begin{description}
-
-\item[Median (Q2)] 50\% quartile
-\item[Inter Quartile Range (IQR)]
-\end{description}
-
-\begin{tabular}{|r| r |r|r|}
-  & Lower & Upper\\
-  \multicolumn{2}{Hinge} & Q1: 25\%  & Q3: 75\% \\
-  \multicolumn{2}{Interquartile range} & \multicolumn{2}{IQR = Q3-Q1}\\
-  \multirow{2}{Fence} & Inner & Q1-1.5*IQR & Q3-1.5*IQR \\
-  & outer & Q1-3*IQR & Q3+3*IQR \\
-  \end{tabular}
+While bar graphs and histograms can give asense of the variability in
+data, they are limited to showing the dispersion of the data only in how the measurements are spread out.
+This is somewhat limiting for very large datasets, so the boxplot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}.
+While a simple boxplot does not inherently show conditional dependencies, many derviations do.
 
@@
 
-In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
-Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional
-information of a h\
-istogram and encodes the state heights between the
-25th and 75th percentiles as a box, the median height as a line cutting across
-the heights, and the extreme state heights as lines coming out of the
-boxes. While a histogram would also show that the distribution of state heights
-has a fat right tail,
-the power of the box and whisker plot is in encoding this information such that
-it's trivial to compare to it's neighboring distribution of volcano
-heights because they can share the Y axis. As the number of distributions grow,
-this technique becomes invaluable because the number of histograms that can be
-placed next to each other or overlayed as pdfs can quickly become
-unweildy.  Many others authors have kept the basic structure of the boxplot,
-but exploited the structure to convey more information about the data. Authors
-used different quantile levels \cite{hyndman1996}s or measures of outliers
+Tukey's introduction to the boxplot \cite{tukey1977}, shown in
+Figure~\ref{fig:boxplot} illustrates how the distributions of these 1d densities differ by category.
+While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data.
+Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
 \cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
@@
   2015}.
-
-\subsubsection{Dependent variables}
-%%scatter segue into bagplot and HDR?
-Arguably the simplest and most well known method of visualizing two
-variables against each other is the scatterplot\cite{friendly2005}. In
-a scatter plot, one variable is mapped to the x coordinate and a
-second variable is mapped to the y coordinate,
-
-\begin{figure}
-  %%\includegraphics{bagplot}
-  \end{figure}
-  The boxplot is inherently limited to 1D data; the bagplot was
-  introduced to
-  visualize 2D and higher data \cite{rousseeuw1999}
-  Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion
-  of
-    boxplot/HDR plot}
-
-\subsection{Multivariate Observations}
-
-%%bubble plot? 
-%% large scale box: letter, spark
-
-%% radial, parallel coordinates, factor plotl
-%%high d-PCA, ICA, markov

### 2017-08-30 paper/sections/densdist.tex
spell checked

--- 2017-08-14 paper/sections/categorical.tex
@@
 \begin{document}
+\section{Densities from Distributions}
 
-Categorical data can be likened to set measurment, where each category is a set
-and researchers often want to know how many members each set has and what
-intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
-  It is the latter that is analogous to a
-conditional probability, the question of the likelihood of membership in one
-set given membership in another. And sometimes there is also a quantative
-component involved too. The most common approach is usually to create a list of frequncies, sometimes of sets of one category, such as \D, or a crosstable wherein each cell is a an element
- in the product of two (or more sets), such as \D\cross \I creates a table
- wherein cell one is the frequency of elements that are in both $\d_0$ and
- $\i_0$.
-
- While a table is encouraged for a small number of variables
- \cite{munznertablesgood}), bar graphs and histograms are typically used to display
- differences between subsets such as \d_0,\dots, \d_n
- \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
- include the hanging rootgram, ord plot, and other distribution oriented
- plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}.
-While conditional  distributions can somewhat be visualized using Fourfold displays, Mosiac displays, and using logit models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
- \cite{friendly_visualizing_2000}.
+Many multivariate visualizations treat the data space as a graph wherin the vertices are the variables and the edge corresponds to some quantified relationship between the two vertices and often these edges are colored by category or form a shape that can be used a proxy for a category.
+These types of visualizations can often be used to derive typical observations in a given category.
 
 \begin{figure}
-   \includegraphics{upsetfig1}
-   \caption{UpSet dashboard displaying the relationships amongst movie
-     genres. On the left is the set visualiztion; the columns are the sets and
-     the rows are the exclusive intersections or aggregates. On the right is
-     the element view, and the scatter plot is a comparison of two sets with
-     each dot equaling one element in the set.}
-   \label{fig:upsetfig1}
+  \includegraphics{flexaxis.png}
+  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
+    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and rader charts.
+  \label{fig:flexlink}
 \end{figure}
 
-The UpSet tool is designed to
-display a much larger number of set intersections and the aggregate
-information of quantative attributes and set membership. \cite{lex_upset:_2014}
-As shown in figure~\ref{fig:upsetfig4}, the Upset tool provides linked set and element views. The set view provides the intersections and their
-aggregates, the frequencies of each set and intersection, and the aggregate
-statistics of the associated quantative variables, while the element view shows
-selected elements and information about their set membership. The Upset tool
-also faciliatates comparison of filtered sets via scatterplot, which acts as a
-visualization of the attributes of the data conditioned on membership in the
-filtered set.
+There are a number of methods for showing how measurments of different
+variables interact, and as showing figure~\ref{fig:flexlink} color is often
+used to illustrate how these relationships are different for each category of
+data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
+visualiztions are matrix of scatters \cite{elmqvist, } , parallel
+coordinates \cite{islenberg, wegman} and rader charts\cite{chambers_tukey}/
 
@@
     respectively). }
-  \label{fig:upsetfig1}
+  \label{fig:parallelsets}
 \end{figure}
@@
 \cite{shneiderman_visualizing_2000}. The Parallel Sets tool
-showin in figure~\ref{fig:upsetfig} is expicilitly designed to show frequencies
+showin in figure~\ref{fig:parallelsets} is expicilitly designed to show frequencies
 of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
@@
   cross tabulations. In Parallel Sets, quantative data is converted to
-  categorical via binning.
+  categorical via binning. Many-to-many PCP is another type of generalized
+  parallel ccoordinates where all possible pairs are explored \cite{lind}.
 
@@
 correlations and on variables of interest.
-
 \end{document}

### 2017-08-30 paper/sections/distdens.tex
spell checked

\section{Distributions from Densities}

Often it is useful to find representative patterns (clusters) from sets of data. These clusters can in turn be considered categories of the groupings of quantitative data and the cluster membership yields the distribution of these categories.

\begin{figure}
\includegraphics{radial}
\caption{Figure from \cite{Albo_Bak_}. Radar chart, flower chart, and circle
  charts}
\label{fig:radial}
\end{figure}
Radial visualizations can be used to show representative
categories from multivariate vectors. For example, as seen in figure~\ref{fig:radial}
radar plot, the clusterings of the area types indicates general categories for
the data, and the number of areas in each cluster indicates the relative
frequency. Flower charts can also be grouped to find these relative
categories, as can the rings in a circle chart. Parallel coordinates charts are often explicitly grouped by category, but the relative frequency of the colors can be used to derive the relative frequencies of each graph.

\begin{figure}
\includegraphics{calendar.png}
\caption{Figure from \cite{van_wijk_van_selow}. They visualized representative clusters of electricty usage and the relative frequency of each cluster using a calendar where day and cluster that day share the same color.}
\label{fig:calendar}
\end{figure}

Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~ref{fig:calendar} shows an analysis of electricity usage in the Netherlands\cite{van_wijk_van_selow}. Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day. They used an agglomerative clustering algorithm \cite{kaufman_rousseeuw} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.

### 2017-08-30 paper/sections/distdist.tex
spell checked

\documentclass[../main.text]{subfiles}
\begin{document}

\section{Distributions}

Categorical data can be likened to set measurement, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
  It is the latter that is analogous to a
conditional probability, the question of the likelihood of membership in one
set given membership in another. And sometimes there is also a quantitative
component involved too. The most common approach is usually to create a list of frequencies, sometimes of sets of one category, such as \D, or a cross-table\cite{goodman} where each cell is a an element in the product of two (or more sets), such as \D\cross \I creates a table
 where cell one is the frequency of elements that are in both $\d_0$ and
 $\i_0$.

 While a table is encouraged for a small number of variables
 \cite{munznertablesgood}), bar graphs and histograms are typically used to display
 differences between subsets such as \d_0,\dots, \d_n
 \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
 include the hanging rootgram, ord plot, and other distribution oriented
 plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional  distributions can somewhat be visualized using Fourfold displays, Mosaic displays, and using logit regression models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 \begin{figure}
   \includegraphics{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie
     genres. On the left is the set visualization; the columns are the sets and
     the rows are the exclusive intersections or aggregates. On the right is
     the element view, and the scatter plot is a comparison of two sets with
     each dot equaling one element in the set.}
   \label{fig:upsetfig}
 \end{figure}

The UpSet tool is designed to
display a much larger number of set intersections and the aggregate
information of quantitative attributes and set membership. \cite{lex_upset:_2014}
As shown in figure~\ref{fig:upsetfig}, the Upset tool provides linked set and element views. The set view provides the intersections and their
aggregates, the frequencies of each set and intersection, and the aggregate
statistics of the associated quantitative variables, while the element view shows
selected elements and information about their set membership. The Upset tool
also facilitates comparison of filtered sets via scatterplot, which acts as a
visualization of the attributes of the data conditioned on membership in the
filtered set.

\end{document}

### 2017-08-30 paper/main.tex
more spellcheck

--- 2017-05-09 paper/main.tex
@@
-==\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
 
@@
 
-\title{Visualizing Functional Data}
+\title{Visualizing Conditional Relationships}
 \author{Hannah Aizenman}
@@
 \submitted{}
-\abstract{There are an almost overwhelming number of machine learning and
-  visualization techniques, but each technique makes some underlying
-  assumptions about how the data is structured. While sometimes the
-  observations are completely independent from each other, in many instances
-  the observations are sampled from 1D continuum like time or 2d surface such
-  as space.This survey uses functional data analysis to classify visualizations
-  based on the number of variables in and level of paramterization of the data.}
+\abstract{The context surrounding data is often as important as the
+  measurements themselves. Income data provides a lens into demographic
+  inequalities when separated by race and gender; temperature data indicates
+  seasons when plotted by time. These are relatively simple relationships, but
+ many datasets are embedded with rather complex conditional relationships. This
+ survey gives an overview of techniques showing conditional dependencies
+ amongst distributions and densities. 
+}
 \begin{document}
@@
 
-\subfile{sections/intro.tex}
-\subfilt{sections/problem.tex}
-\subfile{sections/independent.tex}
+\section{Introduction}
+\label{sec:introduction}
+\begin{figure}
+  \includegraphics{chart_chooser}
+  \caption{From Bertin's Semiology of Graphics \cite{bertin}, this diagram
+    illustrates how there are many ways to display the same information about the 1954 Frnech
+  workforce.}
+  \label{fig:chart_chooser}
+\end{figure}
+
+There are almost an overwhelming number of visualizations, and no real correct
+choice for any given situation. As figure~\ref{fig:chart_chooser} shows, often
+there isn't even a single standout chart for a single dataset; instead the
+choice of figure is dependent on the relationships the researcher would like to
+extract from the dataset. Is it important to know how many people are in each
+department? A bar chart is suited for this, but this data has that information
+segmented into sectors - should it be combined or stay separated? Is the
+spatial distribution of the workforce something that should be preserved? If
+so, how should the sector information be retained? Sectors and space can both
+conditional attributes of the data, yielding information on how the workforce
+varies when disaggregated into these smaller blocks.
+
+Quite often, it is this disaggregation that is interesting to researchers
+because it points to multiple factors that can contribute to the measured
+outcome. The difficulty is in figuring out how to visualize this
+disaggregation, especially as the data structure grows more complex. Cross
+tabulation tables work for a small number of categorical variables, but are
+pushed to the limits at n=3 or 4. And most data is not solely categorical,
+instead often having quantitative densities of one sort or another. A complex but
+important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsey2006, ramsey2002, muller2006}.
+That continuous sample is often over space or time or another parameter or combination thereof, and often it is
+important to preserve those functional dependencies in the visualization. For
+example, for storm modeling it is important to show the expected time and
+location of the rainfall. This survey presents a sampling of visualization
+techniques that attempt to illustrate these conditional and functional
+relationships within a dataset.
+
+\begin{table}
+
+\begin{tabular}
+                & Distribution           & Density 
+   Distribution & P(X|Y),P(A,B,C|E,F,G)  & 
+   Density      &              & p(X|Y), p(A,B,C|E,F,G)
+\end{tabular}
+
+\caption{The rows represent the known distributions & densities while the
+  columns are the distributions and densities that the visualization aims to derive.}
+\end{table}
+
+While this survey focuses on data structured in a linear, tabular, or
+data cube way\cite{munzner data structure}, it could potentially be expanded to network data where the xi
+parameters represent edges in the graph.
+
+\subfile{sections/distdist.tex}
+\subfile{sections/distdens.tex}
+\subfile{sections/densdist.tex}
+\subfile{sections/densdens.tex}
 
@@
 \label{sec:conclusion}
+When trying to choose a visualization technique for a problem, it is often
+crucial to interrogate the questions being asked of the data. What
+relationships are important to preserve and which can be discarded? While there
+are a what feels like infinite number of visualization techniques, this
+visualization task can be rather complicated, especially as the number of
+variables or dependencies increases. Interactive tools can help, but most are limited to showing pairwise or a small number of connected interactions. Machine learning can somewhat be used
+to explore which variables and dimensions, and therefore which relationships, are worth
+exploring further, but often the results of those algorithms need to
+be translated to the visual space and those visualizations need to be
+understandable.
 
-\begin{tabular}{|r|r|r|r|}
-  \hline
-              & univariate & bivariate & multivariate\\
-  \hline
-   0D         &            &           &             \\
-  \hline
-   1D         &            &           &             \\
-   \hline
-   ND         &            &           &             \\
-\end{tabular}
-
-\pagebreak
 \printbibliography

### 2017-08-30 paper/sections/densdens.tex
more spellcheck

@@
 
-While bar graphs and histograms can give asense of the variability in
+While bar graphs and histograms can give a sense of the variability in
 data, they are limited to showing the dispersion of the data only in how the measurements are spread out.
-This is somewhat limiting for very large datasets, so the boxplot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}.
-While a simple boxplot does not inherently show conditional dependencies, many derviations do.
+This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}.
+While a simple box plot does not inherently show conditional dependencies, many derivations do.
 
@@
     the distribution of topology of states and
-    the distribution of volcanos. The strength of the box and whisker is that
+    the distribution of volcanoes. The strength of the box and whisker is that
     the
@@
     that
-    volcanos tend to span a shorter range of average heights, but that the
+    volcanoes tend to span a shorter range of average heights, but that the
     extremes
@@
 
-Tukey's introduction to the boxplot \cite{tukey1977}, shown in
-Figure~\ref{fig:boxplot} illustrates how the distributions of these 1d densities differ by category.
+Tukey's introduction to the box-plot \cite{tukey1977}, shown in
+Figure~\ref{fig:boxplot} illustrates how the distributions of these 1 dimensional densities differ by category.
 While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data.
@@
   2015}.
+
+While box plots were designed to visualize discrete observations, often there
+are no discrete observations and instead the `observation` is a timeseries
+curve (where the connections are as important as the individual points) or
+spatial region. Understanding the distributions of functional observations
+often relies first ordering the functions such that there's some measure of how
+close these functions are to each other--functional depth\cite{febrero2007},
+bivariate score depth\cite{hyndman2009}, and
+bivariate kernel density estimation\cite{scott1992} are some examples. This
+ordering then yields functional analogs to median and other percentile
+metrics. For small values, the ordered curves can then be colored based on the
+ordering metric \cite{hyndman2009}, but this method is unwieldy for larger
+ensembles.
+
+\begin{figure}
+  \includegraphics[width=.75\linewidth]{contour_weather}
+  \caption{From \cite{whitaker2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
+  \label{fig:countour}
+\end{figure}
+Functional bag, box, and HDR plots \cite{hyndman2009, sun2011}} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution.
+A very common task is to compare ensembles, each a function of the same timeseries or geographic region.
+Contour box plots\cite{whitaker2013} aim to capture the variability of the
+uncertainty via a band depth ordering metric. Each ensemble members band depth
+is computed as sum of the probabilities that the observations in any given
+ensemble fall within the max-min envelope defined by any two other
+ensembles. The bands are then sorted by band depth such that the median is the
+ensemble with about 50\% of its members whiten all envelopes formed by other
+bands (so most centered). Outer bands are chosen according to the task at
+hand, which in this paper is temperature visualizations.
+The authors argue that contour box plots are an improvement over spaghetti plots \cite{lau_koa_pang} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
+Instead in
+Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
+the spaghetti, but removes the visual noise of the lines.
+
+One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types.
+The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman1996eyes} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core.
+\end{document}

### 2017-08-30 paper/sections/densdist.tex
more spellcheck

@@
 
-Many multivariate visualizations treat the data space as a graph wherin the vertices are the variables and the edge corresponds to some quantified relationship between the two vertices and often these edges are colored by category or form a shape that can be used a proxy for a category.
+Many multivariate visualizations treat the data space as a graph where the vertices's are the variables and the edge corresponds to some quantified relationship between the two vertexes and often these edges are colored by category or form a shape that can be used a proxy for a category.
 These types of visualizations can often be used to derive typical observations in a given category.
@@
   \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
-    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and rader charts.
+    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
   \label{fig:flexlink}
@@
 
-There are a number of methods for showing how measurments of different
+There are a number of methods for showing how measurements of different
 variables interact, and as showing figure~\ref{fig:flexlink} color is often
@@
 data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
-visualiztions are matrix of scatters \cite{elmqvist, } , parallel
-coordinates \cite{islenberg, wegman} and rader charts\cite{chambers_tukey}/
+visualizations are matrix of scatters \cite{elmqvist, } , parallel
+coordinates \cite{islenberg, wegman} and radar charts\cite{chambers_tukey}/
 
@@
     categories based on other categorical information. For example, in this
-    veiw working family with children is defined as a household with one or
-    more children. On the right is a parallel coordinate graph wherin the
-    dominiant category informs the colors, the size of the box is relative to
-    ghe category size, and the historgrams show the freqiencies of those
-    categories relative to the connecting catgories (market type and income
-    respectively). }
+    view working family with children is defined as a household with one or
+    more children. On the right is a parallel coordinate graph where the
+    dominant category informs the colors, the size of the box is relative to
+    the category size, and the histograms show the frequencies of those
+    categories relative to the connecting categories (market type and income
+    respectively)}
   \label{fig:parallelsets}
@@
 Another important factor in visualizing categorical data is retaining the
-inherent heirachy in many datasets with categorical measurements
+inherent hierarchy in many datasets with categorical measurements
 \cite{shneiderman_visualizing_2000}. The Parallel Sets tool
-showin in figure~\ref{fig:parallelsets} is expicilitly designed to show frequencies
+showin in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
 of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
-  \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
-  the parallel cooridnates plot \cite{claessen_flexible_2011}, Parallel Sets
-  treats each categorical set independantly but indicates conditional
+  \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
+  the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
+  treats each categorical set independently but indicates conditional
   dependency through cross tabulations and by grouping the data, indicated via
   color, based on an active dimension that then becomes the pivot point for the
-  cross tabulations. In Parallel Sets, quantative data is converted to
+  cross tabulations. In Parallel Sets, quantitative data is converted to
   categorical via binning. Many-to-many PCP is another type of generalized
-  parallel ccoordinates where all possible pairs are explored \cite{lind}.
+  parallel coordinates where all possible pairs are explored \cite{lind}.
 
@@
   edge filter to (b). They then allow the user to filter out a variable and
-  either highlight only the vertices of that variable (c) or the edges incident
+  either highlight only the vertexes of that variable (c) or the edges incident
   on that variable(d).
@@
 
-Often though it's important to retain the internel structure of the
-quantative data, such as when it has internal structucture such as time and
+Often though it's important to retain the internal structure of the
+quantitative data, such as when the data is a function of time and
 space. Zhang et al propose a method that retains data and allows for
@@
 figure~\ref{fig:corr}. The authors build out a graph where every variable is a
-vertex and the edges are correlations between the connecting vertices. The
-vertices are sized relative to the variance in the data. The edges are colored
+vertex and the edges are correlations between the connecting vertexes. The
+vertexes are sized relative to the variance in the data. The edges are colored
 red for negative correlations and green for positive correlations. To convert
 the categorical data to numerical space, they developed an algorithm where when
-mapped to numbers, the residual sum of squares of a logostic regression of the
+mapped to numbers, the residual sum of squares of a logistic regression of the
 categorical variables  would be minimized. They then filter edges with low

### 2017-08-30 paper/main.tex
intro pics+bib+table

@@
 \begin{figure}
-  \includegraphics{chart_chooser}
-  \caption{From Bertin's Semiology of Graphics \cite{bertin}, this diagram
+  \includegraphics{chart_chooser.png}
+  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
     illustrates how there are many ways to display the same information about the 1954 Frnech
@@
 instead often having quantitative densities of one sort or another. A complex but
-important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsey2006, ramsey2002, muller2006}.
+important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
 That continuous sample is often over space or time or another parameter or combination thereof, and often it is
@@
 \begin{table}
-
-\begin{tabular}
-                & Distribution           & Density 
-   Distribution & P(X|Y),P(A,B,C|E,F,G)  & 
-   Density      &              & p(X|Y), p(A,B,C|E,F,G)
+\begin{center}
+\begin{tabular}{lll}
+   Variable Type& Distribution           & Density\\ 
+   Distribution & $P(X\|Y),P(A,B,C|E,F,G)$  &  \\
+   Density      &                        & $p(X\|Y), p(A,B,C\|E,F,G)$\\
 \end{tabular}
-
-\caption{The rows represent the known distributions & densities while the
+\caption{The rows represent the known distributions and densities while the
   columns are the distributions and densities that the visualization aims to derive.}
+  \end{center}
 \end{table}
@@
 While this survey focuses on data structured in a linear, tabular, or
-data cube way\cite{munzner data structure}, it could potentially be expanded to network data where the xi
-parameters represent edges in the graph.
+data cube way\cite{munzner_what:_2014}, it could potentially be expanded to network data.
 
-\subfile{sections/distdist.tex}
-\subfile{sections/distdens.tex}
-\subfile{sections/densdist.tex}
-\subfile{sections/densdens.tex}
+%%\subfile{sections/distdist.tex}
+%%\subfile{sections/distdens.tex}
+%%\subfile{sections/densdist.tex}
+%%\subfile{sections/densdens.tex}
 

### 2017-08-30 paper/sections/densdens.tex
intro pics+bib+table

@@
 \section{Densities from Densities}
+
+\begin{figure}
+  \includegraphics{partial_dependence}
+  \caption{Taken from Scikit Learn\cite{scikit_learn}, the line plots show how California housing prices are partially dependent on each x variable.
+The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
+  \label{fig:partialdependence}
+\end{figure}
+
+One of the most straightforward ways to show dependencies is to plot one
+variable against another, but those methods often exclude the effects of the
+other variables in the datasets. Regression methods can be used to analyze the
+ways in which one variable is partially dependent on many other variables in
+the dataset\cite{Hastie_Tibshirani}.
+As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.
 

### 2017-08-30 paper/main.tex
cleaned up bibliographies

@@
 \begin{figure}
-  \includegraphics{chart_chooser.png}
+  \includegraphics[width=\textwidth]{chart_chooser.png}
   \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
-    illustrates how there are many ways to display the same information about the 1954 Frnech
+    illustrates how there are many ways to display the same information about the 1954 French
   workforce.}
@@
 While this survey focuses on data structured in a linear, tabular, or
-data cube way\cite{munzner_what:_2014}, it could potentially be expanded to network data.
+data cube way \cite{munzner_what:_2014}, it could potentially be expanded to network data.
 
-%%\subfile{sections/distdist.tex}
+\subfile{sections/distdist.tex}
 %%\subfile{sections/distdens.tex}

### 2017-08-30 paper/sections/densdens.tex
cleaned up bibliographies

@@
   \includegraphics{partial_dependence}
-  \caption{Taken from Scikit Learn\cite{scikit_learn}, the line plots show how California housing prices are partially dependent on each x variable.
+  \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots show how California housing prices are partially dependent on each x variable.
 The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
@@
 ways in which one variable is partially dependent on many other variables in
-the dataset\cite{Hastie_Tibshirani}.
+the dataset\cite{_elements_2009}.
 As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.
@@
 data, they are limited to showing the dispersion of the data only in how the measurements are spread out.
-This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}.
+This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham_40_2011}.
 While a simple box plot does not inherently show conditional dependencies, many derivations do.
@@
   \includegraphics{boxplot}
-  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
-    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
-    shows
-    the distribution of topology of states and
-    the distribution of volcanoes. The strength of the box and whisker is that
-    the
+  \caption{Tukey's 1977 example of the box plot from
+    Exploratory Data Analytics\cite{tukey_exploratory_1977} illustrates how the box plot shows the distribution of topology of states and
+    the distribution of volcanoes. The strength of the box and whisker is that the
     reader can quickly compare the two distributions and learn, for example,
-    that
-    volcanoes tend to span a shorter range of average heights, but that the
-    extremes
-    are much further from the center. Essentially, the distributions act as
-    references for the other displayed distributions.}
+    that volcanoes tend to span a shorter range of average heights, but that the
+    extremes are much further from the center. Essentially, the distributions act as references for the other displayed distributions.}
   \label{fig:boxplot}
@@
 
-Tukey's introduction to the box-plot \cite{tukey1977}, shown in
+Tukey's introduction to the box-plot \cite{tukey_exploratory_1977}, shown in
 Figure~\ref{fig:boxplot} illustrates how the distributions of these 1 dimensional densities differ by category.
 While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data.
-Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
-\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
-  schwertman2004, schwertman\
-  2007}, or otherwise incorporated skewness,
-kurtosis, and other descriptive distributional statistics \cite{kim2004,
-  hubert2008, marmolejo\
-  2015}.
+Authors used different quantile levels \cite{hyndman_sample_1996}s or measures of outliers
+\cite{frigge_implementations_1989, schwertman_identifying_2007}, or otherwise incorporated skewness, kurtosis, and other descriptive distributional statistics \cite{kim_more_2004, marmolejo-ramos_shifting_2015}.
 
@@
 spatial region. Understanding the distributions of functional observations
-often relies first ordering the functions such that there's some measure of how
-close these functions are to each other--functional depth\cite{febrero2007},
-bivariate score depth\cite{hyndman2009}, and
-bivariate kernel density estimation\cite{scott1992} are some examples. This
+often relies first ordering the functions such that there's some measure of how close these functions are to each other--functional depth\cite{febrero_functional_2007},
+bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
+bivariate kernel density estimation\cite{scott_multivariate_1992} are some examples. This
 ordering then yields functional analogs to median and other percentile
 metrics. For small values, the ordered curves can then be colored based on the
-ordering metric \cite{hyndman2009}, but this method is unwieldy for larger
+ordering metric \cite{rob_j._hyndman_rainbow_2010}, but this method is unwieldy for larger
 ensembles.
@@
   \includegraphics[width=.75\linewidth]{contour_weather}
-  \caption{From \cite{whitaker2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
+  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
   \label{fig:countour}
 \end{figure}
-Functional bag, box, and HDR plots \cite{hyndman2009, sun2011}} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution.
+Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution.
 A very common task is to compare ensembles, each a function of the same timeseries or geographic region.
-Contour box plots\cite{whitaker2013} aim to capture the variability of the
+Contour box plots\cite{whitaker_contour_2013} aim to capture the variability of the
 uncertainty via a band depth ordering metric. Each ensemble members band depth
@@
 hand, which in this paper is temperature visualizations.
-The authors argue that contour box plots are an improvement over spaghetti plots \cite{lau_koa_pang} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
+The authors argue that contour box plots are an improvement over spaghetti plots \cite{luo_visualizing_2003} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
 Instead in
@@
 One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types.
-The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman1996eyes} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core.
+The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman_eyes_1996} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core.
 \end{document}

### 2017-08-30 paper/sections/densdist.tex
cleaned up bibliographies

@@
   \includegraphics{flexaxis.png}
-  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
+  \caption{Figure taken from \cite{claessen_flexible_2011} illustrates the following common
     methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
@@
 data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
-visualizations are matrix of scatters \cite{elmqvist, } , parallel
-coordinates \cite{islenberg, wegman} and radar charts\cite{chambers_tukey}/
+visualizations are matrix of scatters \cite{elmqvist_rolling_2008} , parallel
+coordinates \cite{inselberg_plane_1985, wegman_hyperdimensional_1990} and radar charts\cite{chambers_graphical_1983}.
 
@@
 showin in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
-of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
-  \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
+of set membership in heirarichal ordered categorical sets \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
   the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
@@
   categorical via binning. Many-to-many PCP is another type of generalized
-  parallel coordinates where all possible pairs are explored \cite{lind}.
+  parallel coordinates where all possible pairs are explored \cite{lind_many--many_2009}.
 

### 2017-08-30 paper/sections/distdens.tex
cleaned up bibliographies

@@
 \includegraphics{radial}
-\caption{Figure from \cite{Albo_Bak_}. Radar chart, flower chart, and circle
+\caption{Figure from \cite{albo_off_2016}. Radar chart, flower chart, and circle
   charts}
@@
 \includegraphics{calendar.png}
-\caption{Figure from \cite{van_wijk_van_selow}.
+\caption{Figure from \cite{van_wijk_cluster_1999}.
 They visualized representative clusters of electricty usage and the relative frequency of each cluster using a calendar where day and cluster that day share the same color.}
@@
 This is frequently done through the use of a clustering algorithm.
-Figure~ref{fig:calendar} shows an analysis of electricity usage in the Netherlands\cite{van_wijk_van_selow}.
+Figure~ref{fig:calendar} shows an analysis of electricity usage in the Netherlands\cite{van_wijk_cluster_1999}.
 Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day.
-They used an agglomerative clustering algorithm \cite{kaufman_rousseeuw} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.
+They used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.
+
+Techniques specifically aimed at clustering functional data can be found in the survey by Jacques, Julian and Preda \cite{Jacques_functional_2014}. %%todo: read and write up

### 2017-08-30 paper/sections/distdist.tex
cleaned up bibliographies

@@
-\documentclass[../main.text]{subfiles}
+\documentclass[../main.tex]{subfiles}
 \begin{document}
@@
 component involved too.
-The most common approach is usually to create a list of frequencies, sometimes of sets of one category, such as \D, or a cross-table\cite{goodman} where each cell is a an element in the product of two (or more sets), such as \D\cross \I creates a table
- where cell one is the frequency of elements that are in both $\d_0$ and
- $\i_0$.
+The most common approach is usually to create a list of frequencies, sometimes of sets of one category or a cross-table\cite{goodman_measures_1991} which is an aggregate function applied to the product of two or more sets.
 
  While a table is encouraged for a small number of variables
- \cite{munznertablesgood}), bar graphs and histograms are typically used to display
- differences between subsets such as \d_0,\dots, \d_n
- \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
+ \cite{munzner_what:_2014}), bar graphs and histograms are typically used to display differences between subsets \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
  include the hanging rootgram, ord plot, and other distribution oriented
@@
 
- \begin{figure}
-   \includegraphics{upsetfig1}
-   \caption{UpSet dashboard displaying the relationships amongst movie
-     genres. On the left is the set visualization; the columns are the sets and
-     the rows are the exclusive intersections or aggregates. On the right is
-     the element view, and the scatter plot is a comparison of two sets with
-     each dot equaling one element in the set.}
-   \label{fig:upsetfig}
- \end{figure}
+ %%\begin{figure}
+  %% \includegraphics[width=\textwidth]{upsetfig1}
+   %%\caption{UpSet dashboard displaying the relationships amongst movie
+  %%genres. On the left is the set visualization; the columns are the sets and
+    %% the rows are the exclusive intersections or aggregates. On the right is
+    %% the element view, and the scatter plot is a comparison of two sets with
+     %%each dot equaling one element in the set.}
+   %%\label{fig:upsetfig}
+ %%\end{figure}
 

### 2017-08-31 paper/main.tex
new abstract

@@
 \submitted{}
-\abstract{The context surrounding data is often as important as the
-  measurements themselves. Income data provides a lens into demographic
-  inequalities when separated by race and gender; temperature data indicates
-  seasons when plotted by time. These are relatively simple relationships, but
- many datasets are embedded with rather complex conditional relationships. This
- survey gives an overview of techniques showing conditional dependencies
- amongst distributions and densities. 
-}
+\abstract{Many datasets have qualitative and quantitative variables and  also
+  often have functional attributes like time and location and visualizations
+  provide views on a subset of these variables and how they relate to each
+  other. For example a timeseries plot will keep the link between the variable
+  and time but lose the information encoded in the space variable. These
+  relationships can be modeled as conditional probabilities  such as P(V|t),
+  and visualizations can be organized based on which of these probabilities are
+  implicitly preserved in the visualization techniques. Modern visualizations
+  in particular try to tackle the difficulties of displaying these functional
+  relationships as the number of variables and the number of functional
+  dependencies (for example lat, lon, height, time, and type) increase. }
 \begin{document}
@@
 \subfile{sections/distdist.tex}
-%%\subfile{sections/distdens.tex}
-%%\subfile{sections/densdist.tex}
-%%\subfile{sections/densdens.tex}
+\subfile{sections/distdens.tex}
+\subfile{sections/densdist.tex}
+\subfile{sections/densdens.tex}
 

### 2017-08-31 paper/sections/densdist.tex
new abstract

@@
 \begin{document}
-\section{Densities from Distributions}
+\section{Quantitative Densities Conditioned on Qualitative Distributions}
 
@@
 \begin{figure}
-  \includegraphics{flexaxis.png}
-  \caption{Figure taken from \cite{claessen_flexible_2011} illustrates the following common
-    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
+  \includegraphics[width=\textwidth]{flexaxis.png}
+  \caption{This figure taken from a paper on improving multivariate data visualization using flexible linked axis \cite{claessen_flexible_2011} shows the common techniques of matrix of scatters, parallel coordinates and radar charts.}
   \label{fig:flexlink}
@@
 \begin{figure}
-  \includegraphics{parallelsets.png}
+  \includegraphics[width=\textwidth]{parallelsets.png}
   \caption{On the left is the categorical view, where users can define
@@
 \begin{figure}
-  \includegraphics{corr.png}
-  \label{caption} Zhang et al create an initial map (a) that they then apply an
+  \includegraphics[width=\textwidth]{corr.png}
+  \caption{Zhang et al create an initial map (a) that they then apply an
   edge filter to (b). They then allow the user to filter out a variable and
   either highlight only the vertexes of that variable (c) or the edges incident
-  on that variable(d).
+  on that variable(d).}
   \label{fig:corr}
@@
 space. Zhang et al propose a method that retains data and allows for
-visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
+visualizing more than just pairwise relationships \cite{zhang_visual_2015}. They
 introduce a concept of a correlation map, as shown in

### 2017-08-31 paper/sections/distdist.tex
new abstract

@@
 
-\section{Distributions}
+\section{Qualitative Variables Conditioned on Qualitative Distributions}
 
@@
 
- %%\begin{figure}
-  %% \includegraphics[width=\textwidth]{upsetfig1}
-   %%\caption{UpSet dashboard displaying the relationships amongst movie
-  %%genres. On the left is the set visualization; the columns are the sets and
-    %% the rows are the exclusive intersections or aggregates. On the right is
-    %% the element view, and the scatter plot is a comparison of two sets with
-     %%each dot equaling one element in the set.}
-   %%\label{fig:upsetfig}
- %%\end{figure}
+\begin{figure}
+  \includegraphics[width=\textwidth]{upsetfig1}
+   \caption{UpSet dashboard displaying the relationships amongst movie genres. On the left is the set visualization; the columns are the sets and
+   the rows are the exclusive intersections or aggregates. On the right is the element view, and the scatter plot is a comparison of two sets with each dot equaling one element in the set.}
+   \label{fig:upsetfig}
+\end{figure}
 
-The UpSet tool is designed to
-display a much larger number of set intersections and the aggregate
-information of quantitative attributes and set membership. \cite{lex_upset:_2014}
-As shown in figure~\ref{fig:upsetfig}, the Upset tool provides linked set and element views. The set view provides the intersections and their
+The UpSet tool is designed to display a much larger number of set intersections and the aggregate information of quantitative attributes and set membership.
+\cite{lex_upset:_2014} As shown in figure~\ref{fig:upsetfig}, the Upset tool provides linked set and element views.
+The set view provides the intersections and their
 aggregates, the frequencies of each set and intersection, and the aggregate

### 2017-09-27 paper/graphtable.tex
new abstract + edits

% Please add the following required packages to your document preamble:
% \usepackage{mcr}
% \usepackage{graphicx}

\let\mcn\multicolumn
\let\mcr\multirow

\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{ccl|l|l|l|l|}
\cline{4-7}
\mcn{1}{l}{}                                  & \mcn{1}{l}{}                              &       & \mcn{4}{c|}{Probability of}                                             \\ \cline{4-7} 
\mcn{1}{l}{}                                  & \mcn{1}{l}{}                              &       & \mcn{2}{c|}{Independent} & \mcn{2}{c|}{Dependent (Dimensional)} \\ \cline{4-7} 
\mcn{1}{l}{}                                  & \mcn{1}{l}{}                              &       & Categorical     & Quantative     & Categorical   & Quantative           \\ \hline
\mcn{1}{|c|}{\mcr{4}{*}{Conditioned On}} & \mcn{1}{c|}{\mcr{2}{*}{Independent}} 		 & Cat   &  a diagram, one diagram,  &   three diagrams, one diagram,   &       &                  \\ \cline{3-7} 
\mcn{1}{|c|}{}                                & \mcn{1}{c|}{}                            & Quant &  scatterplot, triangular diagram      &  distribution            &                 &  a cartogram
                      \\ \cline{2-7} 
\mcn{1}{|c|}{}                                & \mcn{1}{c|}{\mcr{2}{*}{Dependent}}        & Cat   &                 &             &    circles (piecharts)  & four maps                       \\ \cline{3-7} 
\mcn{1}{|c|}{}                                & \mcn{1}{c|}{}                             & Quant &   one map     &             &     columns & a map                       \\ \hline
\end{tabular}%
}
\caption{Classification of Bertin's figures on the bases of the conditional dependencies shown in the figures.}
\label{}
\end{table}

### 2017-09-27 paper/main.tex
notes and commitee
new abstract + edits

--- 2017-08-30 paper/main.tex
@@
 
+\usepackage{array,multirow,graphicx}
 \usepackage{subfiles}
@@
 
-\title{Visualizing Conditional Relationships}
+\title{Visualizing Conditional Dependencies}
 \author{Hannah Aizenman}
-\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Huy Vo}
+\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
 \submitted{}
-\abstract{The context surrounding data is often as important as the
-  measurements themselves. Income data provides a lens into demographic
-  inequalities when separated by race and gender; temperature data indicates
-  seasons when plotted by time. These are relatively simple relationships, but
- many datasets are embedded with rather complex conditional relationships. This
- survey gives an overview of techniques showing conditional dependencies
- amongst distributions and densities. 
+\abstract{Many datasets are too large or complex to discern patterns simply through visualizing the observations in them, so researchers instead compute the distributions of select variables and explore under what conditions they occur.
+While there are many visualization techniques that preserve variable type, dimensionality, and mutlivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships.
+This survey presents a sampling of techniques for visualizing distributions and densities conditioned on variables that are categorical or quantitative, discrete or continuous, and potentially have or infer dimensional dependencies.
 }
@@
 \label{sec:introduction}
+
 \begin{figure}
   \includegraphics[width=\textwidth]{chart_chooser.png}
-  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
-    illustrates how there are many ways to display the same information about the 1954 French
-  workforce.}
+  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram illustrates various ways of representing information about the French workforce in 1954.}
   \label{fig:chart_chooser}
@@
 
-There are almost an overwhelming number of visualizations, and no real correct
-choice for any given situation. As figure~\ref{fig:chart_chooser} shows, often
-there isn't even a single standout chart for a single dataset; instead the
-choice of figure is dependent on the relationships the researcher would like to
-extract from the dataset. Is it important to know how many people are in each
-department? A bar chart is suited for this, but this data has that information
-segmented into sectors - should it be combined or stay separated? Is the
-spatial distribution of the workforce something that should be preserved? If
-so, how should the sector information be retained? Sectors and space can both
-conditional attributes of the data, yielding information on how the workforce
-varies when disaggregated into these smaller blocks.
+In Semiology of Graphics, Bertin \cite{bertin_semiology_2011} presents a dataset of the French work force in 1954.
+This dataset aggregates the workers by region and by sector, and so is structured as follows:
+\begin{itemize}
+	\item department (geographic region of France)
+	\item  number of employees per thousands in each sector
+		\begin{itemize}
+			\item primary: agriculture
+			\item secondary: industry 
+			\item tertiary: commerce, transportation, services
+		\end{itemize}
+\end{itemize}
 
-Quite often, it is this disaggregation that is interesting to researchers
-because it points to multiple factors that can contribute to the measured
-outcome. The difficulty is in figuring out how to visualize this
-disaggregation, especially as the data structure grows more complex. Cross
-tabulation tables work for a small number of categorical variables, but are
-pushed to the limits at n=3 or 4. And most data is not solely categorical,
-instead often having quantitative densities of one sort or another. A complex but
-important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
-That continuous sample is often over space or time or another parameter or combination thereof, and often it is
-important to preserve those functional dependencies in the visualization. For
-example, for storm modeling it is important to show the expected time and
-location of the rainfall. This survey presents a sampling of visualization
-techniques that attempt to illustrate these conditional and functional
-relationships within a dataset.
+In figure~\ref{fig:chart_chooser}, Bertin explores some potential visualizations of the data, and in doing so also illustrates the trade-offs these visualizations make.
+The first bar graph in the tree shows the number of works per sector, but discards the categorical and encodes the spatial as each bar.
+Categories can be included using a stacked bar chart or three independent bar charts; the former  better preserves the proportionality between sectors within each region, while the latter better displays the regional distribution within each sector.
+Discarding the geographic attributes allows even more interrogation of the relationships within and between categories, via histograms and scatter plots respectively.
 
-\begin{table}
-\begin{center}
-\begin{tabular}{lll}
-   Variable Type& Distribution           & Density\\ 
-   Distribution & $P(X\|Y),P(A,B,C|E,F,G)$  &  \\
-   Density      &                        & $p(X\|Y), p(A,B,C\|E,F,G)$\\
-\end{tabular}
-\caption{The rows represent the known distributions and densities while the
-  columns are the distributions and densities that the visualization aims to derive.}
-  \end{center}
-\end{table}
+While some of these approaches include the department as a the y variable, to some extent they discard the geographic information encoded in the department because this data has a physical counterpart that is not referenced in the visualization.
+To explicitly show this geographic information, Bertin proposes a second branch of map based visualizations.
+Taking the same sort of path as the upper branch, Bertin first proposes a map of how the entire workforce is distributed spatially over France.
+He then suggests a cartogram to show relative differences between geographic regions.
+As in the upper branch, he then incorporates the sector information; he proposes bubble charts and bar charts to highlight inter-sector differences and disaggregates the sectors and visualizes them as different maps to show how each sector is distributed over France.
+I
 
-While this survey focuses on data structured in a linear, tabular, or
-data cube way \cite{munzner_what:_2014}, it could potentially be expanded to network data.
+In these visualizations, not only is Bertin showing how many people are in the French workforce in 1954, but he is also showing how that number is conditionally dependent on categorical information such as sector and department.
+With respect to department, he also tries to preserve the geographic distribution of the data.
+Figure~\ref{fig:chart_chooser} is mostly showing the number of employees given department, category, or both.
+While disaggregating employees over sector and geography provides a rich set of potential factors in employee count, it is also somewhat difficult to visualize if the goal is to preserve both regional and categorical differences.
+While 4 maps is a small enough number to not be overwhelming, they forewarn that this technique would likely not work for a larger number of sectors.
+
+Despite ostensibly being a categorical variable, department can also be a dimension \cite{munzner_what:_2014} on which either number of employees or sector can depend since it maps to physical space rather than a measurement.
+Number of employees and the sector are traditional quantitative and categorical measurement types.
+This distinction of variable type serves as a basis for understanding what conditional dependencies are explored within each visualization:
+
+\input{graphtable.tex}
+
+This survey presents a sampling of visualization techniques that attempt to visually represent conditional dependencies between a variety of variables, including:
+\begin{enumerate}
+	\item quantitative and categorical variables
+	\item distributions and densities
+	\item variables with dimensional dependencies
+\end{enumerate}
 
 \subfile{sections/distdist.tex}
-%%\subfile{sections/distdens.tex}
-%%\subfile{sections/densdist.tex}
-%%\subfile{sections/densdens.tex}
+\subfile{sections/distdens.tex}
+\subfile{sections/densdist.tex}
+\subfile{sections/densdens.tex}
 

### 2017-09-27 paper/sections/distdens.tex
new abstract
new abstract + edits

@@
-\section{Distributions from Densities}
-
-Often it is useful to find representative patterns (clusters) from sets of data.
-These clusters can in turn be considered categories of the groupings of quantitative data and the cluster membership yields the distribution of these categories.
+\section{Categorical Variables Conditioned on Quantitative Densities}
 
 \begin{figure}
-\includegraphics{radial}
+\includegraphics[width=1\textwidth]{radial}
 \caption{Figure from \cite{albo_off_2016}. Radar chart, flower chart, and circle
@@
 \begin{figure}
-\includegraphics{calendar.png}
+\includegraphics[width=1\textwidth]{calendar.png}
 \caption{Figure from \cite{van_wijk_cluster_1999}.

### 2018-01-31 paper/sections/catKcatV.tex
full draft of intro

--- 2017-08-31 paper/sections/distdist.tex
@@
 
-\section{Qualitative Variables Conditioned on Qualitative Distributions}
+\section{Categorical Variables Conditioned on Categorical Distributions}
 
@@
 statistics of the associated quantitative variables, while the element view shows
-selected elements and information about their set membership. The Upset tool
+selected elements and information about their set membership.
+The set view yields the distributions of each of the attributes.
+Far example the selected row (in the gray box) are the distributions of each quantative variable in the comedy set.
+Every row underneath is the distribution of that quantative variable restricted to movies in the intersection of the selected sets.
+For example, the distribution of ratings of movies that are both comedies and dramas.
+The Upset tool
 also facilitates comparison of filtered sets via scatterplot, which acts as a

### 2018-01-31 paper/sections/quantKquantV.tex
full draft of intro

--- 2017-08-30 paper/sections/densdens.tex
@@
 \begin{document}
-\section{Densities from Densities}
+\section{Quantitative Variables Conditioned on Quantitative Densities}
 
 \begin{figure}
-  \includegraphics{partial_dependence}
+  \includegraphics[width=\textwidth]{partial_dependence}
   \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots show how California housing prices are partially dependent on each x variable.
@@
 While box plots were designed to visualize discrete observations, often there
-are no discrete observations and instead the `observation` is a timeseries
-curve (where the connections are as important as the individual points) or
+are no discrete observations and instead the `observation` is a functional densities, drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
+One such example is timeseries curve (where the connections are as important as the individual points) or
 spatial region. Understanding the distributions of functional observations
-often relies first ordering the functions such that there's some measure of how close these functions are to each other--functional depth\cite{febrero_functional_2007},
+often relies first ordering the functions such that there's some measure of how close these functions are to each other--depth\cite{febrero_functional_2007},
 bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
@@
   \includegraphics[width=.75\linewidth]{contour_weather}
-  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
+  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the trajectory of of the data.}
   \label{fig:countour}

### 2018-01-31 paper/main.tex
compiled clean intro

--- 2017-08-30 paper/main.tex
@@
 \usepackage[backend=bibtex, style=numeric-comp]{biblatex}
-\bibliography{main}
+\bibliography{glasslab_viz}
 
+\usepackage{array,multirow,graphicx}
 \usepackage{subfiles}
@@
 \label{sec:introduction}
+Many datasets are multivariate and often contain a a mix of quantative and categorical variables.
+The Iris dataset \cite{fisher_use_1936-1, _uci_????} is a well known example of the type of data commonly found in machine learning literature.
+It is so wildely used as a sample dataset because it has 150 observations that have both quantative and categorical attributes.
+The version visualized in the introduction to this paper is obtained from the scikit-learn Python machine learning library \cite{scikit-learn}.
+The Iris dataset contains the following attributes:\\
+\begin{tabular} {lllll}
+Sepal length (cm) & Sepal width (cm) & Petal length (cm) & Petal width (cm) & Species name
+\end{tabular}
+\\
+A first step in exploring the dataset is typically to make a visualization, but of what sort? Bertin posits that all visualizations at their core have two parts \cite{bertin_semiology_2011}:
+\begin{description}
+\item[invariant] - invariable common ground
+\item[components] - a finite set of variational concepts
+\end{description}
+
 \begin{figure}
-  \includegraphics[width=\textwidth]{chart_chooser.png}
-  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
-    illustrates how there are many ways to display the same information about the 1954 French
-  workforce.}
-  \label{fig:chart_chooser}
+	\includegraphics[width=\textwidth]{intro/iris_pie}
+  	\caption{Relative amounts of each Iris species in the scikit-learn dataset\cite{scikit-learn}.}
+  	\label{fig:iris_pie}
 \end{figure}
 
-There are almost an overwhelming number of visualizations, and no real correct
-choice for any given situation. As figure~\ref{fig:chart_chooser} shows, often
-there isn't even a single standout chart for a single dataset; instead the
-choice of figure is dependent on the relationships the researcher would like to
-extract from the dataset. Is it important to know how many people are in each
-department? A bar chart is suited for this, but this data has that information
-segmented into sectors - should it be combined or stay separated? Is the
-spatial distribution of the workforce something that should be preserved? If
-so, how should the sector information be retained? Sectors and space can both
-conditional attributes of the data, yielding information on how the workforce
-varies when disaggregated into these smaller blocks.
+\begin{figure}
+\includegraphics[width=1\textwidth]{intro/retinal_variables.png}
+\caption{Bertin's retinal variables are a codification of how position, size, shape, color and texture can be used to illustrate variations in the components of a visualization.
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} reproduced from Making Maps: A Visual Guide to Map Design for GIS \cite{krygier_making_2005}}
+\label{fig:retinal_variables}
+\end{figure}
 
-Quite often, it is this disaggregation that is interesting to researchers
-because it points to multiple factors that can contribute to the measured
-outcome. The difficulty is in figuring out how to visualize this
-disaggregation, especially as the data structure grows more complex. Cross
-tabulation tables work for a small number of categorical variables, but are
-pushed to the limits at n=3 or 4. And most data is not solely categorical,
-instead often having quantitative densities of one sort or another. A complex but
-important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
-That continuous sample is often over space or time or another parameter or combination thereof, and often it is
-important to preserve those functional dependencies in the visualization. For
-example, for storm modeling it is important to show the expected time and
-location of the rainfall. This survey presents a sampling of visualization
-techniques that attempt to illustrate these conditional and functional
-relationships within a dataset.
+In figure~\ref{fig:iris_pie}, the invariant is the count of species and the components are the species themselves.
+In other words, the invariant is the value or calculation being visualized, and the components are usually the values on the axis and any variable encoded by what Bertin terms a retinal variable \cite{bertin_semiology_2011,krygier_making_2005}, which are the position, size, shape, color, and texture of the points, lines, or polygons in any visualization.
+As shown in figure~\ref{fig:retinal_variables}, Bertin recommends that quantative components be represented by retinal variables that change quantatively and that categorical components be represented by retinal variables that vary qualitatively.
+The small number of retinal variables suggests that visualizations are somewhat limited in how many components can be shown on any one graph.
+
+\begin{figure}
+ \includegraphics[width=\textwidth]{intro/munzner_datatypes}
+\caption{Keys are unique lookup values used to find individual values in the dataset. Image modified from a diagram from Munzner's website \cite{_visualization_????}}
+\label{fig:munzner_datatypes}
+\end{figure}
+
+A special class of data variables are those that provide situational information about an observational measurement, such as the time and place the observation was recorded.
+Tamara Munzner provides a classification system for thinking about the semantics of a variable \cite{munzner_what_2014}:
+\begin{description}
+\item[value] measurement of interest 
+\item[key] uniqe index to look up value
+\end{description}
 
@@
 \begin{center}
-\begin{tabular}{lll}
-   Variable Type& Distribution           & Density\\ 
-   Distribution & $P(X\|Y),P(A,B,C|E,F,G)$  &  \\
-   Density      &                        & $p(X\|Y), p(A,B,C\|E,F,G)$\\
+\begin{tabular}{ l c r }
+  Munzner & Statistics & Data Warehousing \\
+  \hline
+  key & independent attribute & dimension \\
+  value & dependent attribute & measurement\\
 \end{tabular}
-\caption{The rows represent the known distributions and densities while the
-  columns are the distributions and densities that the visualization aims to derive.}
+\caption{Munzner's key, value designations are roughly analogous to independent and dependent semantics in statistics and what are commonly termed dimensions and measurements in the computer science community.}
+\label{table:munzner_semantics}
 \end{center}
@@
 
-While this survey focuses on data structured in a linear, tabular, or
-data cube way \cite{munzner_what:_2014}, it could potentially be expanded to network data.
+As displayed in table~\ref{table:munzner_semantics}, keys and values are analogous to terms used within the statistics and data communities; the advantage of using keys and values is that they strictly refer to the real world semantics of the data\cite{munzner_what_2014}.
+Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset.
+In a table or a database, it would be the primary key for a row, and in a map it could be an address or geopgraphic coordinates.
+In a multidimensional table, it could be a unique set of attributes-such as a timestamp, latitude, and longitude-associated with each measurment.
 
-\subfile{sections/distdist.tex}
+Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical location that are sampled from a continous curve, surface, or field.
+While these observations are discrete samples from the continous space, often the continuinous (functional) characteristic\cite{ramsay_functional_2006,muller_functional_2006} of the observational space is what is of interest.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{intro/iris_parallel}
+\caption{The parallel coordinates plot (PCP) of the Iris dataset shows the pairwise relationship between adjacent variables, for example sepal length and sepal width.
+The PCP plot can be used as proxy for illustrating conditional probability because it shows how the quantative measurements distribute amongst the species.
+This is is seen both in the relative widths of each band of color and in the similarites in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands.}
+\label{fig:iris_parallel}
+\end{figure}
+
+While the Iris dataset does not have any variables that seem to traditionally fit into the model of keys-as it has no time or space variables-an analogous look-up task for the Iris dataset is to filter on a component and then look at the probabiluty distributions of the other variables in relation tp that key.
+In effect this treats all variables as potential keys into distributions of the other variables.
+In the parallel coordinates plot (PCP) in figure~\ref{fig:iris_parallel.png}, each line is an observation in the dataset, and each color maps to a species of Iris.
+The color coding then allows for this visualization to be used to show how the four quantative variables (sepal length, sepal width, petal length, and petal width) distribute within each variable.
+
+The visualization then becomes a way of visualizing the conditional probability of A given B:
+\begin{equation}
+P(A\mid B) = \frac{P(A \cap B}{P(B)} 
+\end{equation}
+
+wherein A is the observation vector $\left< \text{sepal length, sepal width, petal length, petal width} \right>$ and B in turn takes on each species of petal.
+The locations of the color and width of the colored band act as proxies for the distribution of each quantative variable and the overall observation vector.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{intro/iris_observations.png}
+\caption{The diaganol of the scatter matrix shows the distribution of each variable relative to the species type.
+This can be used to illustrate the conditional probability of each variable relative to a type of species.
+The scatter matrix plot shows every set of pairwise relationships in the dataset, color code by Iris species.
+Because the scatter matrix shows the cooccurance of two variables and is encoded with information about a third variable, it works well was a tool to explore a potential conditional dependency amongst the variables.}
+\label{fig:iris_observations}
+\end{figure}
+
+As exemplified by the scatter matrix in figure~\ref{fig:iris_observations}, there are a number of visualization techniques that can be used to explore conditional dependencies, but these methods are limited to datasets where the observations are discrete.
+There are visualization techniques for understanding the probability distributions of functional observations that could potentially be leverged in conjunction with these methods to develop methods for visualizing conditional dependencies in a functional space.
+This paper provides a sampling of the following visualizations:
+\begin{enumerate}
+\item methods for visualizing conditional dependency for discrete observations
+\item methods for visualizing uncertainity and probabilistic visualizations for functional observations
+\end{enumerate}
+
+%%\subfile{sections/distdist.tex}
 %%\subfile{sections/distdens.tex}
@@
 \label{sec:conclusion}
-When trying to choose a visualization technique for a problem, it is often
-crucial to interrogate the questions being asked of the data. What
-relationships are important to preserve and which can be discarded? While there
-are a what feels like infinite number of visualization techniques, this
-visualization task can be rather complicated, especially as the number of
-variables or dependencies increases. Interactive tools can help, but most are limited to showing pairwise or a small number of connected interactions. Machine learning can somewhat be used
-to explore which variables and dimensions, and therefore which relationships, are worth
-exploring further, but often the results of those algorithms need to
-be translated to the visual space and those visualizations need to be
-understandable.
+
+1 Densities and distributions are important
+2 Visualization methods fall off as variables get more
+complex
+3 Can we build on current methods to show conditional
+dependencies and functional variables?
 

### 2018-01-31 paper/main.tex
newest version + repo clean up

@@
 \documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+
+\usepackage[utf8]{inputenc}
+\usepackage{tikz}
+\usetikzlibrary{mindmap}
 
@@
 \usepackage{array,multirow,graphicx}
+\usepackage{subcaption}
 \usepackage{subfiles}
@@
 \abstract{Many datasets are too large or complex to discern patterns simply through visualizing the observations in them, so researchers instead compute the distributions of select variables and explore under what conditions they occur.
-While there are many visualization techniques that preserve variable type, dimensionality, and mutlivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships.
+While there are many visualization techniques that preserve variable type, dimensionality, and multivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships.
 This survey presents a sampling of techniques for visualizing distributions and densities conditioned on variables that are categorical or quantitative, discrete or continuous, and potentially have or infer dimensional dependencies.
@@
 \end{tabular}
-\\
+
+\subsection{Formalization of Visualization Terms}
+
 A first step in exploring the dataset is typically to make a visualization, but of what sort? Bertin posits that all visualizations at their core have two parts \cite{bertin_semiology_2011}:
@@
 
+\subsubsection{Bertin's Retinal Variables}
 \begin{figure}
@@
 The small number of retinal variables suggests that visualizations are somewhat limited in how many components can be shown on any one graph.
+
+\subsubsection{Munzner's data semantics}
 
@@
 
+\subsubsection{Conditional Probability}
 \begin{figure}
@@
 \caption{The parallel coordinates plot (PCP) of the Iris dataset shows the pairwise relationship between adjacent variables, for example sepal length and sepal width.
-The PCP plot can be used as proxy for illustrating conditional probability because it shows how the quantative measurements distribute amongst the species.
-This is is seen both in the relative widths of each band of color and in the similarites in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands.}
+The PCP plot can be used as proxy for illustrating conditional probability because it shows how the quantitative measurements distribute amongst the species.
+This is is seen both in the relative widths of each band of color and in the similarities in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands.}
 \label{fig:iris_parallel}
@@
 
-While the Iris dataset does not have any variables that seem to traditionally fit into the model of keys-as it has no time or space variables-an analogous look-up task for the Iris dataset is to filter on a component and then look at the probabiluty distributions of the other variables in relation tp that key.
+While the Iris dataset does not have any variables that seem to traditionally fit into the model of keys-as it has no time or space variables-an analogous look-up task for the Iris dataset is to filter on a component and then look at the probability distributions of the other variables in relation tp that key.
 In effect this treats all variables as potential keys into distributions of the other variables.
-In the parallel coordinates plot (PCP) in figure~\ref{fig:iris_parallel.png}, each line is an observation in the dataset, and each color maps to a species of Iris.
+In the parallel coordinates plot (PCP)\cite{claessen_flexible_2011,_nist/sematech_????,inselberg_plane_1985, wegman_hyperdimensional_1990} in figure~\ref{fig:iris_parallel}, each line is an observation in the dataset, and each color maps to a species of Iris.
 The color coding then allows for this visualization to be used to show how the four quantative variables (sepal length, sepal width, petal length, and petal width) distribute within each variable.
@@
 wherein A is the observation vector $\left< \text{sepal length, sepal width, petal length, petal width} \right>$ and B in turn takes on each species of petal.
-The locations of the color and width of the colored band act as proxies for the distribution of each quantative variable and the overall observation vector.
+The locations of the color and width of the colored band act as proxies for the distribution of each quantitative variable and the overall observation vector.
 
+\subsubsection{Conditional Dependency}
 \begin{figure}
 \includegraphics[width=\textwidth]{intro/iris_observations.png}
-\caption{The diaganol of the scatter matrix shows the distribution of each variable relative to the species type.
+\caption{The diagonal of the scatter matrix shows the distribution of each variable relative to the species type.
 This can be used to illustrate the conditional probability of each variable relative to a type of species.
 The scatter matrix plot shows every set of pairwise relationships in the dataset, color code by Iris species.
-Because the scatter matrix shows the cooccurance of two variables and is encoded with information about a third variable, it works well was a tool to explore a potential conditional dependency amongst the variables.}
+Because the scatter matrix shows the co-occurance of two variables and is encoded with information about a third variable, it works well was a tool to explore a potential conditional dependency amongst the variables.}
 \label{fig:iris_observations}
@@
 
-As exemplified by the scatter matrix in figure~\ref{fig:iris_observations}, there are a number of visualization techniques that can be used to explore conditional dependencies, but these methods are limited to datasets where the observations are discrete.
-There are visualization techniques for understanding the probability distributions of functional observations that could potentially be leverged in conjunction with these methods to develop methods for visualizing conditional dependencies in a functional space.
-This paper provides a sampling of the following visualizations:
-\begin{enumerate}
-\item methods for visualizing conditional dependency for discrete observations
-\item methods for visualizing uncertainity and probabilistic visualizations for functional observations
-\end{enumerate}
+As exemplified by the scatter matrix\cite{elmqvist_rolling_2008,l._wilkinson_high-dimensional_2006} in figure~\ref{fig:iris_observations}, there are a number of visualization techniques that can be used to explore conditional dependencies, but these methods are limited to datasets where the observations are discrete.
+There are visualization techniques for understanding the probability distributions of functional observations that could potentially be leveraged in conjunction with these methods to develop methods for visualizing conditional dependencies in a functional space.
 
-%%\subfile{sections/distdist.tex}
-%%\subfile{sections/distdens.tex}
-%%\subfile{sections/densdist.tex}
-%%\subfile{sections/densdens.tex}
+\begin{figure}
+\input{papermap.tikz}
+\caption{The papers in this survey are organized by the type of observation and then by the type of measurements that makes up the observation.}
+\label{fig:papermap}
+\end{figure}
+
+ This survey organizes the methods for exploring conditional dependency by observation type and data semantics, as shown in figure~\ref{fig:papermap}.
+In the functional space, the focus is on visualizing distributions because that is a necessary component of exploring conditional dependencies.
+
+\section{Discrete Observations}
+\subfile{sections/catset}
+\subfile{sections/radclust}
+\subfile{sections/pairwise}
+\section{Continuous Observations}
+\subfile{sections/quantKquantV}
 
@@
 
-1 Densities and distributions are important
-2 Visualization methods fall off as variables get more
-complex
-3 Can we build on current methods to show conditional
-dependencies and functional variables?
+From climate models to Internet advertisement, a lot of modern datasets are very large and often structurally complex.
+And the questions being asked of these datasets are increasingly probabilistic-what is the distribution of observations, under what events do observations occur, and are there events that have to happen?
+The latter question is the one explored in conditional dependency visualizations.
+As explored in this paper, there are many techniques for exploring these dependencies in the discrete observation space, and there is a lot of work being done to understand distributions in the latter space.
+There is a space for future work that would leverage both to develop conditional dependency visualizations for functional datasets.
 

### 2018-01-31 paper/sections/pairwise.tex
newest version + repo clean up

\documentclass[../main.text]{subfiles}
\begin{document}
\subsection{Quantitative}

Many multivariate visualizations treat the data space as a graph where the vertices's are the variables and the edge corresponds to some quantified relationship between the two vertexes and often these edges are colored by category or form a shape that can be used a proxy for a category. These types of visualizations can often be used to derive typical observations in a given category.

\subsubsection{Parallel Coordinate Plots}
\begin{figure}
  \includegraphics[width=\textwidth]{parallelsets.png}
  \caption{On the left is the categorical view, where users can define
    categories based on other categorical information. For example, in this
    view a working family with children is defined as a household with one or
    more children. On the right is a parallel coordinate graph where the
    dominant category informs the colors, the size of the box is relative to
    the category size, and the histograms show the frequencies of those
    categories relative to the connecting categories (supermarket type and income
    respectively). Because it provides a filtering tool, parallel sets can be used to query whether supermarket and income are conditionally dependent on categorical subsets of family types. This figure is taken from Parallel sets: Interactive exploration and visual analysis of categorical data \cite{kosara_parallel_2006}}
  \label{fig:parallelsets}
\end{figure}

Another important factor in visualizing categorical data is retaining the
inherent hierarchy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
shown in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
of set membership in hierarchal ordered categorical sets \cite{kosara_parallel_2006}. Building on the flexible linked axis version of the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independently but indicates conditional
  dependency through cross tabulations and by grouping the data. This is indicated via
  color and is based on an active subcategory that then becomes the grouping variable for the
  cross tabulations. In Parallel Sets, quantitative data is converted to
  categorical data via binning. Parallel Sets provides a robust filtering interface for users to test out how different subsets of family type affect their choice in supermarket or their income, which in turn can be used to explore whether there is a conditional dependency amongst these variables. Many-to-many parallel coordinate plot is another type of generalized parallel coordinates where all possible pairs are explored \cite{lind_many--many_2009}.

\subsubsection{Partial Dependency Plot}

\begin{figure}
    \begin{subfigure}{\textwidth}
    \centering
      \includegraphics[scale=0.5]{partial_dependence}
      \label{fig:fivechart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.5]{partial_dependency_3}
    \label{fig:threechart}
    \end{subfigure}

  \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots in figure~\subref{fig:fivechart} how California housing prices are partially dependent on each x variable and the joint probability of the variables on housing costs. Figure~\subref{fig:threechart} illustrates the specific relationship between  of the age the house and the number of occupants in the house and the cost.}
  \label{fig:partialdependence}
\end{figure}

One of the most straightforward ways to show dependencies is to plot one
variable against another, but those methods often exclude the effects of the
other variables in the datasets. Regression methods can be used to analyze the
ways in which one variable is partially dependent on many other variables in
the dataset\cite{_elements_2009,scikit-learn}. As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps and surface maps show how California housing prices are marginally dependent on each variable. While marginal dependence does not directly map to conditional dependence, it establishes that there's a probabilistic relationship between housing prices and median income, average number of occupants, and the age of the house. These variables are then explored further in the heatmap which pinpoints that the marginal probabilities between house age and occupancy are for older houses with few people. Because price can not really affect age, it can be inferred that price is either dependent on age or both are correlated with a third variable. In effect partial dependence plots are computing partial conditional dependencies since they show the probabilistic dependence between a target function and a set of target features. The target features are any set of discrete observations, and the target function is a modeling of a continuous observation.

\end{document}

### 2018-01-31 paper/sections/quantKquantV.tex
newest version + repo clean up

--- 2017-08-30 paper/sections/densdens.tex
@@
 \begin{document}
-\section{Densities from Densities}
+\subsection{Variations on Boxplots}
 
-\begin{figure}
-  \includegraphics{partial_dependence}
-  \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots show how California housing prices are partially dependent on each x variable.
-The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
-  \label{fig:partialdependence}
-\end{figure}
-
-One of the most straightforward ways to show dependencies is to plot one
-variable against another, but those methods often exclude the effects of the
-other variables in the datasets. Regression methods can be used to analyze the
-ways in which one variable is partially dependent on many other variables in
-the dataset\cite{_elements_2009}.
-As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.
-
-While bar graphs and histograms can give a sense of the variability in
-data, they are limited to showing the dispersion of the data only in how the measurements are spread out.
+While bar graphs and histograms can give a sense of the variability in data, they are limited to showing the dispersion of the data only in how the measurements are spread out.
 This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham_40_2011}.
-While a simple box plot does not inherently show conditional dependencies, many derivations do.
 
@@
   \caption{Tukey's 1977 example of the box plot from
-    Exploratory Data Analytics\cite{tukey_exploratory_1977} illustrates how the box plot shows the distribution of topology of states and
+    The box plot shows the distribution of topology of states and
     the distribution of volcanoes. The strength of the box and whisker is that the
@@
     that volcanoes tend to span a shorter range of average heights, but that the
-    extremes are much further from the center. Essentially, the distributions act as references for the other displayed distributions.}
+    extremes are much further from the center.
+Essentially, the distributions act as references for the other displayed distributions.
+Figure is from Exploratory Data Analytics\cite{tukey_exploratory_1977} }
   \label{fig:boxplot}
@@
 
+\subsubsection{Boxplots for Curves}
 While box plots were designed to visualize discrete observations, often there
-are no discrete observations and instead the `observation` is a timeseries
-curve (where the connections are as important as the individual points) or
+are no discrete observations and instead the `observation` is a functional densities, drawn from a continuous sample of observations \cite{ramsay_functional_2006, muller_functional_2006}.
+One such example is a timeseries curve (where the connections are as important as the individual points) or
 spatial region. Understanding the distributions of functional observations
-often relies first ordering the functions such that there's some measure of how close these functions are to each other--functional depth\cite{febrero_functional_2007},
+often relies first ordering the functions such that there's some measure of how close these functions are to each other--depth\cite{febrero_functional_2007,lopez-pintado_functional_2007},
 bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
@@
 \begin{figure}
-  \includegraphics[width=.75\linewidth]{contour_weather}
-  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
-  \label{fig:countour}
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[scale=0.8]{spaghetti}
+    \caption{Spaghetti plot of ensemble of 500mb and -15C temperature isocountours.}
+    \label{fig:spaghetti}
+    \end{subfigure}
+
+    \bigskip
+    \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[scale=0.5]{contour_weather}
+    \caption{This countour boxplot is showing the distribution of a temperature field at 500mb and -15C.
+The mean field is in yellow, the median in purple, red as outlier cut offs, and envelops for the minimum and maxmimum bands.}
+    \label{fig:contour}
+    \end{subfigure}
+    \caption{Both figures are from Contour boxplots: A method for characterizing uncertainty in feature sets from simulation ensembles\cite{whitaker_contour_2013}}
 \end{figure}
-Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution.
+
+Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited in the variability they show.
 A very common task is to compare ensembles, each a function of the same timeseries or geographic region.
@@
 hand, which in this paper is temperature visualizations.
-The authors argue that contour box plots are an improvement over spaghetti plots \cite{luo_visualizing_2003} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns.
-Instead in
-Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
-the spaghetti, but removes the visual noise of the lines.
+As illustrated by figure~\ref{fig:spaghetti}, contour boxplots are an improvement over spaghetti plots \cite{luo_visualizing_2003, whitaker_contour_2013} because a lot of the nosiness is filtered out.
+In figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray).
+It retains much of the information of the spaghetti plot in figure~\ref{fig:spaghetti}, but removes the visual noise of the lines.
 
-One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types.
-The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman_eyes_1996} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core.
+\subsubsection{Boxplots for Surfaces}
+\begin{figure}
+  \includegraphics[width=1\linewidth]{surfacebox.png}
+  \caption{The large colored surface in the middle is the median surfac.
+The surfaces on top are minimum and maximum envelopes.
+The graphs at the bottom visualize the distribution of band-depths of surfaces.
+The 3 surfaces on the right have been identified as outliers by the tool.
+This figure is from Surface boxplots \cite{genton_surface_2014}}
+  \label{fig:surface}
+\end{figure}
+Another application of this band depth concept is to surfaces such as collections of images or functional data generated by models.
+In figure~\ref{fig:surface}, Genton et al.
+built a surface boxplot visualization tool with multiple panes\cite{genton_surface_2014}.
+Like a traditional boxplot, they display the median surface from the dataset, minimum and maximum surfaces, and they have a 3D view that gives a sense of how surfaces are distributed.
+The tool also provides histograms of the band depth to give a 2d view on how the surfaces are distributed.
+
 \end{document}

### 2018-01-31 slides/main.tex
newest version + repo clean up

\documentclass[xcolor={dvipsnames}]{beamer}
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{graphicx}

\graphicspath{.figs/}

\usepackage[backend=bibtex, style=numeric-comp, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
\bibliography{../paper/main.bib}

\usetheme{ccnycrest}

\newenvironment{changemargin}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setleng{}th{\parsep}{\parskip}%
}%
\item[]}{\end{list}}

\begin{document}

\title{Visualizing Conditional Dependencies}

\begin{frame}
	\titlepage
    Hannah Aizenman\\
    Advisor: Dr. Michael Grossberg \\
    Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
\end{frame}

\begin{frame}{Predictions are Uncertain}
\begin{figure}
\includegraphics[width=\textwidth]{figs/gettyimages-517387760.jpg}
\end{figure}
\end{frame}

\begin{frame}{So Let's Incorporate Uncertainty}
\begin{figure}
\includegraphics[width=\textwidth]{figs/fivethirtyeight.png}
\end{figure}
\end{frame}

\begin{frame}{But What About Observations?}
\begin{figure}
\includegraphics[width=\textwidth]{figs/ctc_02_img0397.jpg}
\end{figure}
\end{frame}

\begin{frame}{Bertin's 1954 French Workforce\cite{bertin_semiology_2011}}
\begin{itemize}
    \item Geographic region of France
    \item Number of employees (per thousand)
    \item Employment sector: agriculture, industry, everything else
\end{itemize}
\end{frame}

\begin{frame}{What questions are we asking?}
\begin{figure}
\includegraphics[width=\textwidth]{figs/chart_chooser_top.png}
\end{figure}
\end{frame}

\begin{frame}{Incorporate geographic information?}
\begin{figure}
\includegraphics[width=\textwidth]{figs/chart_chooser_bottom.png}
\end{figure}
\end{frame}

\begin{frame}{What if we had more data?}
\includegraphics[width=\textwidth]{figs/growingeconomy.jpg}
\end{frame}

\begin{frame}
{\huge
\begin{align*}
P(v_1, v_2,...v_n \mid k_1, k_2,...k_n)
\end{align*}
}
\begin{block}{Extending Munzner's Keys \& Values \cite{munzner_what:_2014}}
    \begin{description}
        \item[value] unknown distribution/density
        \item[key] known look up measurement 
    \end{description}
\end{block}
\end{frame}

\begin{frame}{Apply this framework to data?}
\begin{block}{What's the likelihood of employment in each sector in Paris?}
    \begin{description}
        \item[value] sector
        \item[keys] region=Lyon
    \end{description}
\end{block}

\pause
\begin{block}{How are transportation workers distributed across France?}
    \begin{description}
        \item[values] region
        \item[key] sector=transportation
    \end{description}
\end{block}
\end{frame}

\begin{frame}{Ordering Visualizations by Keys and Values}
\begin{figure}
\includegraphics[height=.75\textheight]{figs/bertinf.png}
\end{figure}
\end{frame}

\section{Categorical Keys and Categorical Values}
\frame{\sectionpage}

\begin{frame}{Star Wars Super Graphic\cite{leong_star_2017}}
\includegraphics[width=\textwidth]{figs/starwards.jpeg}
\end{frame}

\begin{frame}{Parallel Sets\cite{kosara_parallel_2006}}
\includegraphics[width=\textwidth]{figs/parallelsets.png}
\end{frame}

\begin{frame}{Radar Chart \cite{albo_off_2016}}
\includegraphics[width=\textwidth]{figs/cities_anotated.png}
\end{frame}

\begin{frame}{Flower Chart \cite{albo_off_2016}}
\includegraphics[width=\textwidth]{figs/flower_anotated.png}
\end{frame}

\begin{frame}{Circle Chart\cite{albo_off_2016}}
\includegraphics[width=\textwidth]{figs/circle.png}
\end{frame}

\section{Categorical Keys and Quantitative Values}
\frame{\sectionpage}

\begin{frame}{UpSet\cite{lex_upset:_2014}}
\includegraphics[width=\textwidth]{figs/2014_infovis_upset_teaser.png}
\end{frame}

\begin{frame}{Flexible Axes\cite{claessen_flexible_2011}}
\begin{figure}
\includegraphics[width=\textwidth]{figs/flexaxis.png}
\end{figure}
\begin{description}
        \item[values] ac = acceleration, mg = miles per gallon, w = weight, hp = horse power, yr = model year, cy = cylinders, and dp = displacement
        \item[keys] country of origin: USA (yellow), Europe(blue), Japan (red)
\end{description}
\end{frame}

\section{Quantitative Keys and Categorical Values}
\frame{\sectionpage}

\begin{frame}{Streamgraph\cite{byron_stacked_2008}}
\begin{figure}
\includegraphics[width=\textwidth]{figs/stream.png}
\end{figure}
\begin{description}
        \item[values] musical trends
        \item[key] timeseries
\end{description}

\end{frame}

\section{Quantitative Keys and Quantitative Values}
\frame{\sectionpage}

\begin{frame}{Partial Dependence Plots\cite{_elements_2009}}
\includegraphics[width=\textwidth]{figs/partial_dependence.png}

\end{frame}

\begin{frame}{Cluster \& Calendar\cite{van_wijk_cluster_1999}}
\includegraphics[width=\textwidth]{figs/calendar.png}
\end{frame}

\begin{frame}{Bivariate Colormaps\cite{lucchesi_visualizing_2017}}
\includegraphics[width=\textwidth]{figs/bivariate_anotated.png}
\end{frame}

\section{Functional Values}
\frame{\sectionpage}

\begin{frame}{Contour Boxplots\cite{whitaker_contour_2013}}
\includegraphics[height=.75\textheight]{figs/contour_weather.png}
\end{frame}

\begin{frame}{Surface Boxplots\cite{genton_surface_2014}}
\includegraphics[width=\textwidth]{figs/surfacebox.png}
\end{frame}

\section{Conclusion}
\begin{frame}{Conclusion}
\begin{enumerate}
    \item Densities and distributions are important
    \item Visualization methods fall off as variables get more complex
    \item Can we build on current methods to show conditional dependencies and functional variables? 
\end{enumerate}
\end{frame}

\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\end{document}

### 2018-02-01 paper/sections/section1.tex
flowchart

\documentclass[../main.tex]{subfiles}
\begin{document}

\subsection{Categorical}

\subsubsection{Cross Tabulation Visualization}
\begin{figure}
\includegraphics[width=1\textwidth]{swsg.png}
\caption{Graphic designer Tim Leong visualizes a cross tabulation of the severity of a situation against the language used in each situation to show the relative frequency of really bad situations in Star Wars. The ordering of the phrases by the order in which the movies came out and the slight semantic variations between each phrase can be further mined for potential conditional dependencies between these underlying aspects of the phrase, the phrase itself, and the severity of the situation. Chart is from Star Wars Super Graphic \cite{leong_star_2017}}
\label{fig:starwars}
\end{figure}

Categorical data can be likened to set measurement, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}. In exploring these intersections, the question arises of which categories appear in conjunction with each other and if any one category is the driving force underlying the other categories. For example in frequent item set detection, the task is explicitly to figure out what is the minimum set of items that can be used to predict the occurrence of other items in the basket \cite{leskovec_mining_2014,Srikant:1997:MAR:3001392.3001404} The most common approach is to create a list of frequencies, sometimes of sets of one category or a cross-table\cite{goodman_measures_1991} which is an aggregate function applied to the product of two or more sets. A visual example of a simple cross tabulation can be see in figure~\ref{fig:starwars}, wherein designer Tim Leong looks at the probability of a situation being bad given the structure and content of a given phrase\cite{leong_star_2017}. Each phrase acts as set of size 1, the cross tabulation in this case being the likelihood of the situation being bad. Because the phrases are ordered in temporal order, Leong's visualization suggests that there's a higher likelihood of bad events happening near the middle of the time-line, indicating that severity of situational badness and the type of phrasing used could be conditionally dependent on the location of the story arc.

\begin{figure}
\includegraphics[width=1\textwidth]{hanging_rootgram.png}
\caption{The hanging rootogram moves the histogram bars so that they meet the expected values of the density estimator. This in turn means that deviations between the expected and observed frequencies can be measured more easily as distance between the bottom of the rectangle to a horizontal 0, rather than as a difference between the top of the rectangle and a point on a distribution curve.}
\label{fig:hanging_rootogram}
\end{figure}

 While a table is encouraged for a small number of variables
 \cite{munzner_what_2014}), bar graphs and histograms are typically used to display differences between subsets \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms  include the hanging rootogram shown in figure~\ref{fig:hanging_rootogram} and other distribution oriented plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional relationships can somewhat be visualized using other categorical visualizations such as fourfold displays, mosaic displays, and logit regression models, these plots are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 \subsubsection{Set Visualization}
\begin{figure}
  \includegraphics[width=\textwidth]{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie genres. The left hand side lets the user query the database for a set of movies, listed in the set menu. It then shows how these movies intersect with the other genres of sets. The middle panel shows how many movies are in each intersecting set, and the distribution of select quantitative variables. The right hand panel breaks out the movies in one of these subsets into a scatter plot and a list. This combined visualization tool allows a user to see if quantitative attributes like ratings or times watched can be conditionally dependent on the genre of the movie. This figure comes from UpSet: Visualization of Intersecting Sets \cite{lex_upset_2014}}
   \label{fig:upsetfig}
\end{figure}

The UpSet tool is designed to display a much larger number of set intersections, frequency information about set membership, and the aggregate information of quantitative attributes filtered on sets\cite{lex_upset_2014}. As shown in figure~\ref{fig:upsetfig}, the UpSet tool provides linked set and element views. The set view provides the intersections and their
aggregates, the frequencies of each set and intersection, and the aggregate
statistics of the associated quantitative variables, while the element view shows
selected elements and information about their set membership. The set view yields the distributions of each of the attributes. For example the selected row (in the gray box) are the distributions of each quantitative variable in the comedy set. Every row underneath is the distribution of that quantitative variable restricted to movies in the intersection of the selected sets. For example, the distribution of ratings of movies that are both comedies and dramas. The Upset tool
also facilitates comparison of filtered sets via scatter-plot, which acts as a
visualization of the attributes of the data conditioned on membership in the
filtered set.  The UpSet tool provides a window into conditional dependency because it allows for filtering on sets and then updates the scatter and quantile views; like the scatter matrix, UpSet can isolate a genre or filter on ranges of quantitative variables such that the user can try a variety of approachs to see if there is directionality in the co-occuring events. 
\end{document}

### 2018-02-01 paper/sections/section2.tex
flowchart

\documentclass[../main.tex]{subfiles}
\begin{document}
\subsection{Quantitative as Categorical}

\subsubsection{Radial Visualizations}
\begin{figure}
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{cities.png}
    \caption{The size and shape of each polygon acts as an indicator of Internet usage over time. All events in all cities seem to have expanded over time, indicating a conditional dependence on time.}
    \label{fig:radarchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{flower.png}
    \caption{The size of each flower indicates Internet usage. The grid arrangement in this chart facilitates comparison both across time (horizontally) and city (vertically). As in figure~\ref{fig:radarchart}, Internet usage seems to conditionally dependent on time.}
    \label{fig:flowerchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{circle.png}
    \caption{The circle chart prioritizes intensity of each type of usage, most clearly showing the popularity of shopping and SMS. This visualization facilitates testing a hypothesis of a conditional dependency amongst the Internet usage variables by isolating the similar pairwise color patterns in the presence and absence of a third usage type.}
    \label{fig:circlechart}
    \end{subfigure}
\caption{All three visualizations show Internet usage in 5 cities, from 2003 to 2012. Images are from Off the Radar: Comparative Evaluation of Radial Visualization Solutions for Composite Indicators \cite{albo_off_2016}}
\label{fig:radial}
\end{figure}
Radial visualizations are designed to show composite indicators (CI)\cite{chambers_graphical_1983}; a CI is a type of visualization wherein the multivariate observation is  mapped to a single visual idiom. For example, the whole line across all the Y-axis in a parallel coordinates plot (PCP) is a CI. A radar plot, as shown in figure~\ref{fig:radarchart}, is just a PCP where the two sides of the axis meet. Like a PCP, the radar chart shows pairwise relationships between variables that share nodes, but the radial shape also means it can show pairwise relationships between all variables. In 2010 in figure~\ref{fig:radarchart}, the more rectangular polygon for city 2 relative to the somewhat trapezoid polygon for city 4 indicate that Internet usage is characteristically different for the two cities. The conditional dependency factors in because the radial plot shows changes over time, by city, across multiple components of Internet usage. The flower chart in figure~\ref{fig:flowerchart} is very similar to the radar chart, also highlighting a potential dependency on time due to the use of small multiples and the flowers progressively getting larger. The circle chart in figure~\ref{fig:circlechart} stands out particularly because time progressions are not as readily visible, but instead this chart, especially with a finer grained colormap, best highlights the specific usage variables. This chart most clearly shows that shopping and sms seems to happen in tandem, and that clear color delineation also makes it easier for a user to scan the other variables to see how frequently they occur given the relationship between shopping and SMS.

\subsubsection{Cluster Based Visualizations}
\begin{figure}
\includegraphics[width=1\textwidth]{calendar.png}
\caption{Figure from \cite{van_wijk_cluster_1999}. Van Wijk and Van Selow visualized the number of employees present at a research facility. This visualization shows that daily attendance seems to be more strongly dependent on day of week than time of day or day of the year. Figure is from Cluster and Calendar based Visualization of Time Series Data \cite{van_wijk_cluster_1999}}
\label{fig:calendar}
\end{figure}

Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~\ref{fig:calendar} shows an analysis of employee attendance at a research facility in the Netherlands\cite{van_wijk_cluster_1999}. Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day. They then used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.  In this study, time is separated into three classes of events: time of day, day of week, and date. The researchers found that both the shape and magnitude of the attendance curve changed relative to the day of week, indicating a conditionally dependent relationship between attendance and day of week. 
\end{document}

### 2018-02-07 paper/sections/section1.tex
submitted paper

@@
 \subsection{Categorical}
+\label{sec:categorical}
+Categorical data, including vectors of categorical observations, can be represented as sets; researchers then query set cardinality, membership, and intersections \cite{agresti_categorical_2011,schneider_set-theoretic_2012}.
+In exploring these intersections, the question arises of which categories appear in conjunction with each other and if any one category is the driving force underlying the frequency of other categories.
+For example, in frequent item set detection, the task is to figure out the minimum set of items that can be used to predict the occurrence of other items in the set \cite{leskovec_mining_2014,Srikant:1997:MAR:3001392.3001404}.
+By representing categorical variables as sets, we can exploit methods of set visualization to explore dependencies amongst categorical variables.
 
-\subsubsection{Cross Tabulation Visualization}
-\begin{figure}
+\subsubsection{Cross Table}
+\label{sec:crosstab}
+\begin{figure}[H]
 \includegraphics[width=1\textwidth]{swsg.png}
-\caption{Graphic designer Tim Leong visualizes a cross tabulation of the severity of a situation against the language used in each situation to show the relative frequency of really bad situations in Star Wars.
-The ordering of the phrases by the order in which the movies came out and the slight semantic variations between each phrase can be further mined for potential conditional dependencies between these underlying aspects of the phrase, the phrase itself, and the severity of the situation.
-Chart is from Star Wars Super Graphic \cite{leong_star_2017}}
+\caption{Graphic designer Tim Leong visualizes a cross table of a phrase against the severity of the situation in which it was said.
+Because the situations are ordered chronologically, this illustration shows that the most severe situations occur in the middle of the Star Wars' timeline.
+Chart is from \textit{Star Wars Super Graphic} \cite{leong_star_2017}}
 \label{fig:starwars}
@@
 
-Categorical data can be likened to set measurement, where each category is a set
-and researchers often want to know how many members each set has and what
-intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}.
-In exploring these intersections, the question arises of which categories appear in conjunction with each other and if any one category is the driving force underlying the other categories.
-For example in frequent item set detection, the task is explicitly to figure out what is the minimum set of items that can be used to predict the occurrence of other items in the basket \cite{leskovec_mining_2014,Srikant:1997:MAR:3001392.3001404} The most common approach is to create a list of frequencies, sometimes of sets of one category or a cross-table\cite{goodman_measures_1991} which is an aggregate function applied to the product of two or more sets.
-A visual example of a simple cross tabulation can be see in figure~\ref{fig:starwars}, wherein designer Tim Leong looks at the probability of a situation being bad given the structure and content of a given phrase\cite{leong_star_2017}.
-Each phrase acts as set of size 1, the cross tabulation in this case being the likelihood of the situation being bad.
-Because the phrases are ordered in temporal order, Leong's visualization suggests that there's a higher likelihood of bad events happening near the middle of the time-line, indicating that severity of situational badness and the type of phrasing used could be conditionally dependent on the location of the story arc.
+One of the most common approaches for understanding relationships between categorical variables is to aggregate a function applied to the product of two or more sets; this computation is called a cross table \cite{goodman_measures_1991}.
+Figure~\ref{fig:starwars} is a visual example of a cross table where the sets are a phrase spoken by a character in a Star Wars film and the severity of the situation in which the phrase is spoken \cite{leong_star_2017}.
+Designer Tim Leong structures the visualization such that the phrases are chronologically ordered with respect to the Star Wars' timeline and he lines up the phrases such that variations are readily apparent.
+He then encodes the probability of danger as the height of the yellow bar such that the more dangerous, the more it encroaches on the \textit{Really Bad Situation} side of the visualization.
+Each phrase is a set with one element (the phrase), and the cross tabulation is the likelihood that the speaker is in a dangerous situation.
+Because the phrases are ordered in temporal order, Leong's visualization suggests that there is a higher likelihood of bad events happening near the middle of the timeline, indicating that severity of situational badness and the type of phrasing used could be conditionally dependent on the location of the event- in the story arc.
+Leong effectively uses a cross table to visualize a small number of groups (phrases) and a very small number of variables (phrases, situations).
+Cross-tabulations shine when there are a few variables and a small number of groups, but get very unwieldy as the size of the set product grows.
 
-\begin{figure}
+\begin{figure}[H]
 \includegraphics[width=1\textwidth]{hanging_rootgram.png}
-\caption{The hanging rootogram moves the histogram bars so that they meet the expected values of the density estimator.
-This in turn means that deviations between the expected and observed frequencies can be measured more easily as distance between the bottom of the rectangle to a horizontal 0, rather than as a difference between the top of the rectangle and a point on a distribution curve.}
+\caption{The hanging rootogram shifts the histogram bars up so that they intersect with the expected values of the density estimator.
+This means that deviations between the expected and observed frequencies can be measured as the distance between the bottom of the rectangle and horizontal line at x=0.}
 \label{fig:hanging_rootogram}
 \end{figure}
+ While a simple table is encouraged for a small number of variables or categories \cite{munzner_visualization_2014}), bar graphs and histograms \cite{ioannidis_history_2003-1, friendly_brief_2006} are more appropriate when the variable has a large number of unique categories or the dataset has a large number of variables.
+Variations on histograms  include the hanging rootogram shown in figure~\ref{fig:hanging_rootogram} and other distribution oriented plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}.
+While conditional relationships can be visualized using other categorical visualizations such as fourfold displays, mosaic displays, and logit regression models, these plots are somewhat limited to 3 categorical variables \cite{friendly_visualizing_2000}.
+For datasets with more complex categorical relationships, we can look to set oriented visualization tools
 
- While a table is encouraged for a small number of variables
- \cite{munzner_what_2014}), bar graphs and histograms are typically used to display differences between subsets \cite{ioannidis_history_2003-1, friendly_brief_2006}.
-Variations on histograms  include the hanging rootogram shown in figure~\ref{fig:hanging_rootogram} and other distribution oriented plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}.
-While conditional relationships can somewhat be visualized using other categorical visualizations such as fourfold displays, mosaic displays, and logit regression models, these plots are somewhat limited to 3 sets of categories
- \cite{friendly_visualizing_2000}.
-
- \subsubsection{Set Visualization}
-\begin{figure}
+ \subsubsection{Set Intersection}
+ \label{sec:setintersection}
+\begin{figure}[H]
   \includegraphics[width=\textwidth]{upsetfig1}
-   \caption{UpSet dashboard displaying the relationships amongst movie genres.
-The left hand side lets the user query the database for a set of movies, listed in the set menu.
-It then shows how these movies intersect with the other genres of sets.
-The middle panel shows how many movies are in each intersecting set, and the distribution of select quantitative variables.
-The right hand panel breaks out the movies in one of these subsets into a scatter plot and a list.
-This combined visualization tool allows a user to see if quantitative attributes like ratings or times watched can be conditionally dependent on the genre of the movie.
-This figure comes from UpSet: Visualization of Intersecting Sets \cite{lex_upset_2014}}
+   \caption{UpSet dashboard displaying the relationships between a movie's genre and its release date, ratings, and popularity.
+The left panel of the dashboard lets the user query the database for all movies in a selected genre.
+It then shows which other genres these movies belong to.
+The middle panel shows the distribution of release dates, ratings, and popularity of the movies in each intersecting set of genres.
+The right panel lists the individual movies in one of these intersecting sets and plots their quantitative attributes as scatter plot.
+This combined visualization tool allows a user to see if quantitative attributes like ratings or popularity are conditionally dependent on the genre.
+This figure comes from \textit{UpSet: Visualization of Intersecting Sets} \cite{lex_upset_2014}}
    \label{fig:upsetfig}
@@
 
-The UpSet tool is designed to display a much larger number of set intersections, frequency information about set membership, and the aggregate information of quantitative attributes filtered on sets\cite{lex_upset_2014}.
-As shown in figure~\ref{fig:upsetfig}, the UpSet tool provides linked set and element views.
-The set view provides the intersections and their
-aggregates, the frequencies of each set and intersection, and the aggregate
-statistics of the associated quantitative variables, while the element view shows
-selected elements and information about their set membership.
-The set view yields the distributions of each of the attributes.
-For example the selected row (in the gray box) are the distributions of each quantitative variable in the comedy set.
-Every row underneath is the distribution of that quantitative variable restricted to movies in the intersection of the selected sets.
-For example, the distribution of ratings of movies that are both comedies and dramas.
-The Upset tool
-also facilitates comparison of filtered sets via scatter-plot, which acts as a
-visualization of the attributes of the data conditioned on membership in the
-filtered set.
-The UpSet tool provides a window into conditional dependency because it allows for filtering on sets and then updates the scatter and quantile views; like the scatter matrix, UpSet can isolate a genre or filter on ranges of quantitative variables such that the user can try a variety of approachs to see if there is directionality in the co-occuring events.
+The UpSet tool is designed to display a large number of set intersections, frequency information about set membership, and information about the distribution of quantitative attributes of elements in a set \cite{lex_upset_2014}.
+As shown in figure~\ref{fig:upsetfig}, the UpSet tool provides linked set and element views so that users can get a top level view of the characteristics of a genre of movies and also information about the movies in that genre.
+The set view on the left panel of the dashboard lets users select a genre and then displays information about how many movies are in that genre and what other genres those movies belong to.
+The middle panel shows the distribution of the release dates, average ratings, and popularity of the movies in each of the sets shown on the right side.
+The right panel element view displays the movies in the selected genre and the scatter plot shows each movie's ratings and popularity.
+The UpSet tool provides a window into conditional dependency because a user can see if attributes of the movies in a genre, such as average movie rating and release date, are conditionally dependent on genre.
+Users can select a genre (e.g.
+comedy) and look at how the attributes of the movies in that set distribute and then compare those results against the distributions of other genres.
+While the UpSet tool can handle data that is too complex for cross tables (\ref{sec:crosstab}), it can not preserve structure and is limited to showing intersections of sets of one variable type (e.g.
+only genres) and a low number of quantitative variables.
+
 \end{document}

### 2018-02-07 paper/sections/section2.tex
submitted paper

@@
 
-\subsubsection{Radial Visualizations}
-\begin{figure}
+Unlike in section~\ref{sec:categorical}, some datasets have a very large number of observations or the observations are a vector of quantitative measurements that work in tandem.
+One approach for handling these types of datasets is to develop categorical representations of these quantitative variables.
+
+\subsubsection{Radial Charts}
+\label{sec:radial}
+One method of transforming quantitative vectors into categorical representations is to visualize the vector using a radial graph.
+Radial visualizations are designed to show composite indicators (CI) \cite{chambers_graphical_1983}; a CI is a type of visualization where the multivariate observation is mapped to a single visual idiom.
+For example, the line representing the observation vector in a parallel coordinates plot is a CI.
+\begin{figure}[H]
     \begin{subfigure}{\textwidth}
@@
     \includegraphics[scale=0.2]{cities.png}
-    \caption{The size and shape of each polygon acts as an indicator of Internet usage over time.
-All events in all cities seem to have expanded over time, indicating a conditional dependence on time.}
+    \caption{The size and shape of each polygon acts as an indicator of Internet usage over time.}
     \label{fig:radarchart}
@@
     \centering
-    \includegraphics[scale=0.2]{flower.png}
-    \caption{The size of each flower indicates Internet usage.
-The grid arrangement in this chart facilitates comparison both across time (horizontally) and city (vertically).
-As in figure~\ref{fig:radarchart}, Internet usage seems to conditionally dependent on time.}
+    \includegraphics[scale=0.15]{flower.png}
+    \caption{The size of each flower indicates Internet usage. The grid arrangement in this chart facilitates comparison both across time (horizontally) and city (vertically).}
     \label{fig:flowerchart}
@@
     \centering
-    \includegraphics[scale=0.2]{circle.png}
+    \includegraphics[width=1\linewidth]{circle.png}
     \caption{The circle chart prioritizes intensity of each type of usage, most clearly showing the popularity of shopping and SMS.
-This visualization facilitates testing a hypothesis of a conditional dependency amongst the Internet usage variables by isolating the similar pairwise color patterns in the presence and absence of a third usage type.}
+This visualization facilitates testing a hypothesis of a conditional dependency amongst the Internet usage variables by isolating similar pairwise color patterns in the presence and absence of a third usage type.}
     \label{fig:circlechart}
     \end{subfigure}
-\caption{All three visualizations show Internet usage in 5 cities, from 2003 to 2012.
-Images are from Off the Radar: Comparative Evaluation of Radial Visualization Solutions for Composite Indicators \cite{albo_off_2016}}
+\caption{Internet usage in 5 cities from 2003 to 2012.
+Since all types of Internet usage in all the cities seem to have expanded over time, this suggests a conditional dependence of different types of usage on time.
+Images are from \textit{Off the Radar: Comparative Evaluation of Radial Visualization Solutions for Composite Indicators} \cite{albo_off_2016}}
 \label{fig:radial}
 \end{figure}
-Radial visualizations are designed to show composite indicators (CI)\cite{chambers_graphical_1983}; a CI is a type of visualization wherein the multivariate observation is  mapped to a single visual idiom.
-For example, the whole line across all the Y-axis in a parallel coordinates plot (PCP) is a CI.
-A radar plot, as shown in figure~\ref{fig:radarchart}, is just a PCP where the two sides of the axis meet.
-Like a PCP, the radar chart shows pairwise relationships between variables that share nodes, but the radial shape also means it can show pairwise relationships between all variables.
-In 2010 in figure~\ref{fig:radarchart}, the more rectangular polygon for city 2 relative to the somewhat trapezoid polygon for city 4 indicate that Internet usage is characteristically different for the two cities.
-The conditional dependency factors in because the radial plot shows changes over time, by city, across multiple components of Internet usage.
-The flower chart in figure~\ref{fig:flowerchart} is very similar to the radar chart, also highlighting a potential dependency on time due to the use of small multiples and the flowers progressively getting larger.
-The circle chart in figure~\ref{fig:circlechart} stands out particularly because time progressions are not as readily visible, but instead this chart, especially with a finer grained colormap, best highlights the specific usage variables.
-This chart most clearly shows that shopping and sms seems to happen in tandem, and that clear color delineation also makes it easier for a user to scan the other variables to see how frequently they occur given the relationship between shopping and SMS.
+A radar plot, as shown in figure~\ref{fig:radarchart}, is a parallel coordinate plot (PCP) where the ends of the line representing the observation are joined.
+Like a PCP, the radar chart shows pairwise relationships between variables that share nodes, but the radial shape also means that the polygon it forms encodes information about the relationship between all the variables.
+In figure~\ref{fig:radarchart} in 2010, the more rectangular shape of the city 2 Polygon relative to city 4's trapezoidal polygon signifies a characteristic difference between Internet usage in the two cities.
+Since figure~\ref{fig:radarchart} is arranged in small multiples of radar charts showing changes over time, by city, and across multiple types of usage, the figure can be used to examine whether this is a conditional dependency on time or city or usage type.
+Similarly, the flower chart in figure~\ref{fig:flowerchart} also highlights a potential dependency on time due to the flowers progressively getting larger across time.
+The circle chart in figure~\ref{fig:circlechart} differs from the other charts in figure~\ref{fig:radial} because time progressions are not as readily visible.
+Instead this chart (especially if the colormap had more steps) most clearly visualizes the frequency of the different types of Internet usage.
+Figure~\ref{fig:circlechart} shows that shopping and sms seem to happen in tandem, and that clear color delineation also makes it easier for a user to scan the other variables to see how frequently they occur given the relationship between shopping and SMS.
+The radar, flower, and circle chart visualize Internet usage as circles to varying degrees of success, but radar charts are limited to how much information can reasonably fit on one circle or a small number of circles.
 
-\subsubsection{Cluster Based Visualizations}
-\begin{figure}
+\subsubsection{Clustering}
+
+\label{sec:cluster}
+\begin{figure}[H]
 \includegraphics[width=1\textwidth]{calendar.png}
-\caption{Figure from \cite{van_wijk_cluster_1999}.
-Van Wijk and Van Selow visualized the number of employees present at a research facility.
-This visualization shows that daily attendance seems to be more strongly dependent on day of week than time of day or day of the year.
-Figure is from Cluster and Calendar based Visualization of Time Series Data \cite{van_wijk_cluster_1999}}
+\caption{Van Wijk and Van Selow visualized the number of employees present at a research facility.
+This visualization shows that daily attendance seems to be more strongly dependent on the day of the week than on the time of day or the day of the year.
+Figure is from \textit{Cluster and Calendar based Visualization of Time Series Data} \cite{van_wijk_cluster_1999}}
 \label{fig:calendar}
@@
 
-Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset.
+If the data is too big for techniques that try to visualize the observations directly, one approach is to instead visualize representational fingerprints of common or outlying events in the dataset.
 This is frequently done through the use of a clustering algorithm.
-Figure~\ref{fig:calendar} shows an analysis of employee attendance at a research facility in the Netherlands\cite{van_wijk_cluster_1999}.
-Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day.
-They then used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.
-In this study, time is separated into three classes of events: time of day, day of week, and date.
-The researchers found that both the shape and magnitude of the attendance curve changed relative to the day of week, indicating a conditionally dependent relationship between attendance and day of week.
+Figure~\ref{fig:calendar} shows an analysis of employee attendance at a research facility in the Netherlands \cite{van_wijk_cluster_1999}.
+The individual observations were recorded every 6 hours for a year, so the resulting dataset was a $(\text{day} \times \text{hours})$ matrix of the number of employees at work for that day and time.
+To reduce the dimensionality of this dataset to something easier to visualize, they used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} where the days were observations and the hours were the feature vectors.
+The results of this clustering were used to create daily attendance curves representative of the days in each cluster; these representative curves are shown in the timeseries plot in figure~\ref{fig:calendar}.
+They visualized the cluster membership of the days by coloring them the same as their representative cluster, and preserved the temporal structure of the data by presenting this cluster membership as a calendar.
+The researchers found that both the shape and magnitude of the attendance curve changed relative to the day of the week, indicating a conditionally dependent relationship between attendance and the day of the week.
+While the visualization Van Wijk and Van Selow built is very specific to the question they were tackling, clustering techniques and the use of visual idioms related to the structure of the data (such as a calendar to present discretized time information) can be readily generalized.
+The main limitations to applying their approach to other datasets are that clustering inherently reduces dimensions and clustering techniques are limited when the dataset is a mix of categorical and quantitative variables.
 \end{document}

### 2018-02-07 paper/sections/section3.tex
submitted paper

\documentclass[../main.text]{subfiles}
\begin{document}
\subsection{Quantitative}

While clustering (~\ref{sec:cluster}) techniques are very effective for understanding large datasets, often a researcher explicitly wants to explore the relationship between individual observations. Visualizations that emphasize observational relationships often either group like observations together or explore pairwise interactions.

\subsubsection{Stacked Plot}
\label{sec:stacked}
\begin{figure}[H]
\includegraphics[width=\textwidth]{stream.png}
\caption{Section of graphic designer Lee Byron's \textit{Listening History} \cite{fry_what_????,l._byron_stacked_2008} project based on last.fm data. The spikes in listening to new artists following periods of decreased listening suggest a possible dependency between previous listening behavior and openness to trying new music. The figure is from \textit{Stacked Graphs-Geometry \& Aesthetics} \cite{l._byron_stacked_2008}
}
\label{fig:streamgraph}
\end{figure}

For timeseries, or other continuous functional data, a streamgraph shows how categories of data change across the time \textit{component} of the visualization. A stream graph, which is a type of stacked graph, is created by either horizontally or vertically stacking continuous data \cite{Heer:2010:TTV:1743546.1743567, l._byron_stacked_2008}. In \textit{Listening History} \cite{fry_what_????,l._byron_stacked_2008}, Lee Byron encodes each artist as a unique color, but groups the colors such that more established authors are represented by cooler colors (blues) and newer artists by warmer colors (yellows). In figure~\ref{fig:streamgraph}, the more established artists maintain a stable listener base over time, while newer artists wax and wane in popularity. Figure~\ref{fig:streamgraph} shows a seasonal pattern of fairly consistent listening over the year and a severe acute drop in listening in December, after which follows a spike in popularity for the then new artist \textit{DJ Shadow}. A drop in mid-July leads to a spike for the new bands \textit{Pat Metherly Group} and \textit{Prefuse73}. This suggests a conditional dependency between low periods of listening and choosing to listen to new acts. A stacked graph can be used to infer conditional dependency when encoded in such a way that it can provide information on two or more variables, but the graph is limited to about 3 or 4 variables.

\subsubsection{Parallel Coordinates}
\label{sec:pcp}
\begin{figure}{H}
  \includegraphics[width=\textwidth]{parallelsets.png}
  \caption{On the left is the categorical view where users can select a set of interest, for example a working family with one or more children. At the top of the parallel sets graphs is a set of supermarket categories. The width of the lines flowing out of each category is representational of the number of elements that are in both that category and the category the line is going into. The left most line is representational of the number of households that shop at a low-end supermarket and that have a family type of unemployed. The top histograms show how each family type is distributed across all market types, and the bottom histograms show how each family type is distribute across income levels. Like the lines going into the family type view, the width of the lines represents the number of households of that family type in that income level and are color coded by supermarket type. Because it provides a filtering tool, parallel sets can be used to query whether supermarket and income are conditionally dependent on categorical subsets of family types. This figure is taken from \textit{Parallel sets: Interactive exploration and visual analysis of categorical data} \cite{kosara_parallel_2006}}
  \label{fig:parallelsets}
\end{figure}

When there are no key semantics to preserve, one method for showing conditional dependence is to combine the pairwise visualizations of a parallel coordinates plot with histograms and categorical filtering. The Parallel Sets tool shown in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
of set membership in hierarchal ordered categorical sets \cite{kosara_parallel_2006}. Similarly to the flexible linked axis version of the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets explicitly groups observations by category and shows how the set of observations in those categories flows into another set of categories. Parallel sets treats each categorical set independently, but indicates conditional dependency through cross tables and by grouping the data. This is indicated via color and is based on an active subcategory that then becomes the grouping variable for the cross tables. In Parallel Sets, quantitative data is converted to categorical data via binning. Parallel Sets provides a filtering interface for users to test out how different subsets of family type affect family income or choice of supermarket, which in turn can be used to explore whether there is a conditional dependency amongst these variables. The many-to-many parallel coordinate plot is another type of generalized parallel coordinates where all possible pairs are explored \cite{lind_many--many_2009}. Parallel sets is particularly good for visualizing datasets where the conditional dependency is on categorical datasets, but like many of the other techniques is limited to a relatively small number of variables.

\subsubsection{Partial Dependency}
\label{sec:pdp}

\begin{figure}[H]
\includegraphics[width=1\textwidth]{partial_dependence}
\caption{Taken from \textit{Scikit Learn} \cite{_partial_????}, the line plots show that California housing prices are partially dependent on occupancy and median income. The heatmap shows that housing prices rise for old houses, but only if the house does not have many occupants.}
\label{fig:partialdependence}
\end{figure}

While Streamgraphs (\ref{sec:stacked}) require a continuous variable and Parallel Sets (\ref{sec:pcp} require categorical variables, the partial dependency plot is designed for quantitative observations. One way to show dependencies is to plot one variable against another, but those methods often do not measure the joint effect of the variables in the dataset. One way to analyze this interaction in the California Housing dataset is to use a regression method \cite{_elements_2009,scikit-learn} to measure the partial dependence. As shown in figure~\ref{fig:partialdependence}, the 4 one-way partial dependence line plots and the two-way partial dependence heatmap show how California housing prices are marginally dependent on median income, average occupancy and slightly on the age of the house. While marginal dependence is not conditional dependence, it establishes that there is a probabilistic relationship between housing prices and median income, average number of occupants, and the age of the house. These variables are then explored further in the heatmap, which pinpoints that the marginal probabilities between house age and occupancy are for older houses with few people. Because a house can not get younger, it can be inferred that price is partially dependent on age. In effect, partial dependence plots are computing partial conditional dependencies since they show the probabilistic dependence between a target function and a set of target features. The target features are any set of discrete observations, and the target function is a modeling of a continuous observation. While flexible in terms of what data it can be used with, partial dependence plots are limited in scope to datasets and questions that can be framed as regression problems. 
\end{document}

### 2018-02-07 paper/sections/section4.tex
submitted paper

\documentclass[../main.tex]{subfiles}

\begin{document}
\subsection{Functional Boxplots}
\label{sec:boxplots}

Only some of the previously described visualization methods work for continuous data and none of them can display multiple surface observations (e.g. multiple maps). One of the reasons it is so difficult to visualize conditional dependency in structurally complex datasets is due to the difficulty of even visualizing the probability distributions of these observations.

\begin{figure}[H]
  \center
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot shows the distribution of state and volcano elevations. The strength of the box and whisker is that it clearly shows that volcanoes tend to span a shorter range of average heights, but that the extremes are much further from the center. Figure is from \textit{Exploratory Data Analytics} \cite{tukey_exploratory_1977} }
  \label{fig:boxplot}
\end{figure}

The box plot is the standard method for visualizing the distribution of a quantitative variable and the box plot has evolved to display many more characteristics of the distribution \cite{wickham_40_2011}. Figure~\ref{fig:boxplot} illustrates that the distribution of volcano elevations is more densely packed at the center and has wider but leaner tails then the distribution of state elevations. While figure~\ref{fig:boxplot} shows the median, range between upper and lower interquartile, and lower and upper extreme \cite{tukey_exploratory_1977}, many authors have exploited the structure of the boxplot to encode more information. Authors used different quantile levels \cite{hyndman_sample_1996},xs or measures of outliers \cite{frigge_implementations_1989, schwertman_identifying_2007}, or otherwise incorporated skewness, kurtosis, and other descriptive distributional statistics \cite{kim_more_2004, marmolejo-ramos_shifting_2015}.

While box plots were designed to visualize discrete observations, there are methods for expanding them to functional observations. Understanding the distributions of functional observations requires ordering the functions such that there is some measure of how close these functions are to each other; traditional distance metrics discretize the observation into points sampled from the curve and so lose the overall structure of the curve. Instead functional observations are often ordered using band depth, which is the sum of the probabilities that the observations in any given functional observation fall within the max-min envelope defined by any two other functional observations \cite{Febrero_functional_2007,lopez-pintado_functional_2007}. Band depth and derivatives such as bivariate score depth \cite{rob_j._hyndman_rainbow_2010} and bivariate kernel density estimation \cite{scott_multivariate_1992} yield functional analogs to the median and other percentile metrics. For small values, the ordered curves can then be colored based on the ordering metric \cite{rob_j._hyndman_rainbow_2010}.

\subsubsection{1D Curve}
\label{sec:curve}

\begin{figure}[H]
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.4]{spaghetti}
    \caption{Spaghetti plot of ensemble of 500mb -15C temperature isocountours.}
    \label{fig:spaghetti}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.5]{contour_weather}
    \caption{Contour boxplot showing the distribution of a 500mb -15C temperature field. The mean field is in yellow, the median in purple, and the dashed red lines are the outlier cut off bands.}
    \label{fig:contour}
    \end{subfigure}

    \caption{Figures are from \textit{Contour boxplots: A method for characterizing uncertainty in feature sets from simulation ensembles} \cite{whitaker_contour_2013}}
\end{figure}

Climate and weather models are functional, and increasingly they are ensembles of models with slightly different parameters; to analyze the skill of these models, it is important to visualize the spread in the predictions of the ensemble members. Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} can all be used to show the distribution of ensemble predictions, but contour box plots \cite{whitaker_contour_2013} are designed to highlight the variability in the predictions. In figure~\ref{fig:contour}, the ensemble members are ordered by band depth such that the median band is the ensemble member with about 50\% of its members within all envelopes formed by the other bands. As illustrated by figure~\ref{fig:contour}, the contour boxplots are an improvement over the spaghetti  plots \cite{luo_visualizing_2003, whitaker_contour_2013} in figure~\ref{fig:spaghetti} because the authors replace most of the bands with an outlying envelope in light purple-gray and a more central envelope in darker purple-gray. The contour boxplot effectively displays variability in isocountour data, but is also inherently limited to curves.

\subsubsection{2D Surface}
\label{sec:surface}
\begin{figure}[H]
  \includegraphics[width=1\linewidth]{surfacebox.png}
  \caption{The large colored heatmap in the middle of the dashboard is the median surface. The surfaces at the top of the dashboard are minimum and maximum envelopes. The graphs at the bottom visualize the distribution of the band-depths of the surfaces. The 3 surfaces on the right are outliers. This figure is from \textit{Surface boxplots} \cite{genton_surface_2014}}
  \label{fig:surface}
\end{figure}

Functional climate and weather models are often regional or global, and it is just as critical to understand the spatial distribution of predictions as the temporal ones. In figure~\ref{fig:surface}, Genton et al. built a surface boxplot visualization tool with multiple panes \cite{genton_surface_2014}. Like a traditional boxplot, the surface boxplot tools displays the median surface and minimum and maximum surfaces. The tool also provides a 3D view of how the surfaces are distributed. Like many other functional boxplot methods, the surfaces are ordered using the band depth metric; therefore the tool also visualizes the distribution of the surface band-depths. As with the functional and contour boxplots (\ref{sec:curve}), the surface boxplot is overly specific to data that is a series of surfaces. 
\end{document}

### 2018-02-07 paper/sections/section5.tex
submitted paper

\documentclass[../main.tex]{subfiles}

\begin{document}

\subsection{Uncertainty Estimation}
\label{sec:uncertainty}

\begin{figure}[H]
\includegraphics[width=1\textwidth]{uncertainty.png}
\caption{A model based on previous observations (light blue) and a model based on bias-corrected forecasts (pink) underestimated the likelihood of the temperature being around 294.15 at a grid point in the equatorial Atlantic ocean. The standard deviation of the model curves can be used as a proxy for how certain the models are about their predictions; the narrower observation model was more certain than the bias-corrected forecast model.\cite{aizenman_ensemble_2016}}
\label{fig:uncertainty}
\end{figure}

Another way to explore the probability distribution of semantically complex datasets is to leverage the field of uncertainty visualizations. As data analysis becomes increasingly probabilistic, there is a need to visualize the certainty of a model's prediction. As seen in figure~\ref{fig:uncertainty}, a model based on previous observations (light blue) and one based on a bias-corrected forecast (pink)) expected the temperature to be lower then it actually was\cite{aizenman_ensemble_2016}. The narrower standard deviation of the model based on observations indicates that the observation model was more confident of its prediction than the bias-corrected forecast model. A visualization of this uncertainty would try to show the deviation or a comparable measure of uncertainty. The measuring and encoding of uncertainty suggests methods for represent the structure of functional probabilistic distributions in a discrete less structurally complex manner.

\subsubsection{Bivariate Choropleth}
\begin{figure}[H]
\includegraphics[width=1\textwidth]{bivariate.png}
\caption{American Community Survey poverty rates are encoded using shades of blue and the margin of error is in shades of yellow. Regions that are dark gray have a high poverty rate, but the measurement has a high margin of error. Figure is from \textit{Visualizing uncertainty in areal data} \cite{lucchesi_visualizing_2017}}
\label{fig:bivariate}
\end{figure}
One of the difficulties of visualizing data on maps is that the map takes up the x and y axis, so it is no longer available for other variables. In figure~\ref{fig:bivariate}, bivariate colormaps \cite{trumbo_theory_1981,carstensen_hypothesis_1986} are used to encode the poverty rate and margin of error in each American Community Survey tract. The advantage of using the bivariate colormap is that the uncertainty of the data is encoded in the same symbol as the data itself. This technique easily translates to representing a probability distribution of values at a location as a mean and standard deviation. The bivariate colormap can also natively be used to show conditional dependencies when the events are dependent on location. In figure~\ref{fig:bivariate}, the high poverty and low response rates across the southeastern cotton belt are reminiscent of many other socioeconomic patterns common to that region. The biggest limit of bivariate colormaps is that they can be difficult to read \cite{carstensen_hypothesis_1986}.

A full survey of uncertainty visualization techniques is outside the scope of this paper, but many such methods are listed in \textit{Visualization and Visual Analysis of Multifaceted Scientific Data: A Survey} \cite{kehrer_visualization_2013} and \textit{Visualizing High-Dimensional Data: Advances in the Past Decade} \cite{liu_visualizing_2017}.

\end{document}

### 2018-07-16 paper/main.tex
backup

--- 2017-09-27 paper/main.tex
@@
 \documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
+
+\usepackage[utf8]{inputenc}
+\usepackage{tikz}
+\usetikzlibrary{mindmap}
+\usetikzlibrary{arrows.meta, chains, positioning, shapes.geometric}
 
@@
 \usepackage[backend=bibtex, style=numeric-comp]{biblatex}
-\bibliography{main}
+\bibliography{glasslab_viz}
 
 \usepackage{array,multirow,graphicx}
+\usepackage{subcaption}
 \usepackage{subfiles}
@@
 \usepackage{amsmath}
+\usepackage{float}
 
@@
 \author{Hannah Aizenman}
-\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
+\committee{Dr. Michael Grossberg (Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
 \submitted{}
-\abstract{Many datasets are too large or complex to discern patterns simply through visualizing the observations in them, so researchers instead compute the distributions of select variables and explore under what conditions they occur.
-While there are many visualization techniques that preserve variable type, dimensionality, and mutlivariate relationships, the complexity increases when also trying to illustrate distributions and conditional relationships.
-This survey presents a sampling of techniques for visualizing distributions and densities conditioned on variables that are categorical or quantitative, discrete or continuous, and potentially have or infer dimensional dependencies.
-}
+\abstract{Often the question in data exploration is not if a pattern occurs, but why.
+The search for the why is often the search for a conditionally dependent relationship in the dataset.
+While there are visualization methods that can be used to uncover a potential causal link between variables in the dataset, many of these techniques only work when the causal variables are categorical or can only be applied between two variables at a time or discard the complex, often spatiotemporal, structure of the dataset.
+Instead structure preserving methods for visualizing the distribution of observations provide a way of understanding some of the probabilistic relationships within and between the observations in these complex datasets.
+Using the type and structure of the variables in a dataset as a guide, this survey presents methods for visualizing conditional dependencies for multivariate datasets with discrete observations and visualizing probability distributions for datasets where observations are drawn from a continuous (functional) space.
+Because there is limited literature on explicit ways to visualize conditional dependency, we include techniques that suggest or can be adapted to show these dependencies.}
+
 \begin{document}
@@
 \label{sec:introduction}
+Data visualization can reveal complex relationships in datasets that are not apparent from the descriptive statistical summary\cite{anscombe_graphs_1973}.
+It can be useful to model the variables of a dataset as random variables and explore the probability distributions of those variables, but that is also a more challenging visualization task.
+Many observed relationships are not explicit functional dependencies of variables but instead may be expressed in terms of a probabilistic relationship between random variables.
+For example, a street vendor may want to sell umbrellas.
+They notice that the people who already have umbrellas tend to be wearing coats.
+So should they target people without coats?
+Maybe, but only in that someone who is unprepared for the weather in one way is also likely to be unprepared in other ways.
+So the intervention only make sense on days when the underlying condition for needing the coat and umbrella is present-namely it is raining.
+Being able to visualize that conditional dependence greatly aides in potentially finding it.
 
-\begin{figure}
-  \includegraphics[width=\textwidth]{chart_chooser.png}
-  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram illustrates various ways of representing information about the French workforce in 1954.}
-  \label{fig:chart_chooser}
+\begin{figure}[H]
+	\center
+	\includegraphics[width=1\textwidth]{intro/iris_scatter.png}
+  	\caption{Pairwise relationship between Iris sepal length and Iris petal length (cm) in the Iris scikit-learn dataset\cite{scikit-learn}.}
+  	\label{fig:iris_scatter}
 \end{figure}
 
-In Semiology of Graphics, Bertin \cite{bertin_semiology_2011} presents a dataset of the French work force in 1954.
-This dataset aggregates the workers by region and by sector, and so is structured as follows:
+To understand how to visualize that conditional dependency, we need to talk about visualizations in a formal manner.
+To introduce the visualization terminology used throughout this paper, we use the Iris flower dataset \cite{fisher_use_1936-1, _uci_????} distributed with the scikit-learn Python machine learning library \cite{scikit-learn}.
+The Iris dataset has 150 observations that have a categorical species attribute and quantitative sepal length, sepal width, petal length, and petal width attributes.
+We start probing the Iris dataset for conditional dependencies with the scatter plot in figure~\ref{fig:iris_scatter}.
+The lack of correlation between the petal and length for the Setosa Iris in figure~\ref{fig:iris_scatter}, despite being tightly clustered, indicates that petal and sepal length are dependent on the species of Iris being Setosa.
+While a scatter plot may work for 2 quantitative variables and a relatively small number of data points, even trying to visualize all the variables of the Iris dataset pushes against the limits of the scatter plot.
+This survey will discuss some methods for addressing the challenges of large and complex datasets, but first we must review some visualization terminology.
+
+\subsection{Terminology}
+\begin{figure}[H]
+\input{flowchart.tikz}
+\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant).
+The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
+\label{fig:flowchart}
+\end{figure}
+
+Given a dataset, we need to decide what subset of the data to visualize.
+Bertin describes the set of constraints used to subset the data as the \textit{invariant}.
+Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
+When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertin_semiology_2011}.
+In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
+As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
+components using retinal (visual) variables.
+
+\subsubsection{Retinal (Visual) Variables}
+\begin{figure}[H]
+\includegraphics[width=1\textwidth]{intro/retinal_variables.png}
+\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization.
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} \cite{krygier_making_2005}}
+\label{fig:retinal_variables}
+\end{figure}
+
+Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components}, derived from what Bertin terms a retinal variable and most other visualization theorists call visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014}.
+The columns of figure~\ref{fig:retinal_variables} correspond to the type of observation: discrete points, continuous events (e.g.
+a timeseries), two dimensional continuous events (e.g.
+a vector map).
+The rows of figure~\ref{fig:retinal_variables} describe ways to visualize variations in the \textit{components}; generally, quantitative components are represented by retinal variables that change quantitatively and categorical components are represented by retinal variables that vary qualitatively.
+In figure~\ref{fig:iris_scatter}, the hue of the marker is used to encode differentiation in species and the position of the marker is used to show variation in petal length and sepal length.
+The retinal variables suggest that any single visualization is limited to encoding at most about 8 or 9 components.
+Retinal variables provide guidelines for encoding \textit{components}, but the choice of graph is based on the type and structure of the data.
+
+\subsubsection{Data Type and Structure}
+
+\begin{figure}[H]
+ \includegraphics[width=\textwidth]{intro/munzner_datatypes}
+\caption{Keys are unique lookup values used to find individual observations in the dataset.
+Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
+Image modified from a diagram from Munzner's website \cite{_visualization_????}}
+\label{fig:munzner_datatypes}
+\end{figure}
+
+As shown in figure~\ref{fig:flowchart}, there are multiple ways to translate data into pictures.
+A map is always an option, except when the observations do not have associated coordinates in a physical plane.
+Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munzner_visualization_2014}.
+Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics.
+She defines \textit{keys} as indexes that can be used to look up values, analogous to independent variables in statistics and dimensions in computer science.
+Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset:
 \begin{itemize}
-	\item department (geographic region of France)
-	\item  number of employees per thousands in each sector
-		\begin{itemize}
-			\item primary: agriculture
-			\item secondary: industry 
-			\item tertiary: commerce, transportation, services
-		\end{itemize}
+	\item map: keys are the coordinates of the points
+	\item table: row index, database primary key
+	\item data cube: row, column, etc. (.e.g. $i,j,k$) index
 \end{itemize}
 
-In figure~\ref{fig:chart_chooser}, Bertin explores some potential visualizations of the data, and in doing so also illustrates the trade-offs these visualizations make.
-The first bar graph in the tree shows the number of works per sector, but discards the categorical and encodes the spatial as each bar.
-Categories can be included using a stacked bar chart or three independent bar charts; the former  better preserves the proportionality between sectors within each region, while the latter better displays the regional distribution within each sector.
-Discarding the geographic attributes allows even more interrogation of the relationships within and between categories, via histograms and scatter plots respectively.
+Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field.
+While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsay_functional_2006,muller_functional_2006} of the observational space is also of interest.
+Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.
 
-While some of these approaches include the department as a the y variable, to some extent they discard the geographic information encoded in the department because this data has a physical counterpart that is not referenced in the visualization.
-To explicitly show this geographic information, Bertin proposes a second branch of map based visualizations.
-Taking the same sort of path as the upper branch, Bertin first proposes a map of how the entire workforce is distributed spatially over France.
-He then suggests a cartogram to show relative differences between geographic regions.
-As in the upper branch, he then incorporates the sector information; he proposes bubble charts and bar charts to highlight inter-sector differences and disaggregates the sectors and visualizes them as different maps to show how each sector is distributed over France.
-I
+\subsubsection{Conditional Probability}
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{intro/iris_parallel}
+\caption{The parallel coordinates plot (PCP) of the Iris dataset shows the pairwise relationship between adjacent variables.
+The relative widths of each band of color and the similarities in shape of the orange and green (Iris Versicolour and Iris Virginica) bands and the consistent downward trend amongst the blue (Iris Setosa) bands show how the observations distribute across the species.
+If the visualization only showed one species, this would be the conditional probability of the multivariate observation given that species.}
+\label{fig:iris_parallel}
+\end{figure}
 
-In these visualizations, not only is Bertin showing how many people are in the French workforce in 1954, but he is also showing how that number is conditionally dependent on categorical information such as sector and department.
-With respect to department, he also tries to preserve the geographic distribution of the data.
-Figure~\ref{fig:chart_chooser} is mostly showing the number of employees given department, category, or both.
-While disaggregating employees over sector and geography provides a rich set of potential factors in employee count, it is also somewhat difficult to visualize if the goal is to preserve both regional and categorical differences.
-While 4 maps is a small enough number to not be overwhelming, they forewarn that this technique would likely not work for a larger number of sectors.
+Since the Iris dataset has no geographic or temporal components, we do not need to consider techniques designed to preserve map or time semantics.
+Instead, in figure~\ref{fig:iris_parallel} we plot how the variables in the Iris dataset are related to each other using the the parallel coordinates plot (PCP) \cite{claessen_flexible_2011,_nist/sematech_????,inselberg_plane_1985, wegman_hyperdimensional_1990}.
+In figure~\ref{fig:iris_parallel}, each line is an observation in the dataset and each color maps to a species of Iris.
+The color coding then allows for this visualization to be used to show how the four quantitative variables (sepal length, sepal width, petal length, and petal width) distribute within each variable.
+If we filter the Iris dataset such that we are only looking at one species, then the parallel coordinates plot is visualizing the conditional probability:
+\begin{equation}
+P(A\mid B) = \frac{P(A \cap B}{P(B)} 
+\end{equation}
 
-Despite ostensibly being a categorical variable, department can also be a dimension \cite{munzner_what:_2014} on which either number of employees or sector can depend since it maps to physical space rather than a measurement.
-Number of employees and the sector are traditional quantitative and categorical measurement types.
-This distinction of variable type serves as a basis for understanding what conditional dependencies are explored within each visualization:
+where A is the observation vector $\left< \text{sepal length, sepal width, petal length, petal width} \right>$ and B is the species of petal.
 
-\input{graphtable.tex}
+\subsubsection{Conditional Dependency}
+\label{sec:condep}
+\begin{figure}[H]
+\center
+\includegraphics[width=1\textwidth]{intro/iris_observations.png}
+\caption{The scatter matrix shows the pairwise relationships color coded by Iris species.
+The clustering of the (blue) Setosa Iris measurements even when the measurements are not correlated suggests a conditional dependence on species type.}
+\label{fig:iris_observations}
+\end{figure}
+After observing a relationship between Iris variables in figure~\ref{fig:iris_scatter}, we want to learn if there is an underlying reason for the cooccurence of the the variables.
+The matrix of scatter plots \cite{elmqvist_rolling_2008,l._wilkinson_high-dimensional_2006} in figure~\ref{fig:iris_observations} provides a lot of information about the distribution of variables in the dataset.
+The diagonal of the scatter matrix shows the kernal density estimation (KDE) \cite{scott_multivariate_1992, chambers_graphical_1983, stigler_mathematical_1978} of the distribution of each variable relative to the species type.
+If the KDE plots are filtered by species type, then this visualization illustrates the conditional probability of each variable relative to a type of species.
+The scatter matrix plot also shows every set of pairwise relationships in the dataset, color code by Iris species.
+Because the scatter matrix shows the cooccurence of two variables and is encoded with information about a third variable, it works well as a tool to explore a potential conditional dependency amongst the variables in the dataset \cite{dawid_conditional_1979}.
+As seen in figure~\ref{fig:iris_observations}, the  measurements of sepal width, sepal length, petal width, and petal length are correlated for Versicolour and Virginica Irises but not for Setosa Irises.
+Despite the lack of correlation, sepal width, sepal length, petal width, and petal length are tightly clustered for Setosa Irises; this clustering indicates that these quantitative measurements are conditionally dependent on species.
+The scatter matrix is limited to a relatively small number of variables, can only show pairwise interactions, and does not retain the semantic structure of the data.
 
-This survey presents a sampling of visualization techniques that attempt to visually represent conditional dependencies between a variety of variables, including:
-\begin{enumerate}
-	\item quantitative and categorical variables
-	\item distributions and densities
-	\item variables with dimensional dependencies
-\end{enumerate}
+\begin{figure}[H]
+\center
+\input{papermap.tikz}
+\caption{The papers in this survey are organized by the structure of observations and types of measurements in a dataset.}
+\label{fig:papermap}
+\end{figure}
 
-\subfile{sections/distdist.tex}
-\subfile{sections/distdens.tex}
-\subfile{sections/densdist.tex}
-\subfile{sections/densdens.tex}
+Although the scatter matrix is limited in the datasets it will work for, there are many other visualization techniques to choose from.
+As illustrated in figure~\ref{fig:papermap}, this survey provides recommendations for visualization methods based on the structure of the observations and types of measurements in the dataset.
+This survey presents methods for visualizing conditional dependencies for multivariate datasets with discrete observations and understanding probability distributions for datasets where observations are drawn from a continuous (functional) space.
+Since there are very few visualizations techniques explicitly designed to show conditional dependency, we include techniques that either suggest a conditional dependency or can be adapted to do so.
+
+\section{Discrete Observations}
+\subfile{sections/section1}
+\subfile{sections/section2}
+\subfile{sections/section3}
+\section{Continuous Observations}
+\subfile{sections/section4}
+\subfile{sections/section5}
 
@@
 \label{sec:conclusion}
-When trying to choose a visualization technique for a problem, it is often
-crucial to interrogate the questions being asked of the data. What
-relationships are important to preserve and which can be discarded? While there
-are a what feels like infinite number of visualization techniques, this
-visualization task can be rather complicated, especially as the number of
-variables or dependencies increases. Interactive tools can help, but most are limited to showing pairwise or a small number of connected interactions. Machine learning can somewhat be used
-to explore which variables and dimensions, and therefore which relationships, are worth
-exploring further, but often the results of those algorithms need to
-be translated to the visual space and those visualizations need to be
-understandable.
+While this paper described some methods for visualizing conditional dependencies, they are limited.
+Many visualization techniques only work for a relatively small number of variables, or only when the condition is categorical or temporal.
+When data has multiple keys, most methods at best can only preserve time or one other one dimensional key.
+The most promising techniques in the functional space are methods for visualizing distributions and the uncertainty in those visualizations.
+These techniques could potentially be combined with methods for exploring conditional dependency in discrete and less functional datasets to build new tools for exploring complex data spaces.
 
